---
title: "Registration Benchmark v2 — Study A"
subtitle: "Internal Report — tf Package"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  cache: true
  warning: false
  message: false
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| cache: false
library(ggplot2)
library(data.table)
library(patchwork)

base_dir <- here::here()
source(file.path(base_dir, "sim-analyze.R"))

mcols <- method_colors()
mlabs <- method_labels()
dlabs <- dgp_short_labels()
ddesc <- dgp_desc_labels()
```

```{r}
#| label: load-data
dt <- load_study_a()
```

Study A compares 5 registration methods across 15 DGPs in a crossed design:
15 DGPs $\times$ 2 severity (0.5, 1.0) $\times$ 3 noise (0, 0.1, 0.3) $\times$
5 methods = 450 cells $\times$ 100 reps = 45,000 runs.

The DGP design is **not** full factorial (15 of 24 possible template $\times$ warp
$\times$ amplitude combinations), but enables **paired comparisons** that vary
exactly one factor while holding the other two constant.

```{r}
#| label: dgp-table
knitr::kable(
  dgp_factors(),
  caption = "DGP factor structure. 15 DGPs spanning 2 templates, 3 warp types, 4 amplitude types."
)
```

# DGP Gallery {#sec-gallery}

Example realizations for each DGP (n=20, severity=0.5, no noise). Each row shows
the template, the warping functions $h_i(t)$ (dashed = identity), and the resulting
observed curves $x_i(t)$.

::: {.panel-tabset}

## Baseline (D01--D04)

```{r}
#| label: gallery-baseline
#| fig-height: 12
#| fig-cap: "Baseline DGPs: phase variation only (no amplitude variation)."
make_dgp_example("D01") / make_dgp_example("D02") /
  make_dgp_example("D03") / make_dgp_example("D04")
```

## Low-rank amplitude (D05--D08)

```{r}
#| label: gallery-lowrank
#| fig-height: 12
#| fig-cap: "Low-rank amplitude DGPs: rank-1 multiplicative or rank-2 (scale + shift)."
make_dgp_example("D05") / make_dgp_example("D06") /
  make_dgp_example("D07") / make_dgp_example("D08")
```

## High-rank amplitude (D09--D13, D15)

```{r}
#| label: gallery-highrank
#| fig-height: 18
#| fig-cap: "High-rank amplitude DGPs: 3 FPCs from gait knee data. D13/D15 use non-domain-preserving affine warps."
make_dgp_example("D09") / make_dgp_example("D10") /
  make_dgp_example("D11") / make_dgp_example("D12") /
  make_dgp_example("D13") / make_dgp_example("D15")
```

## New baseline (D14)

```{r}
#| label: gallery-d14
#| fig-height: 3
#| fig-cap: "D14: wiggly template with simple warps and no amplitude variation (completes the wiggly-simple amplitude ladder)."
make_dgp_example("D14")
```

:::

# Overview: DGP $\times$ Method Heatmaps {#sec-overview}

The heatmaps below show the **median** of each metric across all 400 runs per cell
(100 reps $\times$ 2 severity $\times$ 2 noise), giving a quick orientation.
Color encodes the **rank** within each DGP (row): yellow = best (rank 1),
purple = worst (rank 5). Numeric labels show the actual median values.

```{r}
#| label: heatmap-warp-mise
#| fig-height: 7
#| fig-cap: "Median warp MISE across all severity/noise conditions. Color = rank within each DGP; yellow = best, purple = worst. Labels show actual median MISE."
make_heatmap(dt, "warp_mise", "Warp MISE (median, all conditions)")
```

```{r}
#| label: heatmap-alignment-cc
#| fig-height: 7
#| fig-cap: "Median alignment cross-correlation. Color = rank within each DGP; yellow = best, purple = worst."
make_heatmap(dt, "alignment_cc", "Alignment CC (median, all conditions)",
  lower_is_better = FALSE)
```

The win-rate heatmaps show the percentage of individual runs (across all conditions)
where each method achieves the best metric value among all non-failed methods for
that specific run. This complements the median-based heatmaps by revealing how
*consistently* a method dominates, not just on average.

```{r}
#| label: heatmap-winrate-warp-mise
#| fig-height: 7
#| fig-cap: "Win rate for warp MISE: percentage of runs where each method achieves the lowest warp MISE among all methods, per DGP."
make_winrate_heatmap(dt, "warp_mise", "Warp MISE win rate (all conditions)")
```

```{r}
#| label: heatmap-winrate-alignment-cc
#| fig-height: 7
#| fig-cap: "Win rate for alignment CC: percentage of runs where each method achieves the highest cross-correlation."
make_winrate_heatmap(dt, "alignment_cc", "Alignment CC win rate (all conditions)",
  lower_is_better = FALSE)
```

```{r}
#| label: overview-findings
#| results: asis
# Rank distribution per method across DGPs
ranks_all <- dt[failure == FALSE,
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method)
][, rank := rank(mise), by = dgp]

rank_summary <- ranks_all[,
  .(mean_rank = round(mean(rank), 1),
    n_rank1 = sum(rank == 1),
    n_top2 = sum(rank <= 2)),
  by = method
][order(mean_rank)]

cat("**Key observations** (pooling over severity and noise):\n\n")
for (i in seq_len(nrow(rank_summary))) {
  m <- rank_summary$method[i]
  cat(sprintf(
    "- **%s**: mean rank %.1f, rank 1 on %d/%d DGPs, top 2 on %d/%d DGPs.\n",
    mlabs[m], rank_summary$mean_rank[i],
    rank_summary$n_rank1[i], uniqueN(dt$dgp),
    rank_summary$n_top2[i], uniqueN(dt$dgp)
  ))
}
cat("\n- The heatmaps pool over severity and noise; the conditional analyses below provide sharper comparisons.\n")
cat("\n")
```

# Warp Recovery {#sec-warp}

## By warp type (Q1) {#sec-q1}

*Question: Which method best recovers the true warps, conditional on warp type?*

**Warp MISE** measures the integrated squared error between estimated and true warping
functions: $\text{MISE}_h = \int_0^1 (\hat{h}_i(t) - h_i(t))^2\, dt$, averaged over
curves $i$. Lower is better.

**Implementation note:** Warp MISE compares forward warps $\hat{h}_i$ vs $h_i$
for all warp types. Estimated forward warps are recovered by inverting the stored
inverse warps from `tf_registration`.

Warp types represent fundamentally different problems — smooth elastic warps (simple/complex)
vs affine warps — so we stratify all comparisons by warp type. Within each stratum,
we show the full distribution of warp MISE across all DGPs with that warp type.

```{r}
#| label: q1-boxplot
#| fig-height: 9
#| fig-cap: "Distribution of warp MISE by method, stratified by warp type (rows). Columns show the 4 severity × noise conditions. Each boxplot summarizes all reps across all DGPs with that warp type."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = warp_mise, fill = method)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_grid(
    warp_type ~ severity + noise_sd,
    scales = "free_x",
    labeller = cond_labeller
  ) +
  scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  labs(y = NULL, x = "Warp MISE (log scale)") +
  theme_benchmark()
```

```{r}
#| label: q1-findings
#| results: asis
# Compute best method per warp type x severity x noise
q1_ranks <- dt[failure == FALSE,
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(method, warp_type, severity, noise_sd)
][, rank := rank(mise), by = .(warp_type, severity, noise_sd)]

# Best method per warp type (pooled)
q1_pooled <- dt[failure == FALSE,
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(method, warp_type)
][, rank := rank(mise), by = warp_type]

best_simple <- mlabs[q1_pooled[warp_type == "simple"][which.min(mise), method]]
best_complex <- mlabs[q1_pooled[warp_type == "complex"][which.min(mise), method]]
best_affine <- mlabs[q1_pooled[warp_type == "affine"][which.min(mise), method]]

# Check ranking stability: does best method stay rank 1 across conditions?
rank1_simple <- q1_ranks[warp_type == "simple" & rank == 1, unique(method)]
rank1_complex <- q1_ranks[warp_type == "complex" & rank == 1, unique(method)]

cat("**Findings:**\n\n")
cat(sprintf(
  "- On **simple** warps, %s achieves the lowest median warp MISE (pooled). %s\n",
  best_simple,
  if (length(rank1_simple) == 1)
    sprintf("This holds across all severity/noise conditions.")
  else
    sprintf("The best method varies by condition (%s).",
      paste(mlabs[rank1_simple], collapse = ", "))
))
cat(sprintf(
  "- On **complex** warps, %s achieves the lowest median warp MISE (pooled). %s\n",
  best_complex,
  if (length(rank1_complex) == 1)
    sprintf("This holds across all severity/noise conditions.")
  else
    sprintf("The best method varies by condition (%s).",
      paste(mlabs[rank1_complex], collapse = ", "))
))
cat(sprintf(
  "- On **affine** warps, %s achieves the lowest MISE — paradoxical if Affine (S+S) is well-specified (discussed in @sec-affine-puzzle).\n",
  best_affine
))
cat("- Higher severity and noise both increase MISE, but method rankings are largely stable within warp types.\n")
# Check FDA tail behavior
fda_iqr <- dt[failure == FALSE & method == "fda_default",
  .(iqr = IQR(warp_mise, na.rm = TRUE)), by = warp_type]
srvf_iqr <- dt[failure == FALSE & method == "srvf",
  .(iqr = IQR(warp_mise, na.rm = TRUE)), by = warp_type]
if (all(fda_iqr$iqr > srvf_iqr$iqr)) {
  cat("- FDA methods show competitive medians but heavier tails (more variable performance).\n")
}
cat("\n")
```

```{r}
#| label: q1-summary
q1_summary <- mc_summary(dt, warp_mise,
  by = c("method", "warp_type", "severity"))
q1_summary[, median_fmt := fmt(median)]
q1_summary[, se_fmt := fmt(mc_se)]
q1_wide <- dcast(
  q1_summary[, .(method, warp_type, severity, median_fmt, se_fmt)],
  method + warp_type ~ severity,
  value.var = c("median_fmt", "se_fmt")
)
knitr::kable(
  q1_wide[order(warp_type, method)],
  caption = "Median warp MISE (± MC SE) by method, warp type, and severity, pooling over noise and DGPs within each warp type."
)
```

## Phase-amplitude interaction (Q2) {#sec-q2}

*Question: Does amplitude variation degrade warp recovery?*

We use **amplitude ladders** — sequences of DGPs that hold template and warp type constant
while increasing amplitude complexity (none $\to$ rank-1 $\to$ rank-2 $\to$ high-rank).
An upward slope means that method's warp recovery degrades as amplitude complexity increases.

```{r}
#| label: q2-ladders
#| fig-height: 10
#| fig-cap: "Amplitude ladders: median warp MISE (IQR error bars) vs amplitude complexity. Each panel row holds template and warp type fixed; columns show the 4 severity × noise conditions. Upward slopes indicate degraded warp recovery under amplitude variation."
ladders <- amplitude_ladders()
plots <- lapply(names(ladders), function(nm) {
  make_ladder_plot(dt, ladders[[nm]], nm)
})
wrap_plots(plots, ncol = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: q2-findings
#| results: asis
# Check degradation on each ladder: does MISE go up with amplitude complexity?
ladders_q2 <- amplitude_ladders()

cat("**Findings:**\n\n")

# Count how often each method degrades vs improves across ladders and conditions
n_degrade <- 0L
n_improve <- 0L
n_total <- 0L
for (lnm in names(ladders_q2)) {
  ld <- ladders_q2[[lnm]]
  dgps <- names(ld)
  first_dgp <- dgps[1]
  last_dgp <- dgps[length(dgps)]
  for (m in levels(dt$method)) {
    for (s in levels(dt$severity)) {
      for (ns in levels(dt$noise_sd)) {
        v_first <- dt[dgp == first_dgp & method == m & severity == s &
          noise_sd == ns & failure == FALSE, median(warp_mise, na.rm = TRUE)]
        v_last <- dt[dgp == last_dgp & method == m & severity == s &
          noise_sd == ns & failure == FALSE, median(warp_mise, na.rm = TRUE)]
        if (length(v_first) && length(v_last) && !is.na(v_first) && !is.na(v_last)) {
          n_total <- n_total + 1L
          if (v_last > v_first) n_degrade <- n_degrade + 1L
          else n_improve <- n_improve + 1L
        }
      }
    }
  }
}

pct_degrade <- round(100 * n_degrade / n_total)
cat(sprintf(
  "- In %d%% of method-ladder-condition combinations (%d/%d), warp MISE increases with amplitude complexity. The pattern is predominant but not universal — some methods on some ladders show stable or slightly improved MISE under higher-rank amplitude.\n",
  pct_degrade, n_degrade, n_total
))

# Check which method has best absolute MISE at highrank
best_highrank <- dt[failure == FALSE & amplitude == "highrank",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = method][which.min(mise), method]
cat(sprintf(
  "- **%s** generally retains the lowest absolute MISE as amplitude complexity grows, though its *relative* degradation can be comparable to other methods.\n",
  mlabs[best_highrank]
))
cat("- The magnitude and direction of amplitude effects vary across ladders and conditions — see the plots above for per-ladder details.\n")
cat("\n")
```

## Warp complexity effect (Q3) {#sec-q3}

*Question: How much does warp complexity cost each method?*

**Warp ladders** hold template and amplitude constant while increasing warp complexity
(affine $\to$ simple $\to$ complex).

```{r}
#| label: q3-ladders
#| fig-height: 9
#| fig-cap: "Warp ladders: median warp MISE (IQR error bars) vs warp complexity. Each panel row holds template and amplitude fixed; columns show the 4 severity × noise conditions."
ladders_w <- warp_ladders()
plots_w <- lapply(names(ladders_w), function(nm) {
  make_ladder_plot(dt, ladders_w[[nm]], nm)
})
wrap_plots(plots_w, ncol = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: q3-findings
#| results: asis
# Check how often complex > simple MISE across methods and conditions
ladders_w <- warp_ladders()
n_worse <- 0L
n_better <- 0L
n_tot_q3 <- 0L
for (lnm in names(ladders_w)) {
  ld <- ladders_w[[lnm]]
  dgps <- names(ld)
  if (!("simple" %in% ld && "complex" %in% ld)) next
  d_simple <- names(ld)[ld == "simple"]
  d_complex <- names(ld)[ld == "complex"]
  for (m in levels(dt$method)) {
    for (s in levels(dt$severity)) {
      for (ns in levels(dt$noise_sd)) {
        v_s <- dt[dgp == d_simple & method == m & severity == s &
          noise_sd == ns & failure == FALSE, median(warp_mise, na.rm = TRUE)]
        v_c <- dt[dgp == d_complex & method == m & severity == s &
          noise_sd == ns & failure == FALSE, median(warp_mise, na.rm = TRUE)]
        if (length(v_s) && length(v_c) && !is.na(v_s) && !is.na(v_c)) {
          n_tot_q3 <- n_tot_q3 + 1L
          if (v_c > v_s) n_worse <- n_worse + 1L
          else n_better <- n_better + 1L
        }
      }
    }
  }
}
pct_worse <- round(100 * n_worse / n_tot_q3)

cat("**Findings:**\n\n")
cat(sprintf(
  "- In %d%% of comparisons (%d/%d), complex warps yield higher MISE than simple warps. The pattern holds for most methods and conditions but has exceptions — some methods show comparable or slightly lower MISE on complex vs simple warps in specific settings.\n",
  pct_worse, n_worse, n_tot_q3
))
cat("- On **affine** DGPs: SRVF and Landmark maintain low MISE despite the non-domain-preserving warps, while Affine (S+S) has higher MISE despite being well-specified.\n")
cat("- The combination of non-domain-preserving warps and high-rank amplitude (D13) appears especially challenging for affine regression, though the mechanism is not experimentally isolated (see @sec-affine-puzzle).\n")
cat("\n")
```

### The affine puzzle {#sec-affine-puzzle}

The Affine (S+S) method is **well-specified** for D04 and D13 (the true warps are
affine), yet it consistently achieves 3--4$\times$ higher warp MISE than SRVF or
Landmark (auto). Several factors likely contribute:

1. **Template estimation bias** (conjecture): Affine registration uses the arithmetic
   $L^2$ mean as template. Under non-domain-preserving warps, this is not the Fréchet
   mean in shape space. SRVF uses the Karcher mean, which may reduce template bias.
   This hypothesis is plausible but not experimentally isolated in this study.
2. **Weak signal**: At severity=0.5, the affine warps are very close to the identity
   ($\text{sd}_a = 0.025$, $\text{sd}_b = 0.015$), giving the affine regression
   little signal to estimate (a, b) from.
3. **Iteration count**: The affine method uses 10 Procrustes iterations, which may
   not suffice for joint template-warp estimation in this setting.

This is a notable observation: well-specified parametric models *can* underperform
flexible nonparametric methods. The mechanisms above are plausible explanations
but remain conjectures — isolating them would require targeted follow-up experiments
(e.g., oracle template, higher severity).


# Registration Diagnostics {#sec-diag}

## Over/under-registration (Q4) {#sec-q4}

*Question: Do methods systematically over-register or under-register?*

The **registration ratio** measures how much the estimated warp deviates from
the identity relative to the true warp:
$$R_i = \frac{\|\hat{h}_i - \mathrm{id}\|^2}{\|h_i - \mathrm{id}\|^2}$$
where $\|\cdot\|^2 = \int_0^1 (\cdot)^2\,dt$.
Values $R < 1$ indicate under-registration (estimated warp closer to identity
than truth); $R > 1$ indicates over-registration (warped too much).
We report the median $R$ across curves within each run.

```{r}
#| label: q4-reg-ratio
#| fig-height: 8
#| fig-cap: "Registration ratio per DGP and method (median across 100 reps, severity=0.5, noise=0). Error bars: IQR. Dashed line at 1 = perfectly calibrated."
make_reg_ratio_plot(dt)
```

```{r}
#| label: q4-reg-ratio-aggregated
#| fig-height: 5
#| fig-cap: "Registration ratio aggregated by method and warp type (median and IQR across all DGPs within each warp type, severity=0.5, noise=0)."
rr_agg <- dt[
  failure == FALSE & !is.na(registration_ratio_median) &
    severity == "0.5" & noise_sd == "0",
  .(
    median_rr = median(registration_ratio_median, na.rm = TRUE),
    q25 = quantile(registration_ratio_median, 0.25, na.rm = TRUE),
    q75 = quantile(registration_ratio_median, 0.75, na.rm = TRUE)
  ),
  by = .(method, warp_type)
]

ggplot(rr_agg, aes(y = method, x = median_rr, color = method)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = q25, xmax = q75), height = 0.2) +
  facet_wrap(~warp_type) +
  scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_continuous(trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("1/4", "1/2", "1", "2", "4")) +
  labs(y = NULL, x = expression("Registration ratio (" * log[2] * " scale)")) +
  theme_benchmark()
```

```{r}
#| label: q4-findings
#| results: asis
# Compute registration ratio by method x warp_type
rr_by_mw <- dt[
  failure == FALSE & !is.na(registration_ratio_median),
  .(rr = median(registration_ratio_median, na.rm = TRUE)),
  by = .(method, warp_type)
]

rr_fmt <- function(m, w) {
  v <- rr_by_mw[method == m & warp_type == w, rr]
  if (length(v) == 0) return("NA")
  sprintf("%.2f", v)
}

cat("**Findings:**\n\n")
cat(sprintf(
  "- **SRVF** is well-calibrated on simple (median ratio %s) and complex (%s) warps, with ratios close to 1.\n",
  rr_fmt("srvf", "simple"), rr_fmt("srvf", "complex")
))
cat(sprintf(
  "- **Landmark (auto)** systematically under-registers: simple %s, complex %s (ratios well below 1, meaning estimated warps are closer to the identity than the truth).\n",
  rr_fmt("landmark_auto", "simple"), rr_fmt("landmark_auto", "complex")
))
cat(sprintf(
  "- **FDA (default)** and **FDA (crit 1)** tend to over-register, especially on complex warps: FDA default simple/complex = %s/%s, FDA crit 1 = %s/%s.\n",
  rr_fmt("fda_default", "simple"), rr_fmt("fda_default", "complex"),
  rr_fmt("fda_crit1", "simple"), rr_fmt("fda_crit1", "complex")
))
cat(sprintf(
  "- **Affine (S+S)** is well-calibrated on affine DGPs (ratio %s) but shows variable behavior on other warp types.\n",
  rr_fmt("affine_ss", "affine")
))
cat("\n")
```

The **warp slope ratio** measures how much "wiggliness" the estimated warp has
relative to the truth:
$$S_i = \frac{\max \hat{h}_i'(t) - \min \hat{h}_i'(t)}{\max h_i'(t) - \min h_i'(t)}$$
Values $S > 1$ mean the estimated warp is too wiggly; $S < 1$ means too smooth.
This diagnostic is only meaningful for non-affine warps (affine warps have
constant derivative, making the denominator uninformative).

```{r}
#| label: q4-warp-slope
#| fig-height: 5
#| fig-cap: "Warp slope ratio by method and warp type (non-affine DGPs only). Median and IQR across all conditions. Values > 1 = estimated warps too wiggly; < 1 = too smooth."
dt_slope <- dt[
  warp_type != "affine" & failure == FALSE & !is.na(warp_slope_ratio_median)
]
if (nrow(dt_slope) > 0) {
  slope_agg <- dt_slope[, .(
    median_sr = median(warp_slope_ratio_median, na.rm = TRUE),
    q25 = quantile(warp_slope_ratio_median, 0.25, na.rm = TRUE),
    q75 = quantile(warp_slope_ratio_median, 0.75, na.rm = TRUE)
  ), by = .(method, warp_type)]

  ggplot(slope_agg, aes(y = method, x = median_sr, color = method)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
    geom_point(size = 3) +
    geom_errorbarh(aes(xmin = q25, xmax = q75), height = 0.2) +
    facet_wrap(~warp_type) +
    scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_y_discrete(labels = mlabs) +
    scale_x_continuous(trans = "log2") +
    labs(y = NULL, x = expression("Warp slope ratio (" * log[2] * " scale)")) +
    theme_benchmark()
}
```

```{r}
#| label: q4-slope-findings
#| results: asis
# Compute warp slope ratio by method x warp_type
sr_by_mw <- dt[
  warp_type != "affine" & failure == FALSE & !is.na(warp_slope_ratio_median),
  .(sr = median(warp_slope_ratio_median, na.rm = TRUE)),
  by = .(method, warp_type)
]

sr_fmt <- function(m, w) {
  v <- sr_by_mw[method == m & warp_type == w, sr]
  if (length(v) == 0) return("NA")
  sprintf("%.2f", v)
}

cat("**Warp slope findings:**\n\n")
for (m in levels(dt$method)) {
  vals <- sr_by_mw[method == m]
  if (nrow(vals) == 0) next
  parts <- paste(
    sprintf("%s: %s", vals$warp_type, sprintf("%.2f", vals$sr)),
    collapse = ", "
  )
  cat(sprintf("- **%s**: %s\n", mlabs[m], parts))
}
cat("\nValues > 1 indicate estimated warps are wigglier than truth; < 1 indicates smoother.\n\n")
```


## Phase-amplitude separation (Q2b) {#sec-amp-ratio}

*Question: Do methods confuse phase and amplitude variation?*

The **amplitude variance ratio** compares the residual amplitude variance after
registration to the true amplitude variance:
$$\text{AVR} = \frac{\overline{\text{Var}}_t\bigl(\hat{x}_i^{\text{aligned}}(t) - \hat{\mu}(t)\bigr)}{\overline{\text{Var}}_t\bigl(\tilde{x}_i(t) - \mu(t)\bigr)}$$
where $\hat{\mu}$ is the estimated template, $\tilde{x}_i = \mu \circ h_i + a_i$ is
the true pre-warp curve (template composed with true warp, plus true amplitude
component $a_i$), and $\overline{\text{Var}}_t$ denotes the pointwise variance
across curves $i$, averaged over $t$. Thus the numerator measures the residual
curve-to-curve variation after registration, and the denominator measures the
true amplitude variation in the DGP.

- $\text{AVR} < 1$ suggests **over-registration**: the method absorbed some
  amplitude variation into the phase component.
- $\text{AVR} > 1$ suggests **under-registration**: residual phase variation
  remains in the amplitude component, or the estimated template differs from
  the truth, inflating residuals.

This diagnostic is only shown for DGPs with amplitude variation (D05--D13).

```{r}
#| label: q4-amp-variance-ratio
#| fig-height: 9
#| fig-cap: "Amplitude variance ratio by method, stratified by warp type (rows) and severity × noise conditions (columns). DGPs with amplitude variation only (D05--D13). Dashed line at 1. Boxplots pool across DGPs within each warp type."
dt_amp <- dt[amplitude != "none" & failure == FALSE & !is.na(amp_variance_ratio)]

if (nrow(dt_amp) > 0) {
  ggplot(dt_amp, aes(y = method, x = amp_variance_ratio, fill = method)) +
    geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
    geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
    facet_grid(
      warp_type ~ severity + noise_sd,
      scales = "free_x",
      labeller = cond_labeller
    ) +
    scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_y_discrete(labels = mlabs) +
    scale_x_continuous(trans = "log2",
      breaks = c(0.125, 0.25, 0.5, 1, 2, 4, 8),
      labels = c("1/8", "1/4", "1/2", "1", "2", "4", "8")) +
    labs(y = NULL, x = expression("Amplitude variance ratio (" * log[2] * " scale)")) +
    theme_benchmark()
}
```

```{r}
#| label: q2b-findings
#| results: asis
# Compute AVR by method (pooled across amplitude DGPs)
avr_by_m <- dt[
  amplitude != "none" & failure == FALSE & !is.na(amp_variance_ratio),
  .(avr_med = median(amp_variance_ratio, na.rm = TRUE),
    avr_q25 = quantile(amp_variance_ratio, 0.25, na.rm = TRUE),
    avr_q75 = quantile(amp_variance_ratio, 0.75, na.rm = TRUE)),
  by = method
]

cat("**Findings:**\n\n")
for (i in seq_len(nrow(avr_by_m))) {
  m <- avr_by_m$method[i]
  med <- avr_by_m$avr_med[i]
  q25 <- avr_by_m$avr_q25[i]
  q75 <- avr_by_m$avr_q75[i]
  direction <- if (med < 0.9) {
    "over-registration (absorbing amplitude into phase)"
  } else if (med > 1.1) {
    "under-registration or template bias inflating residuals"
  } else {
    "near-calibrated residual amplitude variance"
  }
  cat(sprintf(
    "- **%s**: median AVR = %.2f [IQR: %.2f--%.2f], suggesting %s.\n",
    mlabs[m], med, q25, q75, direction
  ))
}
cat("\nNote: AVR > 1 can reflect either residual phase variation (under-registration) *or* template estimation error inflating residuals — these effects are confounded in this metric.\n")
cat("\n")
```


## Template estimation (Q5) {#sec-q5}

*Question: How well does each method estimate the template?*

**Template MISE** measures $\int_0^1 (\hat{\mu}(t) - \mu(t))^2\, dt$ where
$\hat{\mu}$ is the method's estimated template and $\mu$ the true template.
This conflates shape error with phase error: two templates differing only by
a warp are "the same shape" but have nonzero $L^2$ distance.

```{r}
#| label: q5-template
#| fig-height: 7
#| fig-cap: "Distribution of template MISE by method, conditioned on severity and noise (columns). Boxplots pool across all 15 DGPs (100 reps each)."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = template_mise, fill = method)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_grid(
    . ~ severity + noise_sd,
    labeller = cond_labeller
  ) +
  scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  labs(y = NULL, x = "Template MISE (log scale)") +
  theme_benchmark()
```

To address the phase-shape confound in $L^2$ template MISE, we also compute the
**elastic (SRSF) distance** between estimated and true templates:
$$d_{\text{FR}}(\hat\mu, \mu) = \| q_{\hat\mu} - q_\mu \|_{L^2}, \quad
q_f(t) = \text{sign}(f'(t))\sqrt{|f'(t)|}$$
This measures template *shape* distance invariant to reparameterization.

```{r}
#| label: q5-elastic
#| fig-height: 7
#| fig-cap: "Distribution of elastic (SRSF) template distance by method, conditioned on severity and noise. Lower is better. Unlike L2 MISE, this metric is invariant to phase shifts in the template."
if ("template_elastic_dist" %in% names(dt)) {
  ggplot(
    dt[failure == FALSE & !is.na(template_elastic_dist)],
    aes(y = method, x = template_elastic_dist, fill = method)
  ) +
    geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
    facet_grid(
      . ~ severity + noise_sd,
      labeller = cond_labeller
    ) +
    scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_y_discrete(labels = mlabs) +
    scale_x_log10() +
    labs(y = NULL, x = "Elastic template distance (log scale)") +
    theme_benchmark()
} else {
  cat("*template_elastic_dist not available in current results — will appear after re-run.*")
}
```

```{r}
#| label: q5-findings
#| results: asis
# Best method per severity x noise panel (pooled across DGPs)
q5_best <- dt[failure == FALSE,
  .(tmise = median(template_mise, na.rm = TRUE)),
  by = .(method, severity, noise_sd)
]

has_elastic <- "template_elastic_dist" %in% names(dt)
if (has_elastic) {
  q5_elastic <- dt[failure == FALSE & !is.na(template_elastic_dist),
    .(edist = median(template_elastic_dist, na.rm = TRUE)),
    by = .(method, severity, noise_sd)
  ]
}

cat("**Findings (L2 MISE):**\n\n")
cat("Best method (lowest median template MISE) per condition:\n\n")
for (s in levels(dt$severity)) {
  for (ns in levels(dt$noise_sd)) {
    sub <- q5_best[severity == s & noise_sd == ns]
    best_m <- sub[which.min(tmise), method]
    best_v <- sub[which.min(tmise), tmise]
    cat(sprintf(
      "- sev=%s, noise=%s: **%s** (%.4f)\n",
      s, ns, mlabs[best_m], best_v
    ))
  }
}
cat("\n")

if (has_elastic) {
  cat("**Findings (elastic distance):**\n\n")
  cat("Best method (lowest median elastic distance) per condition:\n\n")
  for (s in levels(dt$severity)) {
    for (ns in levels(dt$noise_sd)) {
      sub <- q5_elastic[severity == s & noise_sd == ns]
      best_m <- sub[which.min(edist), method]
      best_v <- sub[which.min(edist), edist]
      cat(sprintf(
        "- sev=%s, noise=%s: **%s** (%.4f)\n",
        s, ns, mlabs[best_m], best_v
      ))
    }
  }
  cat("\n")
}

# Worst method overall
worst_tpl <- q5_best[, .(tmise = median(tmise)), by = method][which.max(tmise), method]
cat(sprintf(
  "**%s** consistently shows the highest template MISE.\n\n",
  mlabs[worst_tpl]
))
cat("**Caveat:** Template MISE uses $L^2$ distance, which conflates shape error with phase error. The elastic (SRSF) distance above addresses this by measuring shape distance invariant to reparameterization.\n\n")
```

# Sensitivity (Q6) {#sec-sensitivity}

*Question: Which methods are most sensitive to noise and severity?*

We compute **degradation ratios**: the ratio of median warp MISE under the harder condition
to the easier condition. A ratio of 2 means the metric doubled. Each dot below represents
one DGP; the crossbar shows the median ratio across DGPs.

## Noise degradation

```{r}
#| label: q6-noise
#| fig-height: 5
#| fig-cap: "Noise degradation: ratio of median warp MISE at each noise level vs noise=0 (severity=0.5 fixed). Panels show noise=0.1 and noise=0.3. Each dot = one DGP; crossbar = median across DGPs."
ratios <- degradation_ratios(dt)
noise_dt <- ratios$noise[severity == "0.5" & is.finite(noise_ratio)]

if (nrow(noise_dt) > 0) {
  ggplot(noise_dt, aes(x = method, y = noise_ratio, color = method)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.4,
      color = "black", linewidth = 0.4) +
    facet_wrap(~noise_sd, labeller = labeller(
      noise_sd = function(x) paste0("noise=", x, " vs 0"))) +
    scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_x_discrete(labels = mlabs) +
    scale_y_continuous(trans = "log2",
      breaks = c(0.5, 1, 2, 4, 8, 16, 32, 64, 128),
      labels = c("0.5", "1", "2", "4", "8", "16", "32", "64", "128")) +
    labs(
      x = NULL,
      y = expression("Noise degradation ratio (" * log[2] * " scale)")
    ) +
    theme_benchmark() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Severity degradation

```{r}
#| label: q6-severity
#| fig-height: 5
#| fig-cap: "Severity degradation: ratio of median warp MISE at severity=1.0 vs severity=0.5 (noise=0 fixed). Each dot = one DGP; crossbar = median across DGPs."
sev_dt <- ratios$severity[noise_sd == "0" & is.finite(severity_ratio)]

if (nrow(sev_dt) > 0) {
  ggplot(sev_dt, aes(x = method, y = severity_ratio, color = method)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.4,
      color = "black", linewidth = 0.4) +
    scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_x_discrete(labels = mlabs) +
    scale_y_continuous(trans = "log2",
      breaks = c(0.5, 1, 2, 4, 8, 16, 32),
      labels = c("0.5", "1", "2", "4", "8", "16", "32")) +
    labs(
      x = NULL,
      y = expression("Severity degradation ratio (" * log[2] * " scale)")
    ) +
    theme_benchmark() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

```{r}
#| label: q6-findings
#| results: asis
# Compute actual median degradation ratios per noise level
noise_med <- ratios$noise[
  severity == "0.5" & is.finite(noise_ratio),
  .(ratio = median(noise_ratio, na.rm = TRUE)),
  by = .(method, noise_sd)
][order(noise_sd, -ratio)]

sev_med <- ratios$severity[
  noise_sd == "0" & is.finite(severity_ratio),
  .(ratio = median(severity_ratio, na.rm = TRUE)),
  by = method
][order(-ratio)]

cat("**Findings:**\n\n")

# Noise — per level
noise_levels_present <- sort(unique(noise_med$noise_sd))
for (nl in noise_levels_present) {
  nl_sub <- noise_med[noise_sd == nl]
  most <- nl_sub[which.max(ratio)]
  least <- nl_sub[which.min(ratio)]
  cat(sprintf(
    "- **Noise=%s vs 0**: strongest effect on %s (%.1f$\\times$), weakest on %s (%.2f$\\times$).\n",
    nl, mlabs[most$method], most$ratio,
    mlabs[least$method], least$ratio
  ))
}
# SRVF noise note
if (noise_med[noise_sd == noise_levels_present[1] & method == "srvf", .N] > 0) {
  cat("  Unregularized SRVF operates on derivatives (SRSFs), which amplify high-frequency noise. Study B (penalization) will test whether regularization mitigates this.\n")
}

# Severity
most_sev <- sev_med[1]
least_sev <- sev_med[.N]
cat(sprintf(
  "- **Severity** increases MISE for all methods, with the strongest effect on %s (%.1f$\\times$) and the weakest on %s (%.1f$\\times$).\n",
  mlabs[most_sev$method], most_sev$ratio,
  mlabs[least_sev$method], least_sev$ratio
))
cat("\n")

# Absolute degradation context
abs_easy <- dt[failure == FALSE & severity == "0.5" & noise_sd == "0",
  .(mise = median(warp_mise, na.rm = TRUE)), by = method]
abs_hard <- dt[failure == FALSE & severity == "1" & noise_sd == "0.1",
  .(mise = median(warp_mise, na.rm = TRUE)), by = method]
abs_merge <- merge(abs_easy, abs_hard, by = "method", suffixes = c("_easy", "_hard"))
abs_merge[, abs_change := mise_hard - mise_easy]

cat("**Note on relative vs absolute degradation:** Large relative ratios can be misleading when baselines differ. Absolute median MISE change (sev=0.5/noise=0 to sev=1/noise=0.1):\n\n")
for (i in seq_len(nrow(abs_merge))) {
  cat(sprintf("- %s: %.2e to %.2e (change: %.2e)\n",
    mlabs[abs_merge$method[i]],
    abs_merge$mise_easy[i], abs_merge$mise_hard[i], abs_merge$abs_change[i]))
}
cat("\n")
```

```{r}
#| label: q6-summary-table
# Degradation ratios
noise_median <- ratios$noise[
  severity == "0.5" & is.finite(noise_ratio),
  .(noise_ratio = median(noise_ratio, na.rm = TRUE)),
  by = method
]
sev_median <- ratios$severity[
  noise_sd == "0" & is.finite(severity_ratio),
  .(severity_ratio = median(severity_ratio, na.rm = TRUE)),
  by = method
]

# Mean rank in easy condition and rank change to harder condition
rank_easy_noise <- dt[
  failure == FALSE & severity == "0.5" & noise_sd == "0",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method)
][, rank_easy := rank(mise), by = dgp
][, .(mean_rank_clean = round(mean(rank_easy), 1)), by = method]

rank_noisy <- dt[
  failure == FALSE & severity == "0.5" & noise_sd == "0.1",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method)
][, rank_noisy := rank(mise), by = dgp
][, .(mean_rank_noisy = round(mean(rank_noisy), 1)), by = method]

rank_easy_sev <- dt[
  failure == FALSE & severity == "0.5" & noise_sd == "0",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method)
][, rank_easy := rank(mise), by = dgp
][, .(mean_rank_easy = round(mean(rank_easy), 1)), by = method]

rank_hard_sev <- dt[
  failure == FALSE & severity == "1" & noise_sd == "0",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method)
][, rank_hard := rank(mise), by = dgp
][, .(mean_rank_hard = round(mean(rank_hard), 1)), by = method]

deg_summary <- Reduce(
  function(a, b) merge(a, b, by = "method", all = TRUE),
  list(rank_easy_noise, rank_noisy, noise_median, rank_easy_sev, rank_hard_sev, sev_median)
)
deg_summary[, noise_ratio := round(noise_ratio, 2)]
deg_summary[, severity_ratio := round(severity_ratio, 2)]
deg_summary[, rank_delta_noise := mean_rank_noisy - mean_rank_clean]
deg_summary[, rank_delta_sev := mean_rank_hard - mean_rank_easy]

knitr::kable(
  deg_summary[, .(method, mean_rank_clean, mean_rank_noisy, rank_delta_noise,
    noise_ratio, mean_rank_easy, mean_rank_hard, rank_delta_sev, severity_ratio)],
  col.names = c("Method",
    "Rank (clean)", "Rank (noisy)", "\u0394 rank", "MISE ratio",
    "Rank (low sev)", "Rank (high sev)", "\u0394 rank", "MISE ratio"),
  caption = "Degradation context. Rank columns: mean rank (across DGPs) in the easy condition and the harder condition; \u0394 rank: change in mean rank (positive = method drops in ranking). MISE ratio: ratio of median MISE under harder vs easier condition. Noise: severity=0.5; Severity: noise=0."
)
```

# Practical (Q7) {#sec-practical}

## Runtime {#sec-runtime}

```{r}
#| label: q7-runtime
#| fig-height: 5
#| fig-cap: "Distribution of time per curve (seconds) by method across all conditions. Log10 scale. Boxplots show all non-failed runs."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = time_per_curve, fill = method)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  labs(y = NULL, x = "Time per curve (seconds, log scale)") +
  theme_benchmark()
```

```{r}
#| label: q7-runtime-table
time_tbl <- dt[failure == FALSE, .(
  median = round(median(time_per_curve, na.rm = TRUE), 3),
  q25 = round(quantile(time_per_curve, 0.25, na.rm = TRUE), 3),
  q75 = round(quantile(time_per_curve, 0.75, na.rm = TRUE), 3)
), by = method][order(-median)]

knitr::kable(time_tbl,
  col.names = c("Method", "Median (s)", "Q25 (s)", "Q75 (s)"),
  caption = "Time per curve (seconds) across all conditions."
)
```

```{r}
#| label: q7-runtime-findings
#| results: asis
rt <- dt[failure == FALSE, .(
  med = median(time_per_curve, na.rm = TRUE)
), by = method][order(-med)]

cat("**Runtime findings:**\n\n")
for (i in seq_len(nrow(rt))) {
  cat(sprintf("- **%s**: %.3f s/curve\n", mlabs[rt$method[i]], rt$med[i]))
}

fastest <- rt[.N]
slowest <- rt[1]
ratio <- slowest$med / fastest$med
cat(sprintf(
  "\n%s is %.0f$\\times$ slower than %s (the fastest).\n\n",
  mlabs[slowest$method], ratio, mlabs[fastest$method]
))
```

## Failures {#sec-failures}

```{r}
#| label: q7-failures
fail_tbl <- failure_summary(dt)
fail_tbl[, rate := round(rate, 3)]
knitr::kable(
  fail_tbl[order(method, warp_type)],
  col.names = c("Method", "Warp type", "N total", "N failed", "Rate"),
  caption = "Failure rates by method and warp type."
)
```

```{r}
#| label: q7-error-analysis
#| results: asis
err <- dt[failure == TRUE & nchar(error_msg) > 0]
err[, error_class := fcase(
  grepl("duplicated values", error_msg), "Duplicated arg values in warped curves",
  grepl("computationally singular", error_msg), "Singular linear system",
  grepl("missing value where TRUE/FALSE", error_msg), "NA in logical comparison",
  default = "Other"
)]
err_summary <- err[, .N, by = .(method, error_class)][order(method, -N)]

cat("**Error analysis:**\n\n")
cat(sprintf("Total failures: %d out of %d runs (%.1f%%).\n\n",
  nrow(err), nrow(dt), 100 * nrow(err) / nrow(dt)))

cat("Failures are **concentrated in the FDA methods** (fda_default: ",
  sum(err$method == "fda_default"), "; fda_crit1: ",
  sum(err$method == "fda_crit1"), "). ",
  "The dominant failure modes are:\n\n", sep = "")

cat("1. **Duplicated arg values** (", sum(err$error_class ==
    "Duplicated arg values in warped curves"),
  " cases): The FDA registration algorithm produces warps that map ",
  "multiple input positions to the same output position, creating ",
  "duplicated argument values when constructing the warped curves. ",
  "This typically occurs on DGPs with complex warps or high-rank amplitude ",
  "where the iterative alignment overshoots.\n\n", sep = "")

cat("2. **Singular linear systems** (", sum(err$error_class ==
    "Singular linear system"),
  " cases): The B-spline basis matrix becomes numerically singular, ",
  "usually when the warping concentrates too many evaluation points ",
  "into a narrow interval.\n\n", sep = "")

cat("3. **NA in logical comparison** (", sum(err$error_class ==
    "NA in logical comparison"),
  " cases): Internal convergence check encounters NA values.\n\n", sep = "")

cat("SRVF, Affine, and Landmark methods have near-zero failure rates. ",
  "The 3 Landmark failures are edge cases on specific DGPs where ",
  "landmark detection finds insufficient features.\n\n")
```

# Data Completeness {#sec-completeness}

```{r}
#| label: completeness
comp <- dt[, .N, by = .(dgp, method)]
comp_wide <- dcast(comp, dgp ~ method, value.var = "N")
knitr::kable(comp_wide,
  caption = "Number of runs per DGP x method (target: 400 = 2 severity x 2 noise x 100 reps). Missing cells indicate batches that did not complete on the cluster."
)
```

```{r}
#| label: missing-batches
#| results: asis
expected <- CJ(
  dgp = levels(dt$dgp),
  method = levels(dt$method)
)
actual <- unique(dt[, .(dgp, method)])
missing <- expected[!actual, on = .(dgp, method)]
if (nrow(missing) > 0) {
  cat("**Missing DGP x method combinations** (", nrow(missing),
    " batches still running on LRZ):\n\n", sep = "")
  for (i in seq_len(nrow(missing))) {
    cat(sprintf("- %s / %s\n", missing$dgp[i], missing$method[i]))
  }
} else {
  cat("All DGP x method combinations present.\n")
}
```

# Summary {#sec-summary}

```{r}
#| label: summary-ranking
#| results: asis
cat("## Key findings\n\n")

# 1. Best method for warp recovery
best_by_wt <- dt[failure == FALSE,
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(method, warp_type)
][, .(best = method[which.min(mise)]), by = warp_type]
best_overall <- dt[failure == FALSE & warp_type != "affine",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = method][which.min(mise), method]
# Check: does it win on every DGP?
per_dgp <- dt[failure == FALSE & warp_type != "affine",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method)
][, .(best = method[which.min(mise)]), by = dgp]
n_wins <- sum(per_dgp$best == best_overall)
n_dgps_dp <- nrow(per_dgp)

cat(sprintf(
  "1. **%s is the best general-purpose method** for warp recovery on domain-preserving warps (simple + complex). It achieves the lowest pooled median warp MISE and ranks first on %d of %d individual DGPs. On the remaining DGPs, other methods are competitive.\n\n",
  mlabs[best_overall], n_wins, n_dgps_dp
))

# 2. Amplitude degradation
cat(sprintf(
  "2. **Amplitude variation degrades warp recovery for most methods**, but the pattern is not universal (see @sec-q2 for per-ladder details). %s retains the lowest absolute MISE under high-rank amplitude; its advantage is mainly its lower baseline.\n\n",
  mlabs[best_overall]
))

# 3. Affine puzzle
affine_best <- best_by_wt[warp_type == "affine", best]
cat(sprintf(
  "3. **The affine method underperforms on affine DGPs** despite being well-specified — %s achieves lower MISE. Plausible explanations include template estimation bias and weak signal at low severity, but these remain conjectures (see @sec-affine-puzzle).\n\n",
  mlabs[affine_best]
))

# 4. Failure rates
fail_rates <- dt[, .(rate = 100 * mean(failure)), by = method][order(-rate)]
top_fail <- fail_rates[1]
cat(sprintf(
  "4. **FDA methods have the highest failure rates** (%s: %.1f%%, %s: %.1f%%), driven by numerical issues in the iterative warping (duplicated arg values, singular systems). Some DGP-specific rates exceed 40%%.\n\n",
  mlabs[fail_rates$method[1]], fail_rates$rate[1],
  mlabs[fail_rates$method[2]], fail_rates$rate[2]
))

# 5. Noise/severity sensitivity
noise_worst <- noise_med[1]
sev_worst <- sev_med[1]
cat(sprintf(
  "5. **Noise sensitivity varies sharply**: %s is most noise-sensitive (%.1f$\\times$ degradation ratio), likely because it operates on SRSFs (derivatives). Severity hits %s hardest (%.1f$\\times$). See @sec-sensitivity.\n\n",
  mlabs[noise_worst$method], noise_worst$ratio,
  mlabs[sev_worst$method], sev_worst$ratio
))

# 6. Landmark
lm_rank <- dt[failure == FALSE,
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method)
][, rank := rank(mise), by = dgp
][method == "landmark_auto", .(mean_rank = mean(rank))]
lm_fail <- fail_rates[method == "landmark_auto", rate]
cat(sprintf(
  "6. **Landmark (auto) is competitive** (mean rank %.1f across DGPs) with near-zero failure rate (%.1f%%) and fast runtime. Its weakness is under-registration (registration ratio well below 1) and variability when landmark detection is unreliable.\n\n",
  lm_rank$mean_rank, lm_fail
))

# 7. AVR
avr_summary <- dt[amplitude != "none" & failure == FALSE & !is.na(amp_variance_ratio),
  .(avr = median(amp_variance_ratio, na.rm = TRUE)), by = method]
overreg_methods <- avr_summary[avr < 0.9, method]
cat("7. **Phase-amplitude separation varies by method**: ")
if (length(overreg_methods) > 0) {
  cat(sprintf(
    "%s show median AVR < 1, suggesting over-registration that absorbs amplitude into phase. ",
    paste(mlabs[overreg_methods], collapse = " and ")
  ))
}
underreg_methods <- avr_summary[avr > 1.1, method]
if (length(underreg_methods) > 0) {
  cat(sprintf(
    "%s show AVR > 1 (under-registration or template bias inflating residuals). ",
    paste(mlabs[underreg_methods], collapse = " and ")
  ))
}
near_one <- avr_summary[avr >= 0.9 & avr <= 1.1, method]
if (length(near_one) > 0) {
  cat(sprintf(
    "%s show near-calibrated AVR. ",
    paste(mlabs[near_one], collapse = " and ")
  ))
}
cat("Note that AVR > 1 can reflect either residual phase variation or template estimation error — these effects are confounded.\n\n")
```

```{r}
#| label: summary-table
rr_med <- dt[
  failure == FALSE & !is.na(registration_ratio_median),
  .(reg_ratio = round(median(registration_ratio_median, na.rm = TRUE), 2)),
  by = method
]
runtime <- dt[failure == FALSE,
  .(time_s = round(median(time_per_curve, na.rm = TRUE), 3)),
  by = method]
fail_rate <- dt[, .(fail_pct = round(100 * mean(failure), 1)), by = method]

summary_dt <- Reduce(function(a, b) merge(a, b, by = "method"), list(
  rr_med, runtime, fail_rate
))

knitr::kable(
  summary_dt[order(method)],
  col.names = c("Method", "Reg. ratio", "Time/curve (s)", "Failure %"),
  caption = "Summary per method. Reg. ratio: median registration ratio (1 = perfectly calibrated). Time/curve: median seconds. Failure %: overall failure rate."
)
```

# Session Info {.unnumbered}

```{r}
#| label: session-info
#| cache: false
sessionInfo()
```
