---
title: "Study D: Oracle Template"
subtitle: "Registration Benchmark v2"
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| cache: false
library(ggplot2)
library(data.table)
library(patchwork)

base_dir <- here::here()
source(file.path(base_dir, "sim-analyze.R"))

mcols <- method_colors()
mlabs <- method_labels()
dlabs <- dgp_short_labels()
```

```{r}
#| label: load-data
dt <- load_study_d()
```

```{r}
#| label: derived-data
oracle_keys <- c("dgp", "method", "condition")

warp_mode_ratio <- ratio_to_baseline(
  dt = dt,
  metric = "warp_mise",
  by = c(oracle_keys, "template_mode"),
  baseline_col = "template_mode",
  baseline_level = "estimated",
  stat = "median",
  value_name = "warp_mise",
  baseline_name = "estimated",
  ratio_name = "oracle_ratio",
  keep_baseline = TRUE
)
benefit_wide <- warp_mode_ratio[
  template_mode == "oracle",
  .(
    dgp, method, condition,
    estimated,
    oracle = warp_mise,
    pct_improvement = 100 * (1 - oracle_ratio)
  )
]

ae_mode_ratio <- ratio_to_baseline(
  dt = dt,
  metric = "alignment_error",
  by = c(oracle_keys, "template_mode"),
  baseline_col = "template_mode",
  baseline_level = "estimated",
  stat = "median",
  value_name = "alignment_error",
  baseline_name = "estimated",
  ratio_name = "oracle_ratio",
  keep_baseline = TRUE
)
ae_benefit_wide <- ae_mode_ratio[
  template_mode == "oracle",
  .(
    dgp, method, condition,
    estimated,
    oracle = alignment_error,
    pct_improvement = 100 * (1 - oracle_ratio)
  )
]

benefit_rep <- dt[
  failure == FALSE,
  .(dgp, method, condition, rep, template_mode, warp_mise)
]
benefit_rep_wide <- dcast(
  benefit_rep,
  dgp + method + condition + rep ~ template_mode,
  value.var = "warp_mise"
)
benefit_rep_wide[, pct_improvement := 100 * (estimated - oracle) / estimated]
cond_plot_dt <- benefit_rep_wide[, .(
  pct_improvement = median(pct_improvement, na.rm = TRUE),
  q25 = quantile(pct_improvement, 0.25, na.rm = TRUE),
  q75 = quantile(pct_improvement, 0.75, na.rm = TRUE)
), by = .(dgp, method, condition)]

ae_benefit_rep <- dt[
  failure == FALSE,
  .(dgp, method, condition, rep, template_mode, alignment_error)
]
ae_benefit_rep_wide <- dcast(
  ae_benefit_rep,
  dgp + method + condition + rep ~ template_mode,
  value.var = "alignment_error"
)
ae_benefit_rep_wide[, pct_improvement := 100 * (estimated - oracle) / estimated]
```

Study D decomposes total registration error into **template estimation error** vs
**warp estimation error**. When given the true template (oracle mode), template MISE = 0
and any remaining warp MISE is purely from the warp estimation algorithm.

**Design**: 6 DGPs $\times$ 2 template modes (oracle, estimated) $\times$ 2 conditions
$\times$ 5 methods = 120 cells $\times$ 50 reps = 6,000 runs.

Two conditions:

| Condition | Noise | Severity | Purpose |
|-----------|-------|----------|---------|
| Easy | 0.1 | 0.5 | Anchor: template estimation is relatively easy |
| Hard | 0.3 | 1.0 | Stress: high noise + large warps stress iterative estimation |

**Important caveat**: The oracle and estimated pipelines differ not only in template
source but also in the number of Procrustes iterations (see
@sec-limitations for details). The oracle comparison therefore reflects the
joint effect of template quality *and* iterative refinement, not a pure
template-only ablation.

```{r}
#| label: dgp-table
dgps_used <- c("D01", "D03", "D09", "D10", "D13", "D14")
knitr::kable(
  dgp_factors()[dgp %in% dgps_used],
  caption = "DGPs in Study D: 6 DGPs spanning both templates, all 3 warp types, and none/highrank amplitude."
)
```

# Sanity Checks {#sec-sanity}

## Template MISE in Oracle Mode

Template MISE should be exactly 0 (or numerically negligible) in oracle mode for
template-based methods, since the true template is passed directly. Landmark
registration ignores the supplied template (it aligns to detected landmarks),
so its template MISE is nonzero even in oracle mode — this is expected.

```{r}
#| label: sanity-template-mise
#| results: asis
tmise <- dt[failure == FALSE,
  .(template_mise = median(template_mise, na.rm = TRUE)),
  by = .(template_mode, method)
]
tmise_wide <- dcast(tmise, method ~ template_mode, value.var = "template_mise")
tmise_wide[, method := mlabs[method]]
cat("**Median template MISE by mode:**\n\n")
print(knitr::kable(tmise_wide, digits = 8))
oracle_max_no_lm <- dt[
  use_true_template == TRUE & failure == FALSE & method != "landmark_auto",
  max(template_mise, na.rm = TRUE)
]
cat(sprintf(
  "\nMax template MISE in oracle mode (excluding landmarks): %.2e (expect ~0).\n\n",
  oracle_max_no_lm
))
```

## Landmark Oracle Benefit

Landmark registration does not use a template estimate — it aligns to detected landmarks.
The oracle benefit for landmarks should be exactly 0 (serving as a negative control).

```{r}
#| label: sanity-landmark
#| results: asis
lm_dt <- dt[method == "landmark_auto" & failure == FALSE,
  .(warp_mise = median(warp_mise, na.rm = TRUE)),
  by = .(template_mode, condition)
]
lm_wide <- dcast(lm_dt, condition ~ template_mode, value.var = "warp_mise")
lm_wide[, diff := estimated - oracle]
cat("**Landmark warp MISE by condition and template mode:**\n\n")
print(knitr::kable(lm_wide, digits = 6))
```

# Failure Rates

```{r}
#| label: failures
#| results: asis
fail_dt <- dt[,
  .(n_fail = sum(failure), n_total = .N, rate = mean(failure)),
  by = .(method, template_mode)
]
fail_any <- fail_dt[n_fail > 0]
if (nrow(fail_any) == 0) {
  cat("No failures across all 6,000 runs.\n\n")
} else {
  cat("**Failure rates by method and template mode:**\n\n")
  print(knitr::kable(fail_dt[n_fail > 0], digits = 3))
}
```

# Oracle Benefit: Warp MISE {#sec-warp-benefit}

**Oracle benefit** = warp MISE (estimated template) $-$ warp MISE (oracle template).
Positive values mean the oracle helps (lower warp error with true template).
Negative values mean the estimated-template pipeline produces *lower* warp error
than the oracle pipeline — see @sec-cc-default-negative for interpretation.

## Overview Heatmap

```{r}
#| label: oracle-benefit-heatmap
#| fig-height: 7
#| fig-cap: "Oracle benefit (% improvement in warp MISE) by method and DGP, for each condition. Red = oracle helps, blue = oracle hurts. Values beyond the color range are squished to the boundary but printed at their true value."
ggplot(benefit_wide, aes(x = method, y = dgp, fill = pct_improvement)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_label(aes(label = sprintf("%.0f%%", pct_improvement)),
    size = 2.3, fill = "white", alpha = 0.85,
    label.size = 0, label.padding = unit(1.5, "pt")) +
  facet_wrap(~condition) +
  scale_x_discrete(labels = mlabs) +
  scale_y_discrete(labels = dlabs) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "grey95", high = "#B2182B",
    midpoint = 0, name = "Oracle\nbenefit (%)",
    limits = c(-170, 100),
    oob = scales::squish
  ) +
  labs(x = NULL, y = NULL,
    subtitle = "Oracle benefit: % improvement in warp MISE when given true template") +
  theme_benchmark()
```

## Per-Method Oracle Benefit

```{r}
#| label: oracle-benefit-by-method
#| fig-height: 6
#| fig-cap: "Distribution of oracle benefit (% improvement) by method (excluding landmarks). Each dot is one DGP-condition cell. The crossbar shows the median."
ggplot(benefit_wide[method != "landmark_auto"],
  aes(x = pct_improvement, y = method, color = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_jitter(height = 0.15, size = 2.5, alpha = 0.7) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4,
    color = "black", linewidth = 0.4) +
  facet_wrap(~condition) +
  scale_color_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  labs(x = "Oracle benefit (% improvement in warp MISE)", y = NULL) +
  theme_benchmark()
```

```{r}
#| label: oracle-benefit-findings
#| results: asis
method_summary <- benefit_wide[method != "landmark_auto",
  .(
    mean_pct = mean(pct_improvement, na.rm = TRUE),
    se_pct = if (.N > 1) sd(pct_improvement, na.rm = TRUE) / sqrt(.N) else 0,
    median_pct = median(pct_improvement, na.rm = TRUE),
    min_pct = min(pct_improvement, na.rm = TRUE),
    max_pct = max(pct_improvement, na.rm = TRUE)
  ),
  by = .(method, condition)
]

cat("**Oracle benefit by method** (% improvement in warp MISE):\n\n")
for (cond in c("easy", "hard")) {
  cat(sprintf("*%s condition (noise=%s, severity=%s):*\n\n",
    cond,
    ifelse(cond == "easy", "0.1", "0.3"),
    ifelse(cond == "easy", "0.5", "1.0")))
  sub <- method_summary[condition == cond][order(-median_pct)]
  for (i in seq_len(nrow(sub))) {
    cat(sprintf("- **%s**: mean %.0f ± %.0f pp; median %.0f%% [%.0f%%, %.0f%%]\n",
      mlabs[sub$method[i]],
      sub$mean_pct[i], sub$se_pct[i],
      sub$median_pct[i], sub$min_pct[i], sub$max_pct[i]))
  }
  cat("\n")
}
```

## CC (FPC1 cc): Negative Oracle Benefit {#sec-cc-default-negative}

```{r}
#| label: fda-default-negative
#| results: asis
fda_def <- benefit_wide[method == "cc_default"]
n_negative <- sum(fda_def$pct_improvement < 0)
n_total_fda <- nrow(fda_def)
worst <- fda_def[which.min(pct_improvement)]
fda_def_by_cond <- fda_def[, .(
  n_negative = sum(pct_improvement < 0),
  n_total = .N,
  mean_pct = mean(pct_improvement),
  se_pct = if (.N > 1) sd(pct_improvement) / sqrt(.N) else 0,
  median_pct = median(pct_improvement)
), by = condition]

cat(sprintf(
  "**cc_default** shows negative oracle benefit in %d of %d DGP-condition cells ",
  n_negative, n_total_fda
))
cat(sprintf(
  "(worst: %s %s condition, %.0f%%).\n\n",
  worst$dgp, worst$condition, worst$pct_improvement
))
cat("By condition, the hard regime is more uniformly negative, while the easy regime is mixed and skewed:\n\n")
for (i in seq_len(nrow(fda_def_by_cond))) {
  r <- fda_def_by_cond[i]
  cat(sprintf(
    "- **%s**: %d/%d cells negative; mean oracle benefit %.0f ± %.0f pp; median %.0f%%\n",
    r$condition, r$n_negative, r$n_total,
    r$mean_pct, r$se_pct, r$median_pct
  ))
}
cat("\n")

cat("This means the estimated-template pipeline produces *lower* warp MISE than the ",
  "oracle pipeline. Two factors contribute:\n\n")
cat("1. **Iteration confound** (@sec-limitations): The estimated pipeline runs ",
  "10 Procrustes iterations (jointly refining template and warps), while the oracle ",
  "pipeline runs only 1 (since `tf` sets `max_iter = 1` when a template is provided). ",
  "The additional iterations improve warp estimates even though the template is less accurate.\n\n")
cat("2. **Objective mismatch**: cc_default uses criterion 2 (eigenfunction-based), ",
  "which may converge to a biased template that happens to be easier for its ",
  "warping objective to fit. The iteratively estimated template is co-adapted to ",
  "the warping criterion, whereas the true template is not.\n\n")

# Compare with cc_crit1
crit1_neg <- benefit_wide[method == "cc_crit1", sum(pct_improvement < 0)]
cat(sprintf(
  "By contrast, **cc_crit1** (criterion 1, ISE-based) shows negative oracle benefit ",
  "in only %d of %d cells, suggesting criterion 2 is more sensitive to the ",
  "iteration/co-adaptation effect.\n",
  crit1_neg, nrow(benefit_wide[method == "cc_crit1"])
))
```

## Condition Effect: Easy vs Hard

```{r}
#| label: condition-effect
#| fig-height: 6
#| fig-cap: "Oracle benefit by condition. Points show median paired oracle benefit and vertical bars show IQRs across replications for each method-DGP combination."
ggplot(cond_plot_dt[method != "landmark_auto"],
  aes(x = condition, y = pct_improvement, group = interaction(method, dgp),
    color = method)) +
  geom_line(alpha = 0.5) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.08, alpha = 0.3) +
  facet_wrap(~method, labeller = labeller(method = mlabs)) +
  scale_color_manual(values = mcols, guide = "none") +
  labs(x = "Condition", y = "Oracle benefit (%)") +
  theme_benchmark()
```

```{r}
#| label: condition-findings
#| results: asis
cond_change <- dcast(method_summary, method ~ condition, value.var = "mean_pct")
cond_se <- dcast(method_summary, method ~ condition, value.var = "se_pct")
cond_change[, change := hard - easy]

cat("**Does oracle benefit grow from easy to hard?**\n\n")
for (i in seq_len(nrow(cond_change))) {
  m <- cond_change$method[i]
  cat(sprintf("- **%s**: easy %.0f ± %.0f pp → hard %.0f ± %.0f pp (%+.0f pp)\n",
    mlabs[m],
    cond_change$easy[i], cond_se$easy[i],
    cond_change$hard[i], cond_se$hard[i],
    cond_change$change[i]
  ))
}
cat("\n")

# Check if effect is method-dependent
n_pos <- sum(cond_change$change > 0)
n_neg <- sum(cond_change$change < 0)
cat(sprintf(
  "The condition effect is method-dependent: %d methods show increased oracle ",
  n_pos
))
cat(sprintf(
  "benefit under the hard condition, while %d show decreased (or more negative) ",
  n_neg
))
cat("benefit. For SRVF, harder conditions amplify the oracle benefit as expected. ",
  "For cc_default, the negative oracle benefit persists or worsens, consistent ",
  "with the iteration confound (@sec-limitations).\n")
```

# Warp MISE: Oracle vs Estimated {#sec-warp-comparison}

Direct comparison of warp MISE under oracle and estimated template modes.

```{r}
#| label: warp-comparison
#| fig-height: 8
#| fig-cap: "Warp MISE by template mode (oracle vs estimated), faceted by method and condition. Boxplots pool over all DGPs. Landmarks excluded (no template dependence)."
ggplot(
  dt[failure == FALSE & method != "landmark_auto"],
  aes(y = template_mode, x = warp_mise, fill = template_mode)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_grid(method ~ condition,
    scales = "free_x",
    labeller = labeller(method = mlabs)) +
  scale_fill_manual(
    values = c(estimated = "#F0E442", oracle = "#56B4E9"),
    guide = "none"
  ) +
  scale_x_log10() +
  labs(y = NULL, x = "Warp MISE (log scale)") +
  theme_benchmark()
```

## Per-DGP Comparison

```{r}
#| label: dgp-comparison
#| fig-height: 8
#| fig-cap: "Per-DGP warp MISE comparison (template-based methods only). Yellow = estimated, blue = oracle. Methods where the gap is large benefit most from knowing the true template."
dgp_comp <- dt[failure == FALSE & method != "landmark_auto",
  .(
    median = median(warp_mise, na.rm = TRUE),
    q25 = quantile(warp_mise, 0.25, na.rm = TRUE),
    q75 = quantile(warp_mise, 0.75, na.rm = TRUE)
  ),
  by = .(dgp, method, template_mode, condition)
]

ggplot(dgp_comp, aes(x = template_mode, y = median, fill = template_mode)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.2, alpha = 0.5) +
  facet_grid(dgp ~ method + condition,
    scales = "free_y",
    labeller = labeller(method = mlabs, dgp = dlabs)) +
  scale_fill_manual(
    values = c(estimated = "#F0E442", oracle = "#56B4E9"),
    name = "Template"
  ) +
  scale_y_log10() +
  labs(x = NULL, y = "Warp MISE (log scale)") +
  theme_benchmark(base_size = 8) +
  theme(axis.text.x = element_blank())
```

# Oracle Benefit: Alignment Error {#sec-alignment-error}

Alignment error measures $\text{mean}_i \int (\hat{x}_i^{\text{aligned}} - \tilde{x}_i)^2\,dt$
— the practical downstream impact of registration quality. Does knowing the true
template also improve alignment of registered curves?

## Overview Heatmap

```{r}
#| label: ae-oracle-heatmap
#| fig-height: 7
#| fig-cap: "Oracle benefit (% improvement in alignment error) by method and DGP. Red = oracle helps, blue = oracle hurts."
ggplot(ae_benefit_wide, aes(x = method, y = dgp, fill = pct_improvement)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_label(aes(label = sprintf("%.0f%%", pct_improvement)),
    size = 2.3, fill = "white", alpha = 0.85,
    label.size = 0, label.padding = unit(1.5, "pt")) +
  facet_wrap(~condition) +
  scale_x_discrete(labels = mlabs) +
  scale_y_discrete(labels = dlabs) +
  scale_fill_gradient2(
    low = "#2166AC", mid = "grey95", high = "#B2182B",
    midpoint = 0, name = "Oracle\nbenefit (%)",
    limits = c(-170, 100),
    oob = scales::squish
  ) +
  labs(x = NULL, y = NULL,
    subtitle = "Oracle benefit: % improvement in alignment error when given true template") +
  theme_benchmark()
```

## Per-Method Comparison

```{r}
#| label: ae-oracle-by-method
#| fig-height: 6
#| fig-cap: "Distribution of oracle benefit for alignment error by method (excluding landmarks). Each dot is one DGP-condition cell."
ggplot(ae_benefit_wide[method != "landmark_auto"],
  aes(x = pct_improvement, y = method, color = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_jitter(height = 0.15, size = 2.5, alpha = 0.7) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4,
    color = "black", linewidth = 0.4) +
  facet_wrap(~condition) +
  scale_color_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  labs(x = "Oracle benefit (% improvement in alignment error)", y = NULL) +
  theme_benchmark()
```

```{r}
#| label: ae-oracle-findings
#| results: asis
ae_method_summary <- ae_benefit_wide[method != "landmark_auto",
  .(
    mean_pct = mean(pct_improvement, na.rm = TRUE),
    se_pct = if (.N > 1) sd(pct_improvement, na.rm = TRUE) / sqrt(.N) else 0,
    median_pct = median(pct_improvement, na.rm = TRUE),
    min_pct = min(pct_improvement, na.rm = TRUE),
    max_pct = max(pct_improvement, na.rm = TRUE)
  ),
  by = .(method, condition)
]

cat("**Alignment error oracle benefit by method:**\n\n")
for (cond in c("easy", "hard")) {
  cat(sprintf("*%s condition:*\n\n", cond))
  sub <- ae_method_summary[condition == cond][order(-median_pct)]
  for (i in seq_len(nrow(sub))) {
    cat(sprintf("- **%s**: mean %.0f ± %.0f pp; median %.0f%% [%.0f%%, %.0f%%]\n",
      mlabs[sub$method[i]],
      sub$mean_pct[i], sub$se_pct[i],
      sub$median_pct[i], sub$min_pct[i], sub$max_pct[i]))
  }
  cat("\n")
}

# Compare with warp MISE oracle benefit
warp_benefit_sub <- benefit_wide[method != "landmark_auto",
  .(method, dgp, condition, warp_pct = pct_improvement)]
ae_benefit_sub <- ae_benefit_wide[method != "landmark_auto",
  .(method, dgp, condition, ae_pct = pct_improvement)]
both <- merge(warp_benefit_sub, ae_benefit_sub,
  by = c("method", "dgp", "condition"))
r_benefit <- cor(both$warp_pct, both$ae_pct, use = "complete.obs")
cat(sprintf(
  "Correlation between warp MISE and alignment error oracle benefit: r = %.2f. ",
  r_benefit))
if (r_benefit > 0.9) {
  cat("The oracle template produces nearly identical benefit patterns for both metrics.\n\n")
} else if (r_benefit > 0.7) {
  cat("The benefit patterns are strongly correlated — methods that gain warp accuracy ",
    "from the oracle also gain alignment quality, with some exceptions.\n\n")
} else {
  cat("The metrics show moderate agreement — oracle template benefits differ ",
    "between warp recovery and alignment quality.\n\n")
}

# Flag any sign disagreements (oracle helps warp but hurts alignment or vice versa)
sign_disagree <- both[sign(warp_pct) != sign(ae_pct)]
if (nrow(sign_disagree) > 0) {
  cat(sprintf("Sign disagreements (oracle helps one metric but hurts the other): %d/%d cells.\n\n",
    nrow(sign_disagree), nrow(both)))
}

# Key insight: cc_default warp vs alignment divergence
fda_warp <- benefit_wide[method == "cc_default",
  .(warp_med = median(pct_improvement, na.rm = TRUE)), by = condition]
fda_ae <- ae_benefit_wide[method == "cc_default",
  .(ae_med = median(pct_improvement, na.rm = TRUE)), by = condition]
fda_both <- merge(fda_warp, fda_ae, by = "condition")
if (any(sign(fda_both$warp_med) != sign(fda_both$ae_med))) {
  cat("**Key insight — cc_default metric divergence**: The warp MISE oracle ",
    "benefit is negative (estimated pipeline produces better warps due to ",
    "iterative refinement), but the alignment error oracle benefit is positive. ",
    "This means the true template partially compensates for the worse oracle-mode ",
    "warps when computing aligned curves ($\\hat{x}^{\\text{aligned}} = \\hat{\\mu} ",
    "\\circ \\hat{h}$): a perfect template offsets some warp error, while the ",
    "estimated pipeline's better warps cannot fully compensate for its worse ",
    "template. The two metrics thus reveal different facets of the iteration ",
    "confound (@sec-limitations).\n")
} else {
  cat(sprintf(
    "**cc_default** shows consistent sign across metrics: warp MISE median %.0f%%/%.0f%% ",
    fda_both[condition == "easy", warp_med], fda_both[condition == "hard", warp_med]))
  cat(sprintf(
    "vs alignment error %.0f%%/%.0f%% (easy/hard).\n",
    fda_both[condition == "easy", ae_med], fda_both[condition == "hard", ae_med]))
}
```

# Template Estimation Quality {#sec-template-quality}

How good are the template estimates? This section examines within-method
associations between template quality and warp quality in estimated mode.

```{r}
#| label: template-quality
#| fig-height: 6
#| fig-cap: "Template MISE in estimated mode, by method and DGP (template-based methods only)."
est_only <- dt[use_true_template == FALSE & failure == FALSE &
  method != "landmark_auto"]

ggplot(est_only, aes(y = method, x = template_mise, fill = method)) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_grid(dgp ~ condition,
    scales = "free_x",
    labeller = labeller(dgp = dlabs)) +
  scale_fill_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  labs(y = NULL, x = "Template MISE (log scale, estimated mode only)") +
  theme_benchmark(base_size = 9)
```

```{r}
#| label: template-warp-correlation
#| fig-height: 5
#| fig-cap: "Association between template MISE and warp MISE in estimated mode. Each dot is the cell median and horizontal/vertical bars show IQRs over the 50 replications for one (DGP, method, condition) cell."
cor_dt <- est_only[,
  .(
    template_mise = median(template_mise, na.rm = TRUE),
    warp_mise = median(warp_mise, na.rm = TRUE),
    tmise_q25 = quantile(template_mise, 0.25, na.rm = TRUE),
    tmise_q75 = quantile(template_mise, 0.75, na.rm = TRUE),
    warp_q25 = quantile(warp_mise, 0.25, na.rm = TRUE),
    warp_q75 = quantile(warp_mise, 0.75, na.rm = TRUE)
  ),
  by = .(dgp, method, condition)
]

ggplot(cor_dt, aes(x = template_mise, y = warp_mise, color = method)) +
  geom_point(size = 2.5) +
  geom_errorbar(aes(ymin = warp_q25, ymax = warp_q75), width = 0, alpha = 0.3) +
  geom_errorbarh(aes(xmin = tmise_q25, xmax = tmise_q75), height = 0, alpha = 0.3) +
  facet_wrap(~condition) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(values = mcols, labels = mlabs, name = "Method") +
  labs(x = "Template MISE (log scale)", y = "Warp MISE (log scale)") +
  theme_benchmark()
```

```{r}
#| label: template-warp-cor-findings
#| results: asis
# Overall correlation
cors <- cor_dt[,
  .(r = cor(log(template_mise), log(warp_mise), use = "complete.obs")),
  by = condition
]
cat("**Correlation between log(template MISE) and log(warp MISE) ",
  "(estimated mode, excluding landmarks):**\n\n")
for (i in seq_len(nrow(cors))) {
  cat(sprintf("- %s condition: r = %.2f\n", cors$condition[i], cors$r[i]))
}
cat("\n")

# Per-method correlations
cors_method <- cor_dt[,
  .(r = if (.N >= 3) cor(log(template_mise), log(warp_mise),
    use = "complete.obs") else NA_real_),
  by = .(method, condition)
]
cors_method <- cors_method[!is.na(r)]
if (nrow(cors_method) > 0) {
  cat("**Within-method correlations** (each method has 6 DGP points):\n\n")
  for (i in seq_len(nrow(cors_method))) {
    cat(sprintf("- %s, %s: r = %.2f\n",
      mlabs[cors_method$method[i]],
      cors_method$condition[i],
      cors_method$r[i]))
  }
  cat("\n")
}

cat("The overall positive association is partly driven by cross-method ",
  "differences in both template and warp quality. Within-method correlations ",
  "are more informative for assessing whether template quality causally ",
  "affects warp recovery for a given method.\n")
```

# Limitations {#sec-limitations}

```{r}
#| label: limitations
#| results: asis
cat("## max_iter Confound {#sec-max-iter}\n\n")
cat("The oracle and estimated pipelines differ in the number of Procrustes ",
  "iterations, creating a confound:\n\n",
  "| Mode | max_iter | Reason |\n",
  "|------|----------|--------|\n",
  "| Estimated | 10 | Set in `sim-methods.R` for cc/affine methods |\n",
  "| Oracle | 1 | `tf_register()` internally sets `max_iter = 1` when a template is provided |\n\n",
  sep = "")
cat("Each Procrustes iteration re-estimates warps (and, in estimated mode, the ",
  "template). Thus the estimated pipeline gets 10 rounds of warp refinement ",
  "while the oracle pipeline gets only 1. This means:\n\n")
cat("- **Positive oracle benefit** (e.g., SRVF) reflects cases where template ",
  "accuracy outweighs the loss of iterative refinement.\n")
cat("- **Negative oracle benefit** (e.g., cc_default) may partly reflect the ",
  "estimated pipeline's advantage from 10 iterations of joint ",
  "template-warp optimization.\n\n")
cat("A clean ablation (true vs estimated template, both with matched iterations) ",
  "would require modifying `tf_register()` to allow `max_iter > 1` with a supplied ",
  "template. This is left for a future study revision.\n\n")

cat("## Scope of Conclusions\n\n")
cat("Given the max_iter confound, the oracle comparison is best interpreted as: ",
  "**estimated template with iterative refinement** vs **true template with ",
  "single-pass alignment**. This is a practically relevant comparison (it reflects ",
  "what happens when a user supplies a template to `tf_register()`), but it is not ",
  "a pure template-quality ablation.\n\n")
cat("Conclusions about the effect of template quality alone should be drawn ",
  "primarily from SRVF, where the oracle benefit is consistently positive and large ",
  "— indicating that template accuracy dominates even without iterative refinement.\n")
```

# Summary {#sec-summary}

```{r}
#| label: summary
#| results: asis
# Final ranking by oracle benefit
final <- benefit_wide[method != "landmark_auto",
  .(
    mean_pct = mean(pct_improvement, na.rm = TRUE),
    se_pct = if (.N > 1) sd(pct_improvement, na.rm = TRUE) / sqrt(.N) else 0
  ),
  by = method
][order(-mean_pct)]

cat("**Key findings** (see @sec-limitations for caveats on interpretation):\n\n")
cat("1. **Oracle benefit ranking** (mean % improvement ± SE in warp MISE, pooled over conditions and DGPs):\n\n")
for (i in seq_len(nrow(final))) {
  cat(sprintf("   - %s: %.0f ± %.0f pp\n",
    mlabs[final$method[i]], final$mean_pct[i], final$se_pct[i]))
}
cat("\n")

# SRVF-specific finding
srvf_benefit <- benefit_wide[method == "srvf",
  .(
    mean_pct = mean(pct_improvement, na.rm = TRUE),
    se_pct = if (.N > 1) sd(pct_improvement, na.rm = TRUE) / sqrt(.N) else 0
  ),
  by = condition
]
cat(sprintf(
  "2. **SRVF benefits most**: Mean oracle benefit %.0f ± %.0f pp (easy) → %.0f ± %.0f pp (hard). ",
  srvf_benefit[condition == "easy", mean_pct],
  srvf_benefit[condition == "easy", se_pct],
  srvf_benefit[condition == "hard", mean_pct],
  srvf_benefit[condition == "hard", se_pct]
))
cat("SRVF consistently benefits from the true template, with the advantage ",
  "growing under harder conditions. This is the cleanest evidence that template ",
  "quality matters for warp recovery.\n\n")

# cc_default finding
fda_def_benefit <- benefit_wide[method == "cc_default",
  .(
    mean_pct = mean(pct_improvement, na.rm = TRUE),
    se_pct = if (.N > 1) sd(pct_improvement, na.rm = TRUE) / sqrt(.N) else 0,
    median_pct = median(pct_improvement, na.rm = TRUE)
  ),
  by = condition
]
cat(sprintf(
  "3. **cc_default is condition-dependent**: mean oracle benefit %.0f ± %.0f pp in the easy condition but %.0f ± %.0f pp in the hard condition (medians %.0f%% and %.0f%%). ",
  fda_def_benefit[condition == "easy", mean_pct],
  fda_def_benefit[condition == "easy", se_pct],
  fda_def_benefit[condition == "hard", mean_pct],
  fda_def_benefit[condition == "hard", se_pct],
  fda_def_benefit[condition == "easy", median_pct],
  fda_def_benefit[condition == "hard", median_pct]
))
cat("The estimated-template pipeline (with 10 Procrustes iterations) outperforms ",
  "the oracle pipeline mainly in the hard regime, suggesting that iterative ",
  "refinement and/or co-adaptation of template and warps under criterion 2 compensates ",
  "for template estimation error there.\n\n")

cat("4. **Landmark sanity**: Oracle benefit is exactly 0%% for landmarks (no template used), validating the experimental design.\n\n")

# DGP with largest oracle benefit
max_dgp <- benefit_wide[method != "landmark_auto"][which.max(pct_improvement)]
cat(sprintf(
  "5. **Largest oracle benefit**: %s on %s (%s condition): %.0f%% improvement.\n\n",
  mlabs[max_dgp$method], max_dgp$dgp, max_dgp$condition, max_dgp$pct_improvement
))

cat("6. **Implication for `tf_register()`**: The asymmetry between methods suggests ",
  "that (a) SRVF registration would benefit most from better template initialization ",
  "or multi-resolution estimation, and (b) cc_default's criterion 2 benefits from ",
  "iterative joint optimization, making the number of Procrustes iterations an ",
  "important tuning parameter for this method.\n\n")

# Alignment error finding
ae_final <- ae_benefit_wide[method != "landmark_auto",
  .(
    mean_pct = mean(pct_improvement, na.rm = TRUE),
    se_pct = if (.N > 1) sd(pct_improvement, na.rm = TRUE) / sqrt(.N) else 0
  ),
  by = method
][order(-mean_pct)]
cat("7. **Alignment error tells a different story** (@sec-alignment-error): ",
  "Oracle benefit is positive for all methods when measured by alignment error, ",
  "even cc_default. ")
fda_ae_summary <- ae_final[method == "cc_default"]
cat(sprintf(
  "cc_default's alignment error oracle benefit is %.0f ± %.0f pp (positive), while its ",
  fda_ae_summary$mean_pct, fda_ae_summary$se_pct))
cat("warp MISE oracle benefit is negative. The true template compensates for worse ",
  "oracle-mode warps when computing aligned curves, revealing that the two metrics ",
  "capture different aspects of the iteration confound.\n")
```
