---
title: "Registration Benchmark v2 — Study C: Grid Resolution"
subtitle: "Internal Report — tf Package"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  cache: true
  warning: false
  message: false
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| cache: false
library(ggplot2)
library(data.table)
library(patchwork)

base_dir <- here::here()
source(file.path(base_dir, "sim-analyze.R"))

mcols <- method_colors()
mlabs <- method_labels()
dlabs <- dgp_short_labels()
```

```{r}
#| label: load-data
dt <- load_study_c()
```

Study C tests whether **grid resolution** is a confound for method comparison.
SRVF operates on SRSFs (numerical derivatives) which amplify discretization artifacts
at finer grids. A preliminary pilot (job 5131384) showed SRVF warp MISE 1.82× worse at
$n_\text{grid}=201$ vs 101.

**Design**: 5 DGPs $\times$ 3 grid sizes (51, 101, 201) $\times$ 3 noise (0, 0.1, 0.3)
$\times$ 5 methods $\times$ severity=1.0 = 225 cells $\times$ 50 reps = 11,250 runs.

```{r}
#| label: dgp-table
dgps_used <- c("D01", "D03", "D04", "D09", "D12")
knitr::kable(
  dgp_factors()[dgp %in% dgps_used],
  caption = "DGPs in Study C: covering both templates, all 3 warp types, and none/highrank amplitude."
)
```

# Overview {#sec-overview}

## Failure Rates

```{r}
#| label: failures
#| results: asis
fail_dt <- dt[,
  .(n_fail = sum(failure), n_total = .N, rate = mean(failure)),
  by = .(method, n_grid)
]
fail_any <- fail_dt[n_fail > 0]
if (nrow(fail_any) == 0) {
  cat("No failures across all 11,250 runs.\n\n")
} else {
  cat("**Failure rates by method and grid:**\n\n")
  print(knitr::kable(fail_dt[n_fail > 0]))
}
```

## Warp MISE by Grid $\times$ Method

The central question: does grid resolution systematically affect method performance?

```{r}
#| label: warp-mise-grid-method
#| fig-height: 7
#| fig-cap: "Warp MISE by grid size and method, faceted by noise level. Each boxplot pools over all 5 DGPs and 50 reps."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = warp_mise, fill = n_grid)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, position = "dodge") +
  facet_wrap(~noise_sd, labeller = labeller(noise_sd = function(x) paste0("noise=", x))) +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  scale_fill_brewer(palette = "Set2", name = "Grid size") +
  labs(y = NULL, x = "Warp MISE (log scale)") +
  theme_benchmark()
```

# Grid Degradation Ratios {#sec-ratios}

Grid degradation ratios compare each grid size to the baseline $n_\text{grid}=101$.
Values > 1 mean worse performance at that grid; < 1 means better.

```{r}
#| label: grid-ratios
#| fig-height: 8
#| fig-cap: "Warp MISE ratio relative to grid=101, per method and noise level. Each labeled point is one DGP; horizontal bars show the IQR of paired per-rep ratios and the crossbar shows the median across DGPs. Values > 1 (right of dashed line) indicate degradation."
ratio_rep <- dcast(
  dt[failure == FALSE, .(dgp, method, noise_sd, rep, n_grid, warp_mise)],
  dgp + method + noise_sd + rep ~ n_grid,
  value.var = "warp_mise"
)
ratio_rep[, ratio_51 := `51` / `101`]
ratio_rep[, ratio_201 := `201` / `101`]
ratio_rep <- melt(
  ratio_rep,
  id.vars = c("dgp", "method", "noise_sd", "rep"),
  measure.vars = c("ratio_51", "ratio_201"),
  variable.name = "n_grid",
  value.name = "ratio"
)
ratio_rep[, n_grid := sub("^ratio_", "", n_grid)]
ratio_dt <- ratio_rep[, .(
  ratio = median(ratio, na.rm = TRUE),
  q25 = quantile(ratio, 0.25, na.rm = TRUE),
  q75 = quantile(ratio, 0.75, na.rm = TRUE)
), by = .(dgp, method, n_grid, noise_sd)]

ggplot(ratio_dt, aes(x = ratio, y = method, color = method)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_point(size = 2, alpha = 0.7) +
  geom_errorbarh(aes(xmin = q25, xmax = q75), height = 0.15, alpha = 0.3) +
  geom_text(aes(label = dgp), size = 3, vjust = -0.8, show.legend = FALSE) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4,
    color = "black", linewidth = 0.4) +
  facet_grid(n_grid ~ noise_sd,
    labeller = labeller(
      noise_sd = function(x) paste0("noise=", x),
      n_grid = function(x) paste0("grid=", x)
    )
  ) +
  scale_color_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_continuous(trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("1/4", "1/2", "1", "2", "4")
  ) +
  labs(x = expression("Warp MISE ratio vs grid=101 (" * log[2] * " scale)"), y = NULL) +
  theme_benchmark()
```

```{r}
#| label: grid-ratio-findings
#| results: asis
# Median ratio per method x grid x noise
ratio_by_noise <- ratio_dt[,
  .(median_ratio = median(ratio, na.rm = TRUE)),
  by = .(method, n_grid, noise_sd)
]
ratio_uncertainty <- mean_se_table(ratio_dt, "ratio", c("method", "n_grid", "noise_sd"))

# SRVF noise-dependent reversal (the key finding)
srvf_rats <- ratio_by_noise[method == "srvf"]
srvf_unc <- ratio_uncertainty[method == "srvf"]
cat("**Key finding: SRVF grid sensitivity reverses with noise level.**\n\n")
tbl_lines <- c(
  "SRVF warp MISE ratios (vs grid=101):\n",
  "| Noise | grid=51 | grid=201 |",
  "|-------|---------|----------|"
)
for (ns in c("0", "0.1", "0.3")) {
  r51 <- srvf_rats[n_grid == "51" & noise_sd == ns, median_ratio]
  r201 <- srvf_rats[n_grid == "201" & noise_sd == ns, median_ratio]
  se51 <- srvf_unc[n_grid == "51" & noise_sd == ns, se]
  se201 <- srvf_unc[n_grid == "201" & noise_sd == ns, se]
  tbl_lines <- c(tbl_lines, sprintf(
    "| %s | %.2f ± %.2f | %.2f ± %.2f |",
    ns, r51, se51, r201, se201
  ))
}
cat(paste(tbl_lines, collapse = "\n"), "\n\n")
cat("At noise=0, coarser grid *hurts* SRVF — DP alignment on 51 points is too ",
  "imprecise to resolve the warp. At noise>0, finer grid *hurts* SRVF — ",
  "numerical differentiation amplifies noise at 201 points, acting as ",
  "de facto anti-regularization.\n\n")

# All methods summary
cat("**All methods** (median ratio, pooled over DGPs and noise — ",
  "descriptive only; SRVF pooled medians mask the noise-dependent reversal above):\n\n")
ratio_summary <- ratio_dt[,
  .(
    median_ratio = median(ratio, na.rm = TRUE),
    mean_ratio = mean(ratio, na.rm = TRUE),
    se_ratio = if (.N > 1) sd(ratio, na.rm = TRUE) / sqrt(.N) else 0,
    min_ratio = min(ratio, na.rm = TRUE),
    max_ratio = max(ratio, na.rm = TRUE)
  ),
  by = .(method, n_grid)
]
for (g in sort(unique(as.character(ratio_summary$n_grid)))) {
  cat(sprintf("*Grid = %s:*\n\n", g))
  sub <- ratio_summary[n_grid == g][order(median_ratio)]
  for (i in seq_len(nrow(sub))) {
    cat(sprintf("- **%s**: median %.2f; mean %.2f ± %.2f; range [%.2f, %.2f]\n",
      mlabs[sub$method[i]], sub$median_ratio[i],
      sub$mean_ratio[i], sub$se_ratio[i],
      sub$min_ratio[i], sub$max_ratio[i]))
  }
  cat("\n")
}
```

# Template MISE by Grid {#sec-template}

```{r}
#| label: template-mise-grid
#| fig-height: 7
#| fig-cap: "Template MISE by grid size and method, faceted by noise level."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = template_mise, fill = n_grid)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, position = "dodge") +
  facet_wrap(~noise_sd, labeller = labeller(noise_sd = function(x) paste0("noise=", x))) +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  scale_fill_brewer(palette = "Set2", name = "Grid size") +
  labs(y = NULL, x = "Template MISE (log scale)") +
  theme_benchmark()
```

```{r}
#| label: template-ratio
#| fig-height: 8
#| fig-cap: "Template MISE ratio relative to grid=101. Each labeled point is one DGP and horizontal bars show the IQR of paired per-rep ratios. Values < 1 mean that grid improves template estimation."
t_ratio_rep <- dcast(
  dt[failure == FALSE, .(dgp, method, noise_sd, rep, n_grid, template_mise)],
  dgp + method + noise_sd + rep ~ n_grid,
  value.var = "template_mise"
)
t_ratio_rep[, ratio_51 := `51` / `101`]
t_ratio_rep[, ratio_201 := `201` / `101`]
t_ratio_rep <- melt(
  t_ratio_rep,
  id.vars = c("dgp", "method", "noise_sd", "rep"),
  measure.vars = c("ratio_51", "ratio_201"),
  variable.name = "n_grid",
  value.name = "ratio"
)
t_ratio_rep[, n_grid := sub("^ratio_", "", n_grid)]
t_ratio_dt <- t_ratio_rep[, .(
  ratio = median(ratio, na.rm = TRUE),
  q25 = quantile(ratio, 0.25, na.rm = TRUE),
  q75 = quantile(ratio, 0.75, na.rm = TRUE)
), by = .(dgp, method, n_grid, noise_sd)]

ggplot(t_ratio_dt, aes(x = ratio, y = method, color = method)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_point(size = 2, alpha = 0.7) +
  geom_errorbarh(aes(xmin = q25, xmax = q75), height = 0.15, alpha = 0.3) +
  geom_text(aes(label = dgp), size = 3, vjust = -0.8, show.legend = FALSE) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4,
    color = "black", linewidth = 0.4) +
  facet_grid(n_grid ~ noise_sd,
    labeller = labeller(
      noise_sd = function(x) paste0("noise=", x),
      n_grid = function(x) paste0("grid=", x)
    )
  ) +
  scale_color_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_continuous(trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("1/4", "1/2", "1", "2", "4")
  ) +
  labs(x = expression("Template MISE ratio vs grid=101 (" * log[2] * " scale)"), y = NULL) +
  theme_benchmark()
```

# Elastic Distance by Grid {#sec-elastic}

SRSF elastic distance is reparameterization-invariant. A numerical convergence test
shows that finer grids yield slightly larger elastic distances for fixed function pairs
(~4% from grid=51 to 1001 for smooth functions), but this effect is far too small to
explain the patterns below. The large grid effects visible under noise are genuine:
finer grids provide more noisy observation points to the template estimator, degrading
the estimated template shape.

```{r}
#| label: elastic-grid
#| fig-height: 7
#| fig-cap: "Elastic (SRSF) template distance by grid size and method."
if ("template_elastic_dist" %in% names(dt) && any(!is.na(dt$template_elastic_dist))) {
  ggplot(
    dt[failure == FALSE & !is.na(template_elastic_dist)],
    aes(y = method, x = template_elastic_dist, fill = n_grid)
  ) +
    geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, position = "dodge") +
    facet_wrap(~noise_sd,
      labeller = labeller(noise_sd = function(x) paste0("noise=", x))) +
    scale_y_discrete(labels = mlabs) +
    scale_x_log10() +
    scale_fill_brewer(palette = "Set2", name = "Grid size") +
    labs(y = NULL, x = "Elastic distance (log scale)") +
    theme_benchmark()
} else {
  cat("*Elastic distance not available in results.*")
}
```

```{r}
#| label: elastic-ratio
#| fig-height: 8
#| fig-cap: "Elastic distance ratio relative to grid=101. Each labeled point is one DGP and horizontal bars show the IQR of paired per-rep ratios. Values > 1 mean that grid degrades template shape recovery."
if ("template_elastic_dist" %in% names(dt) && any(!is.na(dt$template_elastic_dist))) {
  e_ratio_rep <- dcast(
    dt[failure == FALSE & !is.na(template_elastic_dist),
      .(dgp, method, noise_sd, rep, n_grid, template_elastic_dist)],
    dgp + method + noise_sd + rep ~ n_grid,
    value.var = "template_elastic_dist"
  )
  e_ratio_rep[, ratio_51 := `51` / `101`]
  e_ratio_rep[, ratio_201 := `201` / `101`]
  e_ratio_rep <- melt(
    e_ratio_rep,
    id.vars = c("dgp", "method", "noise_sd", "rep"),
    measure.vars = c("ratio_51", "ratio_201"),
    variable.name = "n_grid",
    value.name = "ratio"
  )
  e_ratio_rep[, n_grid := sub("^ratio_", "", n_grid)]
  e_ratio_dt <- e_ratio_rep[, .(
    ratio = median(ratio, na.rm = TRUE),
    q25 = quantile(ratio, 0.25, na.rm = TRUE),
    q75 = quantile(ratio, 0.75, na.rm = TRUE)
  ), by = .(dgp, method, n_grid, noise_sd)]

  print(
    ggplot(e_ratio_dt, aes(x = ratio, y = method, color = method)) +
      geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
      geom_point(size = 2, alpha = 0.7) +
      geom_errorbarh(aes(xmin = q25, xmax = q75), height = 0.15, alpha = 0.3) +
      geom_text(aes(label = dgp), size = 3, vjust = -0.8, show.legend = FALSE) +
      stat_summary(fun = median, geom = "crossbar", width = 0.4,
        color = "black", linewidth = 0.4) +
      facet_grid(n_grid ~ noise_sd,
        labeller = labeller(
          noise_sd = function(x) paste0("noise=", x),
          n_grid = function(x) paste0("grid=", x)
        )
      ) +
      scale_color_manual(values = mcols, guide = "none") +
      scale_y_discrete(labels = mlabs) +
      scale_x_continuous(trans = "log2",
        breaks = c(0.25, 0.5, 1, 2, 4),
        labels = c("1/4", "1/2", "1", "2", "4")
      ) +
      labs(
        x = expression("Elastic distance ratio vs grid=101 (" * log[2] * " scale)"),
        y = NULL
      ) +
      theme_benchmark()
  )
}
```

```{r}
#| label: elastic-ratio-findings
#| results: asis
if ("template_elastic_dist" %in% names(dt) && any(!is.na(dt$template_elastic_dist))) {
  e_ratio_dt2 <- dt[failure == FALSE & !is.na(template_elastic_dist),
    .(elastic_dist = median(template_elastic_dist, na.rm = TRUE)),
    by = .(dgp, method, n_grid, noise_sd)
  ]
  e_base2 <- e_ratio_dt2[n_grid == "101",
    .(dgp, method, noise_sd, base = elastic_dist)]
  e_ratio_dt2 <- merge(e_ratio_dt2, e_base2, by = c("dgp", "method", "noise_sd"))
  e_ratio_dt2[, ratio := elastic_dist / base]
  e_ratio_dt2 <- e_ratio_dt2[n_grid != "101"]

  e_summary <- e_ratio_dt2[,
    .(median_ratio = median(ratio, na.rm = TRUE),
      min_ratio = min(ratio, na.rm = TRUE),
      max_ratio = max(ratio, na.rm = TRUE)),
    by = .(method, n_grid)
  ]

  cat("**Elastic distance ratios vs grid=101** (median [min, max] across DGPs and noise):\n\n")
  for (g in sort(unique(as.character(e_summary$n_grid)))) {
    cat(sprintf("*Grid = %s:*\n\n", g))
    sub <- e_summary[n_grid == g][order(median_ratio)]
    for (i in seq_len(nrow(sub))) {
      cat(sprintf("- **%s**: %.2f [%.2f, %.2f]\n",
        mlabs[sub$method[i]], sub$median_ratio[i],
        sub$min_ratio[i], sub$max_ratio[i]))
    }
    cat("\n")
  }

  # Check if elastic distance shows same SRVF pattern as warp MISE
  srvf_e <- e_ratio_dt2[method == "srvf",
    .(median_ratio = median(ratio, na.rm = TRUE)),
    by = .(n_grid, noise_sd)
  ]
  srvf_e201_n0 <- srvf_e[n_grid == "201" & noise_sd == "0", median_ratio]
  srvf_e201_n3 <- srvf_e[n_grid == "201" & noise_sd == "0.3", median_ratio]
  if (length(srvf_e201_n0) > 0 && length(srvf_e201_n3) > 0) {
    cat(sprintf(
      "SRVF elastic distance ratio at grid=201: %.2f (noise=0), %.2f (noise=0.3). ",
      srvf_e201_n0, srvf_e201_n3
    ))
    if (srvf_e201_n3 > srvf_e201_n0) {
      cat("Finer grids degrade template shape recovery more under noise, ",
        "consistent with the warp MISE pattern.\n")
    } else {
      cat("The pattern differs from warp MISE — elastic distance is less ",
        "sensitive to the grid × noise interaction for template estimation.\n")
    }
  }
}
```

# Alignment Error by Grid {#sec-alignment-error}

Alignment error measures the practical downstream impact:
$\text{mean}_i \int (\hat{x}_i^{\text{aligned}} - \tilde{x}_i)^2\,dt$.
Do grid artifacts in warp estimation translate into worse alignment?

```{r}
#| label: ae-grid-method
#| fig-height: 7
#| fig-cap: "Alignment error by grid size and method, faceted by noise level."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = alignment_error, fill = n_grid)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, position = "dodge") +
  facet_wrap(~noise_sd, labeller = labeller(noise_sd = function(x) paste0("noise=", x))) +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  scale_fill_brewer(palette = "Set2", name = "Grid size") +
  labs(y = NULL, x = "Alignment error (log scale)") +
  theme_benchmark()
```

```{r}
#| label: ae-grid-ratios
#| fig-height: 8
#| fig-cap: "Alignment error ratio relative to grid=101, per method and noise level. Each labeled point is one DGP; horizontal bars show the IQR of paired per-rep ratios and the crossbar shows the median across DGPs."
ae_ratio_rep <- dcast(
  dt[failure == FALSE, .(dgp, method, noise_sd, rep, n_grid, alignment_error)],
  dgp + method + noise_sd + rep ~ n_grid,
  value.var = "alignment_error"
)
ae_ratio_rep[, ratio_51 := `51` / `101`]
ae_ratio_rep[, ratio_201 := `201` / `101`]
ae_ratio_rep <- melt(
  ae_ratio_rep,
  id.vars = c("dgp", "method", "noise_sd", "rep"),
  measure.vars = c("ratio_51", "ratio_201"),
  variable.name = "n_grid",
  value.name = "ratio"
)
ae_ratio_rep[, n_grid := sub("^ratio_", "", n_grid)]
ae_ratio_dt <- ae_ratio_rep[, .(
  ratio = median(ratio, na.rm = TRUE),
  q25 = quantile(ratio, 0.25, na.rm = TRUE),
  q75 = quantile(ratio, 0.75, na.rm = TRUE)
), by = .(dgp, method, n_grid, noise_sd)]

ggplot(ae_ratio_dt, aes(x = ratio, y = method, color = method)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_point(size = 2, alpha = 0.7) +
  geom_errorbarh(aes(xmin = q25, xmax = q75), height = 0.15, alpha = 0.3) +
  geom_text(aes(label = dgp), size = 3, vjust = -0.8, show.legend = FALSE) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4,
    color = "black", linewidth = 0.4) +
  facet_grid(n_grid ~ noise_sd,
    labeller = labeller(
      noise_sd = function(x) paste0("noise=", x),
      n_grid = function(x) paste0("grid=", x)
    )
  ) +
  scale_color_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_continuous(trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("1/4", "1/2", "1", "2", "4")
  ) +
  labs(x = expression("Alignment error ratio vs grid=101 (" * log[2] * " scale)"),
    y = NULL) +
  theme_benchmark()
```

```{r}
#| label: ae-grid-findings
#| results: asis
ae_ratio_summary <- ae_ratio_dt[,
  .(median_ratio = median(ratio, na.rm = TRUE),
    min_ratio = min(ratio, na.rm = TRUE),
    max_ratio = max(ratio, na.rm = TRUE)),
  by = .(method, n_grid)
]

cat("**Alignment error ratios vs grid=101** (median [min, max] across DGPs and noise):\n\n")
for (g in sort(unique(as.character(ae_ratio_summary$n_grid)))) {
  cat(sprintf("*Grid = %s:*\n\n", g))
  sub <- ae_ratio_summary[n_grid == g][order(median_ratio)]
  for (i in seq_len(nrow(sub))) {
    cat(sprintf("- **%s**: %.2f [%.2f, %.2f]\n",
      mlabs[sub$method[i]], sub$median_ratio[i],
      sub$min_ratio[i], sub$max_ratio[i]))
  }
  cat("\n")
}

# Compare with warp MISE ratios
warp_ratio_summary <- ratio_dt[,
  .(warp_median = median(ratio, na.rm = TRUE)),
  by = .(method, n_grid)
]
ae_ratio_summary2 <- ae_ratio_dt[,
  .(ae_median = median(ratio, na.rm = TRUE)),
  by = .(method, n_grid)
]
both <- merge(warp_ratio_summary, ae_ratio_summary2, by = c("method", "n_grid"))
r_grid <- cor(both$warp_median, both$ae_median, use = "complete.obs")
cat(sprintf(
  "Correlation between warp MISE and alignment error grid ratios: r = %.2f. ",
  r_grid))
if (r_grid > 0.9) {
  cat("Grid effects on warp recovery translate almost directly into alignment quality.\n\n")
} else {
  cat("Grid effects on alignment error partially decouple from warp MISE — ",
    "alignment quality has some robustness to grid-induced warp artifacts.\n\n")
}

# SRVF-specific: does the noise-dependent reversal hold for alignment error too?
srvf_ae <- ae_ratio_dt[method == "srvf",
  .(median_ratio = median(ratio, na.rm = TRUE)),
  by = .(n_grid, noise_sd)]
srvf_ae51_n0 <- srvf_ae[n_grid == "51" & noise_sd == "0", median_ratio]
srvf_ae51_n3 <- srvf_ae[n_grid == "51" & noise_sd == "0.3", median_ratio]
srvf_ae201_n0 <- srvf_ae[n_grid == "201" & noise_sd == "0", median_ratio]
srvf_ae201_n3 <- srvf_ae[n_grid == "201" & noise_sd == "0.3", median_ratio]

if (length(srvf_ae51_n0) > 0 && length(srvf_ae201_n3) > 0) {
  cat(sprintf(
    "SRVF alignment error ratios: grid=51 is %.2f (noise=0), %.2f (noise=0.3); grid=201 is %.2f (noise=0), %.2f (noise=0.3). ",
    srvf_ae51_n0, srvf_ae51_n3, srvf_ae201_n0, srvf_ae201_n3))
  reversal <- (srvf_ae51_n0 > 1 && srvf_ae51_n3 < 1) ||
    (srvf_ae201_n0 < 1 && srvf_ae201_n3 > 1)
  if (reversal) {
    cat("The noise-dependent reversal observed for warp MISE is also present for alignment error.\n")
  } else {
    cat("The noise-dependent reversal pattern differs somewhat for alignment error.\n")
  }
}
```

# Grid $\times$ Noise Interaction {#sec-interaction}

The key question: does the combination of fine grid + high noise compound for SRVF?

```{r}
#| label: grid-noise-interaction
#| fig-height: 7
#| fig-cap: "Grid x noise interaction: warp MISE profiles per method. Points show medians and vertical bars show IQRs; each line connects the 3 grid sizes for one DGP. Steeper upward slopes at grid=201 for noise=0.3 indicate grid-noise compounding. Note: y-axis scales differ across methods due to large baseline differences."
int_dt <- dt[failure == FALSE,
  .(
    warp_mise = median(warp_mise, na.rm = TRUE),
    q25 = quantile(warp_mise, 0.25, na.rm = TRUE),
    q75 = quantile(warp_mise, 0.75, na.rm = TRUE)
  ),
  by = .(dgp, method, n_grid, noise_sd)
]

ggplot(int_dt, aes(x = n_grid, y = warp_mise, group = dgp, color = dgp)) +
  geom_line(alpha = 0.7) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.08, alpha = 0.25) +
  facet_grid(method ~ noise_sd,
    scales = "free_y",
    labeller = labeller(
      noise_sd = function(x) paste0("noise=", x),
      method = mlabs
    )
  ) +
  scale_y_log10() +
  labs(x = "Grid size", y = "Warp MISE (log scale)", color = "DGP") +
  theme_benchmark()
```

```{r}
#| label: grid-noise-findings
#| results: asis
# Decompose compound effect into main effects + interaction
compound_dt <- dt[failure == FALSE,
  .(warp_mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method, n_grid, noise_sd)
]

# For SRVF: decompose the compound ratio
# Compound = (grid=201, noise=0.3) / (grid=101, noise=0)
# Noise main effect = (grid=101, noise=0.3) / (grid=101, noise=0)
# Grid main effect = (grid=201, noise=0) / (grid=101, noise=0)
# Interaction = compound / (noise_main * grid_main)

ref_clean <- compound_dt[n_grid == "101" & noise_sd == "0",
  .(dgp, method, ref = warp_mise)]
noise_main <- compound_dt[n_grid == "101" & noise_sd == "0.3",
  .(dgp, method, noisy = warp_mise)]
grid_main <- compound_dt[n_grid == "201" & noise_sd == "0",
  .(dgp, method, fine = warp_mise)]
stress <- compound_dt[n_grid == "201" & noise_sd == "0.3",
  .(dgp, method, stressed = warp_mise)]

decomp <- Reduce(function(a, b) merge(a, b, by = c("dgp", "method")),
  list(ref_clean, noise_main, grid_main, stress))
decomp[, `:=`(
  compound = stressed / ref,
  noise_effect = noisy / ref,
  grid_effect = fine / ref,
  interaction = (stressed / ref) / ((noisy / ref) * (fine / ref))
)]

cat("**Compound effect decomposition** ",
  "(grid=201/noise=0.3 vs grid=101/noise=0):\n\n")

decomp_summary <- decomp[,
  .(
    compound = median(compound, na.rm = TRUE),
    noise_main = median(noise_effect, na.rm = TRUE),
    grid_main = median(grid_effect, na.rm = TRUE),
    interaction = median(interaction, na.rm = TRUE)
  ),
  by = method
][order(-compound)]

tbl2 <- c(
  "| Method | Compound | Noise main | Grid main | Interaction |",
  "|--------|----------|------------|-----------|-------------|"
)
for (i in seq_len(nrow(decomp_summary))) {
  tbl2 <- c(tbl2, sprintf("| %s | %.1f\u00d7 | %.1f\u00d7 | %.2f\u00d7 | %.2f\u00d7 |",
    mlabs[decomp_summary$method[i]],
    decomp_summary$compound[i],
    decomp_summary$noise_main[i],
    decomp_summary$grid_main[i],
    decomp_summary$interaction[i]))
}
cat(paste(tbl2, collapse = "\n"), "\n\n")

srvf_decomp <- decomp_summary[method == "srvf"]
cat(sprintf(
  "For SRVF, the compound ratio of %.1f× decomposes into a noise main effect ",
  srvf_decomp$compound
))
cat(sprintf(
  "of %.1f×, a grid main effect of %.2f×, and a super-multiplicative interaction ",
  srvf_decomp$noise_main, srvf_decomp$grid_main
))
cat(sprintf(
  "of %.2f×. ", srvf_decomp$interaction
))
if (srvf_decomp$interaction > 1.1) {
  cat("The interaction exceeds 1, confirming that grid and noise effects ",
    "are super-additive for SRVF — fine grids amplify the noise problem ",
    "beyond what either factor alone would predict.\n\n")
} else if (srvf_decomp$interaction < 0.9) {
  cat("The interaction is below 1, suggesting the compound effect is ",
    "sub-multiplicative — the large compound ratio is mostly driven by ",
    "the noise main effect.\n\n")
} else {
  cat("The interaction is near 1, suggesting the compound effect is ",
    "approximately multiplicative — grid and noise contribute independently ",
    "with no synergy.\n\n")
}

# Note on extreme max values
max_compound <- decomp[, .(max_compound = max(compound)), by = method]
extreme <- max_compound[max_compound > 10]
if (nrow(extreme) > 0) {
  cat("*Note:* Extreme compound ratios (")
  cat(paste(sprintf("%s: %.0f×", mlabs[extreme$method], extreme$max_compound),
    collapse = ", "))
  cat(") are driven by DGPs where the baseline (grid=101, noise=0) warp MISE ",
    "is very small, making the ratio sensitive to small absolute differences.\n\n")
}
```

# Per-DGP Profiles {#sec-dgp}

```{r}
#| label: dgp-profiles
#| fig-height: 10
#| fig-cap: "Per-DGP warp MISE by grid size, faceted by method. Lines connect grid levels within each noise condition. IQR error bars show variability across replications."
profile_dt <- dt[failure == FALSE,
  .(
    median = median(warp_mise, na.rm = TRUE),
    q25 = quantile(warp_mise, 0.25, na.rm = TRUE),
    q75 = quantile(warp_mise, 0.75, na.rm = TRUE)
  ),
  by = .(dgp, method, n_grid, noise_sd)
]

ggplot(profile_dt, aes(x = n_grid, y = median, color = noise_sd, group = noise_sd)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.1, alpha = 0.4) +
  facet_grid(dgp ~ method, scales = "free_y",
    labeller = labeller(method = mlabs, dgp = dlabs)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Dark2", name = "Noise") +
  labs(x = "Grid size", y = "Warp MISE (log scale)") +
  theme_benchmark(base_size = 9)
```

# Timing {#sec-timing}

```{r}
#| label: timing
#| fig-height: 4
#| fig-width: 12
#| fig-cap: "Computation time per curve (seconds) by grid size and method. Common y-axis (log scale) enables direct comparison. SRVF scales steeply due to O(n²) DP alignment."
ggplot(
  dt[failure == FALSE],
  aes(x = n_grid, y = time_per_curve, fill = n_grid)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_wrap(~method, nrow = 1, labeller = labeller(method = mlabs)) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2", name = "Grid size") +
  labs(x = "Grid size", y = "Time per curve (sec, log scale)") +
  theme_benchmark()
```

```{r}
#| label: timing-table
#| results: asis
timing_dt <- dt[failure == FALSE,
  .(time_per_curve = median(time_per_curve, na.rm = TRUE)),
  by = .(method, n_grid)
]
timing_wide <- dcast(timing_dt, method ~ n_grid, value.var = "time_per_curve")

# Compute timing ratios
timing_wide[, ratio_201_101 := `201` / `101`]
timing_wide[, method := mlabs[method]]
cat("**Median time per curve (seconds) by method and grid size:**\n\n")
print(knitr::kable(timing_wide, digits = 4))
cat("\n")

srvf_ratio <- timing_wide[method == "SRVF", ratio_201_101]
cat(sprintf(
  "SRVF timing scales %.1f× from grid=101 to grid=201 (DP alignment is O(n²)). ",
  srvf_ratio
))
cat("FDA methods show negligible timing change across grid sizes.\n")
```

# Summary {#sec-summary}

```{r}
#| label: summary-table
#| results: asis
# Summary: for each method, median warp MISE ratio at grid=51 and grid=201 vs 101
ratio_summary_all <- ratio_dt[,
  .(median_ratio = median(ratio, na.rm = TRUE)),
  by = .(method, n_grid, noise_sd)
]
summary_wide <- dcast(ratio_summary_all,
  method + noise_sd ~ n_grid,
  value.var = "median_ratio"
)
summary_wide[, method := mlabs[method]]
cat("**Median warp MISE ratio vs grid=101** (by method and noise level):\n\n")
print(knitr::kable(summary_wide, digits = 2))
cat("\n")
```

```{r}
#| label: key-findings
#| results: asis
cat("**Key findings:**\n\n")

# 1. SRVF reversal
srvf_51_n0 <- ratio_dt[method == "srvf" & n_grid == "51" & noise_sd == "0",
  median(ratio, na.rm = TRUE)]
srvf_51_n3 <- ratio_dt[method == "srvf" & n_grid == "51" & noise_sd == "0.3",
  median(ratio, na.rm = TRUE)]
srvf_201_n0 <- ratio_dt[method == "srvf" & n_grid == "201" & noise_sd == "0",
  median(ratio, na.rm = TRUE)]
srvf_201_n3 <- ratio_dt[method == "srvf" & n_grid == "201" & noise_sd == "0.3",
  median(ratio, na.rm = TRUE)]

cat(sprintf(
  "1. **SRVF grid sensitivity reverses with noise.** At noise=0, grid=51 ",
  "degrades SRVF (%.2f×) while grid=201 helps (%.2f×). At noise=0.3, ",
  "grid=51 helps (%.2f×) while grid=201 degrades (%.2f×). Coarse grids ",
  "act as implicit regularization by smoothing numerical derivatives.\n\n",
  srvf_51_n0, srvf_201_n0, srvf_51_n3, srvf_201_n3
))

# 2. FDA stability
fda_def_range <- ratio_dt[method == "fda_default",
  .(lo = min(ratio, na.rm = TRUE), hi = max(ratio, na.rm = TRUE))]
cat(sprintf(
  "2. **FDA methods are grid-insensitive.** fda_default ratios range from ",
  "%.2f to %.2f across all grids, noise levels, and DGPs. fda_crit1 shows ",
  "slight improvement at finer grids under noise.\n\n",
  fda_def_range$lo, fda_def_range$hi
))

# 3. Timing
cat("3. **SRVF timing scales steeply** (~4× per grid doubling due to O(n²) DP ",
  "alignment). FDA timing is essentially constant across grids.\n\n")

# 4. Practical
cat("4. **Grid=101 is a robust default** for `tf_register()`. It avoids SRVF ",
  "degradation at finer grids while providing sufficient resolution for FDA ",
  "methods. Users who need finer grids with SRVF should investigate whether ",
  "penalization (Study B) can mitigate the derivative noise amplification — ",
  "this hypothesis is not tested in Study C.\n\n")

cat("5. **Grid resolution is a confound for SRVF benchmarking.** ",
  "Method comparisons involving SRVF at grid=201 (or finer) ",
  "will underestimate SRVF performance relative to FDA methods, especially ",
  "under noise. Studies A and B use grid=101, which is neutral territory.\n")
```
