---
title: "Study F: Pre-Smoothing"
subtitle: "Registration Benchmark v2"
execute:
  echo: false
format:
  html:
    code-fold: false
---

```{r}
#| label: setup
library(data.table)
library(ggplot2)
library(patchwork)

source("sim-analyze.R")

EPS_MISE <- 1e-12
RESCUE_BOUND <- log(0.9)
NONINF_BOUND <- log(1.05)

f_dt <- tryCatch(load_study_f(), error = function(e) e)
has_results <- !inherits(f_dt, "error")

primary_variants <- c("lowess_f015", "spline_local_k25")
all_variants <- c(
  "none", "lowess_f010", "lowess_f015",
  "spline_local_k15", "spline_local_k25", "spline_global_k25"
)
variant_labels <- c(
  none = "Raw SRVF",
  lowess_f010 = "LOWESS (f=0.10)",
  lowess_f015 = "LOWESS (f=0.15)",
  spline_local_k15 = "Spline local (k=15)",
  spline_local_k25 = "Spline local (k=25)",
  spline_global_k25 = "Spline global (k=25)"
)
variant_colors <- c(
  none = "#999999",
  lowess_f010 = "#E41A1C",
  lowess_f015 = "#984EA3",
  spline_local_k15 = "#FF7F00",
  spline_local_k25 = "#377EB8",
  spline_global_k25 = "#4DAF4A"
)
weak_dgps <- c("D01", "D02", "D05", "D07", "D09", "D10")
control_dgps <- c("D03", "D12", "D14")
```

```{r}
#| label: bail-if-missing
if (!has_results) {
  msgs <- c(if (inherits(f_dt, "error")) conditionMessage(f_dt))
  cat("Study F results are not available yet.\n\n")
  cat(paste0("- ", msgs, collapse = "\n"))
  knitr::knit_exit()
}
```

# Overview

```{r}
#| label: overview
#| results: asis
f_ok <- f_dt[failure == FALSE]
n_rows <- nrow(f_dt)
n_fail <- sum(f_dt$failure)
n_dgps <- uniqueN(f_dt$dgp)
n_variants <- uniqueN(f_dt$preproc_id)

cat(sprintf(
  "**Study F**: %s runs | %d DGPs | %d pre-processing variants | %.1f%% failures\n\n",
  format(n_rows, big.mark = ","), n_dgps, n_variants, 100 * n_fail / n_rows
))
cat(sprintf(
  "- **Weak DGPs** (SRVF degrades under noise in Study A): %s\n",
  paste(weak_dgps, collapse = ", ")
))
cat(sprintf(
  "- **Control DGPs** (wiggly templates, harder to smooth): %s\n",
  paste(control_dgps, collapse = ", ")
))
cat("- **Method**: SRVF only (all variants)\n")
cat("- **Design**: 9 DGPs x 6 variants x 3 noise x 2 severity x 100 reps\n")
cat("- **Evaluation**: Warps estimated from pre-smoothed curves, then transferred back to raw curves for comparable metrics\n")
```

# Primary Estimand: Paired Log-Ratio

The primary estimand is the paired log-ratio of warp MISE (pre-smoothed vs raw SRVF)
on the same `(dgp, rep)` seed. Negative values mean pre-smoothing helps.

```{r}
#| label: primary-estimand
none_dt <- f_ok[
  preproc_id == "none",
  .(dgp, severity, noise_sd, rep, warp_mise_none = warp_mise)
]

# Verify pairing: exactly one baseline per (dgp, severity, noise_sd, rep)
stopifnot(
  "Non-unique baselines" = !anyDuplicated(none_dt, by = c("dgp", "severity", "noise_sd", "rep"))
)

ratio_dt <- merge(
  f_ok[preproc_id != "none"],
  none_dt,
  by = c("dgp", "severity", "noise_sd", "rep")
)

expected_paired <- nrow(f_ok[preproc_id != "none"])
stopifnot(
  "Pairing lost rows" = nrow(ratio_dt) == expected_paired
)

ratio_dt[,
  log_ratio := log(pmax(warp_mise, EPS_MISE) / pmax(warp_mise_none, EPS_MISE))
]

summary_dt <- ratio_dt[,
  .(
    mean_log_ratio = mean(log_ratio, na.rm = TRUE),
    mcse = sd(log_ratio, na.rm = TRUE) / sqrt(.N),
    upper_2mcse = mean(log_ratio, na.rm = TRUE) +
      2 * sd(log_ratio, na.rm = TRUE) / sqrt(.N),
    pct_improved = 100 * mean(log_ratio < 0),
    n = .N
  ),
  by = .(dgp, severity, noise_sd, preproc_id, preproc_primary)
]

summary_dt[, noise_num := as.numeric(as.character(noise_sd))]
summary_dt[, cell_type := fifelse(
  dgp %in% weak_dgps & noise_num > 0,
  "weak_noisy",
  fifelse(dgp %in% control_dgps | noise_num == 0, "control", "other")
)]

summary_dt[,
  decision := fifelse(
    cell_type == "weak_noisy" & upper_2mcse < RESCUE_BOUND,
    "rescue",
    fifelse(
      cell_type == "control" & upper_2mcse < NONINF_BOUND,
      "noninferior",
      "undecided"
    )
  )
]
```

## All Variants: Warp MISE Ratio Distribution

```{r}
#| label: all-variants-ratio
#| fig-width: 11
#| fig-height: 6
ratio_dt[, ratio := exp(log_ratio)]
ratio_dt[, preproc_id_plot := factor(preproc_id, levels = setdiff(all_variants, "none"))]

ratio_breaks <- c(0.02, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 50)

ggplot(
  ratio_dt,
  aes(x = noise_sd, y = ratio, fill = preproc_id_plot)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_manual(values = variant_colors, labels = variant_labels) +
  scale_y_log10(
    breaks = ratio_breaks,
    labels = ratio_breaks
  ) +
  labs(
    x = "Noise level",
    y = "Warp MISE ratio (vs raw SRVF)",
    fill = "Pre-processing",
    subtitle = "< 1 = pre-smoothing helps, > 1 = pre-smoothing hurts"
  ) +
  theme_benchmark()
```

```{r}
#| label: all-variants-table
ratio_overall <- ratio_dt[,
  .(
    median_ratio = exp(median(log_ratio)),
    pct_improved = 100 * mean(log_ratio < 0)
  ),
  by = .(preproc_id, noise_sd)
]
knitr::kable(
  dcast(ratio_overall, preproc_id ~ noise_sd,
    value.var = c("median_ratio", "pct_improved")
  ),
  digits = 2,
  caption = "Median warp MISE ratio (variant / raw SRVF) and % of runs improved, by noise level."
)
```

## Primary Variants: Cell-Level Ratio Distribution

```{r}
#| label: primary-logratio-plot
#| fig-width: 12
#| fig-height: 12
primary_raw <- ratio_dt[preproc_id %in% primary_variants]
primary_raw[, ratio := exp(log_ratio)]

ratio_breaks <- c(0.01, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 50, 200)

ggplot(
  primary_raw,
  aes(x = noise_sd, y = ratio, fill = preproc_id)
) +
  geom_hline(yintercept = 1, color = "grey40") +
  geom_hline(
    yintercept = exp(RESCUE_BOUND),
    linetype = "dashed",
    color = "darkgreen",
    linewidth = 0.3
  ) +
  geom_hline(
    yintercept = exp(NONINF_BOUND),
    linetype = "dotted",
    color = "darkred",
    linewidth = 0.3
  ) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_grid(dgp ~ severity, labeller = labeller(
    severity = function(x) paste0("severity = ", x)
  )) +
  scale_fill_manual(
    values = variant_colors[primary_variants],
    labels = variant_labels[primary_variants]
  ) +
  scale_y_log10(
    breaks = ratio_breaks,
    labels = ratio_breaks
  ) +
  labs(
    x = "Noise level",
    y = "Warp MISE ratio (vs raw SRVF)",
    fill = "Variant",
    subtitle = "Dashed green = 0.9 (rescue); dotted red = 1.05 (non-inferiority). < 1 = pre-smoothing helps."
  ) +
  theme_benchmark(base_size = 10)
```

# Study-Level Decision Rule

```{r}
#| label: study-rule
#| results: asis
rule_dt <- summary_dt[preproc_id %in% primary_variants]

study_rule <- rule_dt[,
  .(
    rescue_pass = sum(cell_type == "weak_noisy" & decision == "rescue"),
    rescue_total = sum(cell_type == "weak_noisy"),
    control_pass = sum(cell_type == "control" & decision == "noninferior"),
    control_total = sum(cell_type == "control")
  ),
  by = preproc_id
]
study_rule[,
  plausible_default := rescue_pass >= 18 & (control_total - control_pass) == 0
]

knitr::kable(
  study_rule[order(preproc_id)],
  caption = "Study-level decision rule for the two primary variants."
)

cat("\n\n")
for (pid in primary_variants) {
  sr <- study_rule[preproc_id == pid]
  cat(sprintf(
    "**%s**: %d/%d rescue cells pass, %d/%d control cells pass non-inferiority. **%s as unconditional default.**\n\n",
    variant_labels[pid],
    sr$rescue_pass, sr$rescue_total,
    sr$control_pass, sr$control_total,
    ifelse(sr$plausible_default, "Plausible", "Not plausible")
  ))
}
```

# Cross-Over Analysis

```{r}
#| label: crossover
#| fig-width: 9
#| fig-height: 5
xover <- ratio_dt[
  preproc_id == "spline_local_k25",
  .(
    mean_lr = mean(log_ratio),
    pct_better = 100 * mean(log_ratio < 0),
    mcse = sd(log_ratio) / sqrt(.N)
  ),
  by = .(dgp_group = fifelse(dgp %in% weak_dgps, "Weak DGPs", "Control DGPs"),
    noise_sd, severity)
]

xover[, mean_ratio := exp(mean_lr)]
xover[, ratio_lo := exp(mean_lr - 2 * mcse)]
xover[, ratio_hi := exp(mean_lr + 2 * mcse)]

ratio_breaks_xover <- c(0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 20)

ggplot(xover, aes(x = noise_sd, y = mean_ratio, color = dgp_group, group = dgp_group)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = ratio_lo, ymax = ratio_hi),
    width = 0.06
  ) +
  facet_wrap(~severity, labeller = labeller(
    severity = function(x) paste0("severity = ", x)
  )) +
  scale_color_manual(values = c("Weak DGPs" = "#E41A1C", "Control DGPs" = "#377EB8")) +
  scale_y_log10(
    breaks = ratio_breaks_xover,
    labels = ratio_breaks_xover
  ) +
  labs(
    x = "Noise level",
    y = "Warp MISE ratio (spline_local_k25 vs raw SRVF)",
    color = NULL,
    subtitle = "Cross-over: pre-smoothing helps below 1"
  ) +
  theme_benchmark()
```

```{r}
#| label: crossover-text
#| results: asis
noise_levels <- sort(as.numeric(as.character(unique(xover$noise_sd))))
xover_wide <- dcast(xover, dgp_group + severity ~ noise_sd,
  value.var = "mean_lr"
)

for (i in seq_len(nrow(xover_wide))) {
  r <- xover_wide[i]
  vals <- as.numeric(r[, as.character(noise_levels), with = FALSE])
  # Find first transition from positive to negative (or report no cross-over)
  transitions <- which(vals[-length(vals)] > 0 & vals[-1] < 0)
  if (length(transitions) > 0) {
    t1 <- transitions[1]
    cat(sprintf(
      "- **%s, severity=%s**: cross-over between noise %s and %s\n",
      r$dgp_group, r$severity,
      noise_levels[t1], noise_levels[t1 + 1]
    ))
  } else if (all(vals < 0)) {
    cat(sprintf(
      "- **%s, severity=%s**: pre-smoothing helps at all noise levels\n",
      r$dgp_group, r$severity
    ))
  } else {
    cat(sprintf(
      "- **%s, severity=%s**: pre-smoothing hurts at all noise levels\n",
      r$dgp_group, r$severity
    ))
  }
}
```

# Secondary Metrics

## Template Quality

```{r}
#| label: template-metrics
#| fig-width: 10
#| fig-height: 5
tpl_dt <- f_ok[,
  .(
    tmise = median(template_mise, na.rm = TRUE),
    edist = median(template_elastic_dist, na.rm = TRUE)
  ),
  by = .(preproc_id, noise_sd)
]
tpl_dt[, preproc_id := factor(preproc_id, levels = all_variants)]

p_tmise <- ggplot(tpl_dt, aes(x = noise_sd, y = tmise, color = preproc_id, group = preproc_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  scale_color_manual(values = variant_colors, labels = variant_labels) +
  scale_y_log10() +
  labs(x = "Noise", y = "Median template MISE (log)", color = NULL) +
  theme_benchmark(base_size = 10)

p_edist <- ggplot(tpl_dt, aes(x = noise_sd, y = edist, color = preproc_id, group = preproc_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  scale_color_manual(values = variant_colors, labels = variant_labels) +
  labs(x = "Noise", y = "Median elastic distance", color = NULL) +
  theme_benchmark(base_size = 10)

p_tmise + p_edist + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: template-table
knitr::kable(
  dcast(tpl_dt, preproc_id ~ noise_sd, value.var = "tmise"),
  digits = 4,
  caption = "Median template MISE by variant and noise level."
)
knitr::kable(
  dcast(tpl_dt, preproc_id ~ noise_sd, value.var = "edist"),
  digits = 3,
  caption = "Median template elastic distance by variant and noise level."
)
```

## Alignment Error

```{r}
#| label: alignment-error
#| fig-width: 8
#| fig-height: 4
ae_dt <- f_ok[,
  .(align_err = median(alignment_error, na.rm = TRUE)),
  by = .(preproc_id, noise_sd)
]
ae_dt[, preproc_id := factor(preproc_id, levels = all_variants)]

ggplot(ae_dt, aes(x = noise_sd, y = align_err, color = preproc_id, group = preproc_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2.5) +
  scale_color_manual(values = variant_colors, labels = variant_labels) +
  scale_y_log10() +
  labs(
    x = "Noise level",
    y = "Median alignment error (log scale)",
    color = NULL
  ) +
  theme_benchmark()
```

```{r}
#| label: alignment-error-table
knitr::kable(
  dcast(ae_dt, preproc_id ~ noise_sd, value.var = "align_err"),
  digits = 4,
  caption = "Median alignment error by variant and noise level."
)
```

# Per-DGP Breakdown

## Warp MISE by DGP and Variant

```{r}
#| label: per-dgp-heatmap
#| fig-width: 10
#| fig-height: 8
mise_by_dgp <- f_ok[,
  .(warp_mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, preproc_id, noise_sd)
]

ggplot(
  mise_by_dgp,
  aes(x = preproc_id, y = dgp, fill = log10(warp_mise))
) +
  geom_tile(color = "white", linewidth = 0.3) +
  geom_text(
    aes(label = fmt(warp_mise)),
    size = 2.2
  ) +
  facet_wrap(~noise_sd, labeller = labeller(
    noise_sd = function(x) paste0("noise = ", x)
  )) +
  scale_x_discrete(labels = function(x) sub("spline_", "spl_", x)) +
  scale_fill_viridis_c(option = "plasma", direction = -1, name = "log10(MISE)") +
  labs(x = NULL, y = NULL, subtitle = "Median warp MISE by DGP, variant, and noise level") +
  theme_benchmark(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Per-DGP Improvement at Highest Noise

```{r}
#| label: per-dgp-ratio-high-noise
#| fig-width: 10
#| fig-height: 5
hard_noise <- ratio_dt[
  as.numeric(as.character(noise_sd)) == 0.3 &
    as.numeric(as.character(severity)) == 1 &
    preproc_id %in% c("lowess_f015", "spline_local_k25", "spline_global_k25")
]

ratio_by_dgp <- hard_noise[,
  .(
    median_ratio = exp(median(log_ratio)),
    pct_improved = 100 * mean(log_ratio < 0)
  ),
  by = .(dgp, preproc_id)
]
ratio_by_dgp[, dgp_type := fifelse(dgp %in% weak_dgps, "Weak", "Control")]

ggplot(
  ratio_by_dgp,
  aes(x = dgp, y = median_ratio, fill = preproc_id)
) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  facet_grid(~dgp_type, scales = "free_x", space = "free_x") +
  scale_fill_manual(
    values = variant_colors[c("lowess_f015", "spline_local_k25", "spline_global_k25")],
    labels = variant_labels[c("lowess_f015", "spline_local_k25", "spline_global_k25")]
  ) +
  scale_y_log10(
    breaks = c(0.05, 0.1, 0.2, 0.5, 1, 2),
    labels = c("0.05", "0.1", "0.2", "0.5", "1", "2")
  ) +
  labs(
    x = NULL,
    y = "Warp MISE ratio (vs raw SRVF)",
    fill = NULL,
    subtitle = "Noise = 0.3, severity = 1.0 | < 1 = pre-smoothing helps"
  ) +
  theme_benchmark()
```

```{r}
#| label: per-dgp-ratio-table
knitr::kable(
  dcast(
    ratio_by_dgp[, .(dgp, preproc_id, dgp_type, median_ratio, pct_improved)],
    dgp + dgp_type ~ preproc_id,
    value.var = c("median_ratio")
  )[order(dgp)],
  digits = 3,
  caption = "Median warp MISE ratio by DGP (noise=0.3, severity=1.0). Values < 1 indicate improvement."
)
```

# Variant Comparison

```{r}
#| label: variant-rank
#| fig-width: 9
#| fig-height: 5
# Rank variants within each cell (lower warp MISE = rank 1)
rank_dt <- f_ok[,
  .(warp_mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, severity, noise_sd, preproc_id)
]
rank_dt[, rank := rank(warp_mise, ties.method = "min"), by = .(dgp, severity, noise_sd)]

rank_summary <- rank_dt[,
  .(mean_rank = mean(rank), n_rank1 = sum(rank == 1), n_top2 = sum(rank <= 2)),
  by = .(preproc_id, noise_sd)
]
rank_summary[, preproc_id := factor(preproc_id, levels = all_variants)]

ggplot(rank_summary, aes(x = noise_sd, y = mean_rank, color = preproc_id, group = preproc_id)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 3) +
  scale_color_manual(values = variant_colors, labels = variant_labels) +
  scale_y_reverse(breaks = 1:6) +
  labs(
    x = "Noise level",
    y = "Mean rank (lower = better)",
    color = NULL,
    subtitle = "Rank 1 = lowest median warp MISE in that cell"
  ) +
  theme_benchmark()
```

```{r}
#| label: variant-rank-table
knitr::kable(
  dcast(rank_summary, preproc_id ~ noise_sd, value.var = "mean_rank"),
  digits = 2,
  caption = "Mean rank across cells by variant and noise level."
)
```

# Ratio Distributions by DGP Group

```{r}
#| label: ratio-distributions
#| fig-width: 11
#| fig-height: 7
box_dt <- ratio_dt[
  preproc_id %in% c("lowess_f015", "spline_local_k25", "spline_global_k25")
]
box_dt[, dgp_group := fifelse(dgp %in% weak_dgps, "Weak DGPs", "Control DGPs")]

ratio_breaks_dist <- c(0.01, 0.05, 0.1, 0.2, 0.5, 1, 2, 5, 10, 50, 200)

ggplot(box_dt, aes(x = noise_sd, y = ratio, fill = preproc_id)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  geom_boxplot(outlier.size = 0.5, outlier.alpha = 0.3) +
  facet_grid(dgp_group ~ severity, labeller = labeller(
    severity = function(x) paste0("severity = ", x)
  )) +
  scale_fill_manual(
    values = variant_colors[c("lowess_f015", "spline_local_k25", "spline_global_k25")],
    labels = variant_labels[c("lowess_f015", "spline_local_k25", "spline_global_k25")]
  ) +
  scale_y_log10(
    breaks = ratio_breaks_dist,
    labels = ratio_breaks_dist
  ) +
  labs(
    x = "Noise level",
    y = "Warp MISE ratio (vs raw SRVF)",
    fill = NULL,
    subtitle = "Full distribution across reps. Below 1 = pre-smoothing helps."
  ) +
  theme_benchmark()
```

# Comparison with Study A

```{r}
#| label: study-a-comparison
#| fig-width: 10
#| fig-height: 6
a_dt <- tryCatch(load_study_a(), error = function(e) NULL)
has_study_a <- !is.null(a_dt)
```

```{r}
#| label: study-a-comparison-helper
#| eval: !expr has_study_a
study_f_dgps <- as.character(sort(unique(f_ok$dgp)))

cmp_colors <- c(
  "SRVF (raw)" = "grey75",
  "SRVF + spline(k=25)" = "grey30",
  "cc_default" = "#377EB8",
  "cc_crit1" = "#4DAF4A",
  "affine_ss" = "#984EA3",
  "landmark_auto" = "#FF7F00"
)
cmp_labels <- c(
  "SRVF (raw)" = "SRVF (raw)",
  "SRVF + spline(k=25)" = "SRVF + spline(k=25)",
  method_labels()
)
cmp_method_levels <- c(
  "SRVF (raw)", "SRVF + spline(k=25)",
  "cc_default", "cc_crit1", "affine_ss", "landmark_auto"
)

build_cmp_dt <- function(metric, f_data = f_ok, a_data = a_dt) {
  preproc <- f_data[
    preproc_id == "spline_local_k25",
    .(
      med = median(get(metric), na.rm = TRUE),
      q25 = quantile(get(metric), 0.25, na.rm = TRUE),
      q75 = quantile(get(metric), 0.75, na.rm = TRUE)
    ),
    by = .(dgp, noise_sd, severity)
  ]
  preproc[, method := "SRVF + spline(k=25)"]

  raw <- f_data[
    preproc_id == "none",
    .(
      med = median(get(metric), na.rm = TRUE),
      q25 = quantile(get(metric), 0.25, na.rm = TRUE),
      q75 = quantile(get(metric), 0.75, na.rm = TRUE)
    ),
    by = .(dgp, noise_sd, severity)
  ]
  raw[, method := "SRVF (raw)"]

  a_sub <- a_data[
    dgp %in% study_f_dgps & failure == FALSE,
    .(
      med = median(get(metric), na.rm = TRUE),
      q25 = quantile(get(metric), 0.25, na.rm = TRUE),
      q75 = quantile(get(metric), 0.75, na.rm = TRUE)
    ),
    by = .(dgp, noise_sd, severity, method)
  ]
  a_sub[, method := as.character(method)]

  dt <- rbindlist(list(preproc, raw, a_sub), use.names = TRUE)
  dt <- dt[!is.na(method)]
  dt[, method := factor(method, levels = c(cmp_method_levels, "srvf"))]
  dt <- dt[method != "srvf"]
  dt[, method := droplevels(method)]
  dt
}

plot_cmp <- function(dt, metric_label, log_scale = TRUE) {
  p <- ggplot(
    dt[as.numeric(as.character(noise_sd)) > 0],
    aes(y = dgp, x = med, color = method)
  ) +
    geom_pointrange(
      aes(xmin = q25, xmax = q75),
      position = position_dodge(width = 0.7),
      size = 0.3,
      linewidth = 0.5
    ) +
    facet_grid(noise_sd ~ severity, labeller = labeller(
      severity = function(x) paste0("severity = ", x),
      noise_sd = function(x) paste0("noise = ", x)
    )) +
    scale_color_manual(values = cmp_colors, labels = cmp_labels) +
    labs(y = NULL, x = metric_label, color = "Method") +
    theme_benchmark(base_size = 9)
  if (log_scale) {
    rng <- range(dt[as.numeric(as.character(noise_sd)) > 0, med], na.rm = TRUE)
    brks <- 10^seq(floor(log10(max(rng[1], 1e-6))), ceiling(log10(rng[2])))
    brks <- sort(unique(c(brks, brks * 5)))
    brks <- brks[brks >= rng[1] * 0.5 & brks <= rng[2] * 2]
    p <- p + scale_x_log10(breaks = brks, labels = brks)
  }
  p
}
```

## Warp MISE

```{r}
#| label: study-a-warp-mise
#| fig-width: 10
#| fig-height: 8
#| eval: !expr has_study_a
cmp_warp <- build_cmp_dt("warp_mise")
plot_cmp(cmp_warp, "Warp MISE (median + IQR)")
```

## Alignment Error

```{r}
#| label: study-a-alignment-error
#| fig-width: 10
#| fig-height: 8
#| eval: !expr has_study_a
cmp_ae <- build_cmp_dt("alignment_error")
plot_cmp(cmp_ae, "Alignment error (median + IQR)")
```

## Template MISE

```{r}
#| label: study-a-template-mise
#| fig-width: 10
#| fig-height: 8
#| eval: !expr has_study_a
cmp_tmise <- build_cmp_dt("template_mise")
plot_cmp(cmp_tmise, "Template MISE (median + IQR)")
```

## Template Elastic Distance

```{r}
#| label: study-a-elastic-dist
#| fig-width: 10
#| fig-height: 8
#| eval: !expr has_study_a
cmp_edist <- build_cmp_dt("template_elastic_dist")
plot_cmp(cmp_edist, "Template elastic distance (median + IQR)")
```

## Computation Time

```{r}
#| label: study-a-time
#| fig-width: 10
#| fig-height: 8
#| eval: !expr has_study_a
cmp_time <- build_cmp_dt("time")
plot_cmp(cmp_time, "Computation time in seconds (median + IQR)")
```

## Summary Table

```{r}
#| label: study-a-comparison-table
#| results: asis
#| eval: !expr has_study_a
metrics_list <- list(
  warp_mise = cmp_warp,
  alignment_error = cmp_ae,
  template_mise = cmp_tmise,
  elastic_dist = cmp_edist,
  time = cmp_time
)

for (metric_name in names(metrics_list)) {
  cmp_pooled <- metrics_list[[metric_name]][
    as.numeric(as.character(noise_sd)) > 0,
    .(value = median(med, na.rm = TRUE)),
    by = .(method, noise_sd)
  ]
  cat("\n")
  print(knitr::kable(
    dcast(cmp_pooled, method ~ noise_sd, value.var = "value"),
    digits = 4,
    caption = sprintf(
      "Pooled median %s across 9 Study F DGPs by noise level.",
      metric_name
    )
  ))
  cat("\n")
}
```

```{r}
#| label: study-a-missing
#| results: asis
if (!has_study_a) {
  cat("Study A results not available locally. Skipping cross-study comparison.\n")
}
```

# Timing

```{r}
#| label: timing
#| results: asis
time_dt <- f_ok[,
  .(median_time = median(time, na.rm = TRUE)),
  by = preproc_id
][order(median_time)]

time_overhead <- max(time_dt$median_time) - min(time_dt$median_time)
cat(sprintf("Pre-smoothing adds ~%.1fs overhead per run:\n\n", time_overhead))
knitr::kable(time_dt, digits = 2, col.names = c("Variant", "Median time (s)"))
```

# Conclusions

```{r}
#| label: conclusions
#| results: asis
# Compute summary statistics for data-driven conclusions
overall_noisy <- ratio_dt[
  as.numeric(as.character(noise_sd)) > 0,
  .(
    median_reduction = 1 - exp(median(log_ratio))
  ),
  by = preproc_id
]
best_noisy <- overall_noisy[which.max(median_reduction)]

clean_harm <- ratio_dt[
  as.numeric(as.character(noise_sd)) == 0,
  .(median_inflation = exp(median(log_ratio))),
  by = preproc_id
]
best_clean <- clean_harm[which.min(median_inflation)]

time_none <- f_ok[preproc_id == "none", median(time, na.rm = TRUE)]
time_max <- f_ok[, max(median(time, na.rm = TRUE)), by = preproc_id]$V1 |> max()
time_overhead <- time_max - time_none

cat(sprintf(
  "1. **Pre-smoothing strongly rescues SRVF under noise**: The best variant (%s) reduces median warp MISE by %.0f%% under noise (pooled across noise > 0).\n\n",
  best_noisy$preproc_id, 100 * best_noisy$median_reduction
))

cat(sprintf(
  "2. **Pre-smoothing hurts in clean settings**: Even the least harmful variant (%s) inflates warp MISE by %.2fx at noise = 0.\n\n",
  best_clean$preproc_id, best_clean$median_inflation
))

cat("3. **Neither primary variant qualifies as an unconditional default**: Both fail too many control cells for non-inferiority, primarily because of the clean-data penalty.\n\n")

cat("4. **Noise-adaptive recommendation**: Pre-smoothing with penalized splines (k=25) is strongly beneficial when noise is known to be present (noise >= 0.1 for weak DGPs, noise >= 0.3 for all DGPs). It should be offered as an option with guidance, not as the default.\n\n")

cat("5. **Spline variants dominate LOWESS**: Penalized splines (k=25) produce less clean-data harm and comparable or better noise rescue than LOWESS smoothing.\n\n")

cat(sprintf(
  "6. **Timing is not a concern**: Pre-smoothing adds ~%.1fs overhead per run.\n",
  time_overhead
))
```
