---
title: "Registration Benchmark v2 — Study E: Outlier Contamination"
subtitle: "Internal Report — tf Package"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  cache: true
  warning: false
  message: false
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| cache: false
library(ggplot2)
library(data.table)
library(patchwork)

base_dir <- here::here()
source(file.path(base_dir, "sim-analyze.R"))

mcols <- method_colors()
mlabs <- method_labels()
dlabs <- dgp_short_labels()
```

```{r}
#| label: load-data
dt <- load_study_e()

# Load Study A baseline (matching DGPs, noise=0.1/0.3, severity=0.5)
dt_a <- load_study_a()
baseline_dgps <- levels(dt$dgp)
dt_baseline <- dt_a[
  dgp %in% baseline_dgps &
  noise_sd %in% c("0.1", "0.3") &
  severity == "0.5"
]
dt_baseline[, contam_frac := NULL]
dt_baseline[, contam_frac := factor("0")]
dt_baseline[, outlier_type := NULL]
dt_baseline[, outlier_type := factor(NA, levels = c("shape", "phase"))]
```

Study E tests method robustness when a fraction of curves are outliers.
Design: 3 DGPs $\times$ 2 outlier types (shape, phase) $\times$ 3 fractions
(10%, 20%, 30%) $\times$ 5 methods $\times$ 2 noise (0.1, 0.3) $\times$
severity = 0.5 $\times$ 50 reps = 9,000 runs.

Study A results at matching conditions (contam_frac = 0) serve as baseline.

```{r}
#| label: overview
#| results: asis
cat(sprintf("- Rows: %d | Failures: %d (%.1f%%)\n",
  nrow(dt), sum(dt$failure), 100 * mean(dt$failure)))
cat(sprintf("- DGPs: %s\n", paste(levels(dt$dgp), collapse = ", ")))
cat(sprintf("- Methods: %s\n", paste(levels(dt$method), collapse = ", ")))
cat(sprintf("- Fractions: %s\n",
  paste(levels(dt$contam_frac), collapse = ", ")))
cat(sprintf("- Outlier types: %s\n",
  paste(levels(dt$outlier_type), collapse = ", ")))
cat(sprintf("- Noise: %s\n", paste(levels(dt$noise_sd), collapse = ", ")))
cat(sprintf("- Baseline rows from Study A: %d\n", nrow(dt_baseline)))
```

# DGP Gallery {#sec-gallery}

Example realizations with 20% contamination (noise = 0.1, severity = 0.5).
Outlier curves shown in red. Each row shows the template, warping functions
$h_i(t)$ (dashed = identity), and the resulting observed curves $x_i(t)$.

For **phase outliers**, the warps panel shows the actual extreme warps
(severity = 3.0, red) used to generate the contaminated curves, alongside
the original inlier warps (black).

**Design note (v2):** Shape outlier curves are standardized to match the
pointwise mean and standard deviation of inlier curves, so they are shape
outliers only (not magnitude outliers). Both shape and phase outlier curves
have the same noise level as inliers.

::: {.panel-tabset}

## Shape outliers

```{r}
#| label: gallery-shape
#| fig-height: 9
#| fig-cap: "Shape outliers: random sinusoids standardized to match inlier pointwise distribution."
make_contam_example("D02", "shape") /
  make_contam_example("D09", "shape") /
  make_contam_example("D12", "shape")
```

## Phase outliers

```{r}
#| label: gallery-phase
#| fig-height: 9
#| fig-cap: "Phase outliers: extreme warps (severity=3.0) applied to contaminated curves."
make_contam_example("D02", "phase") /
  make_contam_example("D09", "phase") /
  make_contam_example("D12", "phase")
```

:::

# Failure Rates

Contamination may cause methods to fail more often, biasing
conditional-on-success comparisons. This section checks whether failure
rates increase with contamination fraction.

```{r}
#| label: failure-by-contam
#| results: asis
fail_by_contam <- dt[,
  .(n_total = .N,
    n_fail = sum(failure),
    rate_pct = sprintf("%.1f", 100 * mean(failure))),
  by = .(method, outlier_type, contam_frac)
]
fail_wide <- dcast(fail_by_contam,
  method + outlier_type ~ contam_frac,
  value.var = "rate_pct")
knitr::kable(fail_wide,
  caption = "Failure rate (%) by method, outlier type, and contamination fraction.")
```

```{r}
#| label: failure-by-contam-findings
#| results: asis
fail_rates <- dt[, .(rate = mean(failure)), by = .(method, contam_frac)]
worst <- fail_rates[which.max(rate)]
if (worst$rate > 0.05) {
  cat(sprintf(
    "**Warning**: %s has %.1f%% failure rate at %s contamination. ",
    worst$method, 100 * worst$rate, worst$contam_frac))
  cat("Conditional-on-success comparisons may be biased for this method.\n\n")
} else {
  cat(sprintf(
    "All failure rates are below 5%%. Maximum: %s at %.1f%% (contam = %s).\n\n",
    worst$method, 100 * worst$rate, worst$contam_frac))
}
# Baseline failure rate
base_fail <- dt_baseline[dgp %in% baseline_dgps & severity == "0.5",
  .(rate = mean(failure)), by = method]
contam_fail <- dt[, .(rate = mean(failure)), by = method]
merged_fail <- merge(base_fail, contam_fail, by = "method",
  suffixes = c("_base", "_contam"))
merged_fail[, increase := rate_contam - rate_base]
if (any(merged_fail$increase > 0.02)) {
  cat("Methods with elevated failure rates vs baseline (>2pp increase):\n\n")
  elevated <- merged_fail[increase > 0.02]
  for (i in seq_len(nrow(elevated))) {
    cat(sprintf("- %s: %.1f%% → %.1f%% (+%.1fpp)\n",
      elevated$method[i],
      100 * elevated$rate_base[i],
      100 * elevated$rate_contam[i],
      100 * elevated$increase[i]))
  }
} else {
  cat("No method shows a failure rate increase > 2 percentage points vs baseline.\n\n")
}
```

# Warp MISE Degradation

How does warp recovery degrade as contamination fraction increases?
Baseline (0%) from Study A at matching conditions (noise = 0.1/0.3,
severity = 0.5, no contamination).

## Degradation curves by DGP

```{r}
#| label: warp-mise-degradation
#| fig-height: 10
#| fig-cap: "Warp MISE vs contamination fraction. Points show medians and vertical bars show IQRs, with separate panels by DGP and noise level."
# Combine contaminated + baseline for degradation curves
dt_ok <- dt[failure == FALSE]
dt_base_ok <- dt_baseline[failure == FALSE]

# Aggregate: median warp MISE by (method, dgp, noise, outlier_type, fraction)
contam_agg <- aggregate_median_iqr(
  dt = dt_ok,
  metric = "warp_mise",
  by = c("method", "dgp", "noise_sd", "outlier_type", "contam_frac"),
  value_name = "warp_mise"
)

base_agg <- aggregate_median_iqr(
  dt = dt_base_ok,
  metric = "warp_mise",
  by = c("method", "dgp", "noise_sd"),
  value_name = "warp_mise"
)

# Plot: one panel per DGP x noise, x = fraction, y = MISE, color = method
# Separate by outlier type (facet rows)
plots <- list()
for (otype in c("shape", "phase")) {
  sub <- contam_agg[outlier_type == otype]

  # Add baseline as fraction = 0
  base_for_plot <- copy(base_agg)
  base_for_plot[, outlier_type := otype]
  base_for_plot[, contam_frac := "0"]
  combined <- rbind(sub, base_for_plot, fill = TRUE)
  combined[, frac_num := as.numeric(as.character(contam_frac))]

  p <- ggplot(combined,
    aes(x = frac_num, y = warp_mise, color = method, group = method)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.015, alpha = 0.3) +
    facet_grid(dgp ~ noise_sd,
      scales = "free_y",
      labeller = labeller(
        noise_sd = function(x) paste0("noise = ", x),
        dgp = dlabs)) +
    scale_color_manual(values = mcols, labels = mlabs) +
    scale_x_continuous(
      breaks = c(0, 0.1, 0.2, 0.3),
      labels = c("0%", "10%", "20%", "30%")) +
    scale_y_log10() +
    labs(
      title = paste0("Warp MISE vs contamination fraction — ",
        otype, " outliers"),
      x = "Contamination fraction",
      y = "Median warp MISE (log scale)",
      color = "Method") +
    theme_benchmark()

  plots[[otype]] <- p
}

plots[["shape"]] / plots[["phase"]] +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## Degradation ratios

```{r}
#| label: degradation-ratios
#| results: asis
# Compute degradation ratio: MISE(contaminated) / MISE(clean)
contam_ratios <- merge(contam_agg, base_agg,
  by = c("method", "dgp", "noise_sd"),
  suffixes = c("_contam", "_base"))
contam_ratios[, ratio := warp_mise_contam / warp_mise_base]

# Summary by method, outlier_type, fraction
ratio_summary <- contam_ratios[,
  .(
    median_ratio = round(median(ratio, na.rm = TRUE), 2),
    q25 = round(quantile(ratio, 0.25, na.rm = TRUE), 2),
    q75 = round(quantile(ratio, 0.75, na.rm = TRUE), 2)
  ),
  by = .(method, outlier_type, contam_frac)
]

cat("Degradation ratio (contaminated / baseline warp MISE) ",
    "— median [Q25, Q75] across DGPs and noise levels:\n\n")

for (otype in c("shape", "phase")) {
  cat(sprintf("\n### %s outliers\n\n", tools::toTitleCase(otype)))
  sub <- ratio_summary[outlier_type == otype]
  sub[, display := sprintf("%.2f [%.2f, %.2f]", median_ratio, q25, q75)]
  wide <- dcast(sub, method ~ contam_frac, value.var = "display")
  print(knitr::kable(wide))
  cat("\n")
}
```

```{r}
#| label: degradation-findings
#| results: asis
# Which outlier type is more damaging?
type_comparison <- contam_ratios[,
  .(median_ratio = median(ratio, na.rm = TRUE)),
  by = .(method, outlier_type, contam_frac)
]
type_wide <- dcast(type_comparison,
  method + contam_frac ~ outlier_type,
  value.var = "median_ratio")
type_wide[, worse_type := ifelse(shape > phase, "shape", "phase")]

worse_counts <- type_wide[, .N, by = worse_type]
dominant <- worse_counts[which.max(N)]$worse_type
pct_dominant <- round(100 * max(worse_counts$N) / sum(worse_counts$N))

cat(sprintf(
  "**%s outliers** are more damaging in %d%% (%d/%d) of method × fraction cells.\n\n",
  tools::toTitleCase(dominant), pct_dominant,
  max(worse_counts$N), sum(worse_counts$N)))

# Per-method breakdown
cat("Per-method dominant outlier type (across fractions):\n\n")
method_dominant <- type_wide[,
  .(n_shape = sum(worse_type == "shape"),
    n_phase = sum(worse_type == "phase")),
  by = method]
method_dominant[, dominant := ifelse(n_shape > n_phase, "shape", "phase")]
for (i in seq_len(nrow(method_dominant))) {
  cat(sprintf("- **%s**: %s outliers more damaging (%d/%d fractions)\n",
    mlabs[as.character(method_dominant$method[i])],
    method_dominant$dominant[i],
    pmax(method_dominant$n_shape[i], method_dominant$n_phase[i]),
    method_dominant$n_shape[i] + method_dominant$n_phase[i]))
}
```

# Shape vs Phase Outlier Comparison

Direct comparison of outlier types at each contamination fraction.
Ratio > 1 means shape outliers are more damaging than phase outliers.

```{r}
#| label: shape-vs-phase
#| fig-height: 8
# Ratio of shape MISE to phase MISE (>1 means shape is worse)
shape_phase <- dcast(contam_agg,
  method + dgp + noise_sd + contam_frac ~ outlier_type,
  value.var = "warp_mise")
shape_phase[, ratio := shape / phase]

ggplot(shape_phase,
  aes(x = method, y = ratio, color = method)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  geom_jitter(width = 0.15, size = 2, alpha = 0.6) +
  stat_summary(fun = median, geom = "crossbar",
    width = 0.4, color = "black", linewidth = 0.4) +
  facet_wrap(~ contam_frac,
    labeller = labeller(contam_frac = function(x) paste0(x, " contam."))) +
  scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_x_discrete(labels = mlabs) +
  scale_y_log10() +
  labs(
    title = "Shape / phase outlier warp MISE ratio",
    subtitle = "> 1 means shape outliers are more damaging",
    x = NULL,
    y = "MISE ratio (shape / phase, log scale)") +
  theme_benchmark() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: shape-phase-findings
#| results: asis
# Per-method median shape/phase ratio at 30%
sp_at_30 <- shape_phase[contam_frac == "0.3",
  .(
    mean_ratio = mean(ratio, na.rm = TRUE),
    se_ratio = if (.N > 1) sd(ratio, na.rm = TRUE) / sqrt(.N) else 0,
    median_ratio = round(median(ratio, na.rm = TRUE), 2)
  ),
  by = method]

cat("At 30% contamination, the median shape/phase warp MISE ratio by method:\n\n")
for (i in seq_len(nrow(sp_at_30))) {
  m <- as.character(sp_at_30$method[i])
  r <- sp_at_30$median_ratio[i]
  direction <- if (r > 1) "shape worse" else "phase worse"
  cat(sprintf("- **%s**: median %.2f×; mean %s (%s)\n",
    mlabs[m], r, fmt_mean_se(sp_at_30$mean_ratio[i], sp_at_30$se_ratio[i]), direction))
}

cat("\nThe pattern is consistent across contamination fractions: **iterative ",
    "template-based methods** (SRVF, FDA) are more damaged by phase outliers, ",
    "which pull the iterative template estimate toward extreme warps. ",
    "**Landmark** is roughly equally affected by both types. ",
    "**Affine** is approximately equally sensitive to both, likely because ",
    "its shift+scale model cannot accommodate either type of outlier.\n")
```

# Noise × Contamination Interaction

Does high noise compound with contamination? If yes, robustness at
noise = 0.1 overestimates practical robustness.

```{r}
#| label: noise-contam-interaction
#| fig-height: 8
#| fig-cap: "Noise × contamination interaction. Points show median degradation ratios and vertical bars show IQRs across DGPs."
# Compare degradation ratios at noise=0.1 vs noise=0.3
noise_interact <- contam_ratios[,
  .(
    median_ratio = median(ratio, na.rm = TRUE),
    q25 = quantile(ratio, 0.25, na.rm = TRUE),
    q75 = quantile(ratio, 0.75, na.rm = TRUE)
  ),
  by = .(method, noise_sd, outlier_type, contam_frac)
]

ggplot(noise_interact,
  aes(x = as.numeric(as.character(contam_frac)),
      y = median_ratio,
      color = noise_sd,
      group = noise_sd)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.015, alpha = 0.3) +
  facet_grid(outlier_type ~ method,
    labeller = labeller(method = mlabs)) +
  scale_color_manual(
    values = c("0.1" = "#2166AC", "0.3" = "#B2182B"),
    labels = c("0.1" = "noise = 0.1", "0.3" = "noise = 0.3")) +
  scale_x_continuous(
    breaks = c(0.1, 0.2, 0.3),
    labels = c("10%", "20%", "30%")) +
  labs(
    title = "Noise × contamination interaction",
    subtitle = "Degradation ratio (contaminated / baseline) by noise level",
    x = "Contamination fraction",
    y = "Degradation ratio",
    color = NULL) +
  theme_benchmark()
```

```{r}
#| label: noise-contam-findings
#| results: asis
# Compare degradation at noise=0.1 vs 0.3
noise_comparison <- dcast(noise_interact,
  method + outlier_type + contam_frac ~ noise_sd,
  value.var = "median_ratio")
setnames(noise_comparison, c("0.1", "0.3"), c("low_noise", "high_noise"))
noise_comparison[, amplification := high_noise / low_noise]

# Summarize: does high noise amplify contamination effects?
amp_summary <- noise_comparison[,
  .(median_amp = round(median(amplification, na.rm = TRUE), 2)),
  by = method]

cat("Noise amplification of contamination effects ",
    "(ratio of degradation at noise=0.3 vs 0.1):\n\n")
for (i in seq_len(nrow(amp_summary))) {
  m <- as.character(amp_summary$method[i])
  amp <- amp_summary$median_amp[i]
  if (amp > 1.2) {
    txt <- sprintf("amplification (%.2f×)", amp)
  } else if (amp < 0.8) {
    txt <- sprintf("attenuation (%.2f×)", amp)
  } else {
    txt <- sprintf("roughly constant (%.2f×)", amp)
  }
  cat(sprintf("- **%s**: %s\n", mlabs[m], txt))
}

cat("\n**Interpretation**: The degradation ratio measures *relative* worsening ",
    "vs each method's own baseline at that noise level. ",
    "SRVF shows apparent attenuation at high noise because its baseline MISE ",
    "is already elevated at noise=0.3 (SRVF operates on derivatives, amplifying noise). ",
    "The *absolute* MISE at noise=0.3 with contamination is still worse — ",
    "the ratio just appears smaller because the denominator is larger.\n")
```

# Template Bias Under Contamination

Outliers can bias template estimation, degrading all downstream
warp estimates. This section examines template quality via L$^2$ MISE
and Fisher-Rao elastic distance.

**Context**: Template MISE rankings are strongly DGP-dependent (see Study A).
SRVF has high template MISE on harmonic DGPs (D02, D09) but low on
wiggly DGPs (D12). The per-DGP panels below show this clearly — method
rankings at baseline (0%) differ across DGPs, and contamination effects
compound on top of these baseline differences.

## Template MISE

```{r}
#| label: template-mise-contam
#| fig-height: 8
#| fig-cap: "Template MISE vs contamination fraction. Points show medians and vertical bars show IQRs."
# Template MISE with baseline
tmise_contam <- aggregate_median_iqr(
  dt = dt_ok,
  metric = "template_mise",
  by = c("method", "dgp", "noise_sd", "outlier_type", "contam_frac"),
  value_name = "template_mise"
)
tmise_base <- aggregate_median_iqr(
  dt = dt_base_ok,
  metric = "template_mise",
  by = c("method", "dgp", "noise_sd"),
  value_name = "template_mise"
)

plots_tmise <- list()
for (otype in c("shape", "phase")) {
  sub <- tmise_contam[outlier_type == otype]
  base_for_plot <- copy(tmise_base)
  base_for_plot[, outlier_type := otype]
  base_for_plot[, contam_frac := "0"]
  combined <- rbind(sub, base_for_plot, fill = TRUE)
  combined[, frac_num := as.numeric(as.character(contam_frac))]

  p <- ggplot(combined,
    aes(x = frac_num, y = template_mise, color = method, group = method)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.015, alpha = 0.3) +
    facet_grid(dgp ~ noise_sd,
      scales = "free_y",
      labeller = labeller(
        noise_sd = function(x) paste0("noise = ", x),
        dgp = dlabs)) +
    scale_color_manual(values = mcols, labels = mlabs) +
    scale_x_continuous(
      breaks = c(0, 0.1, 0.2, 0.3),
      labels = c("0%", "10%", "20%", "30%")) +
    scale_y_log10() +
    labs(
      title = paste0("Template MISE — ", otype, " outliers"),
      x = "Contamination fraction",
      y = "Median template MISE (log scale)",
      color = "Method") +
    theme_benchmark()

  plots_tmise[[otype]] <- p
}

plots_tmise[["shape"]] / plots_tmise[["phase"]] +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## Template elastic distance

```{r}
#| label: template-edist-contam
#| fig-height: 8
#| fig-cap: "Template elastic distance vs contamination fraction. Points show medians and vertical bars show IQRs."
edist_contam <- aggregate_median_iqr(
  dt = dt_ok[!is.na(template_elastic_dist)],
  metric = "template_elastic_dist",
  by = c("method", "dgp", "noise_sd", "outlier_type", "contam_frac"),
  value_name = "edist"
)
edist_base <- aggregate_median_iqr(
  dt = dt_base_ok[!is.na(template_elastic_dist)],
  metric = "template_elastic_dist",
  by = c("method", "dgp", "noise_sd"),
  value_name = "edist"
)

plots_edist <- list()
for (otype in c("shape", "phase")) {
  sub <- edist_contam[outlier_type == otype]
  base_for_plot <- copy(edist_base)
  base_for_plot[, outlier_type := otype]
  base_for_plot[, contam_frac := "0"]
  combined <- rbind(sub, base_for_plot, fill = TRUE)
  combined[, frac_num := as.numeric(as.character(contam_frac))]

  p <- ggplot(combined,
    aes(x = frac_num, y = edist, color = method, group = method)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.015, alpha = 0.3) +
    facet_grid(dgp ~ noise_sd,
      scales = "free_y",
      labeller = labeller(
        noise_sd = function(x) paste0("noise = ", x),
        dgp = dlabs)) +
    scale_color_manual(values = mcols, labels = mlabs) +
    scale_x_continuous(
      breaks = c(0, 0.1, 0.2, 0.3),
      labels = c("0%", "10%", "20%", "30%")) +
    labs(
      title = paste0("Template elastic distance — ", otype, " outliers"),
      x = "Contamination fraction",
      y = "Median elastic distance",
      color = "Method") +
    theme_benchmark()

  plots_edist[[otype]] <- p
}

plots_edist[["shape"]] / plots_edist[["phase"]] +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

## Case study: SRVF template improvement under contamination

The elastic distance results show SRVF's template quality can paradoxically
*improve* under contamination. This section visualizes a specific instance:
all five methods are run on the same dataset with and without contamination.
Black solid line = true template; colored dashed line = estimated template;
grey lines = aligned curves.

```{r}
#| label: case-study-setup
#| fig-height: 7
#| fig-width: 12
# D12 (wiggly/GP/high-rank) + phase outliers, seed=7: SRVF elastic distance
# *decreases* from 0.703 to 0.557 (ratio 0.79) while template MISE doubles.
cs <- make_contam_case_study(
  dgp_name = "D12",
  outlier_type = "phase",
  contam_frac = 0.20,
  noise_sd = 0.1,
  severity = 0.5,
  n = 50,
  seed = 7
)
cs$plot
```

```{r}
#| label: case-study-table
mdf <- cs$metrics
mdf$tmise_ratio <- mdf$tmise_contam / mdf$tmise_clean
mdf$edist_ratio <- mdf$edist_contam / mdf$edist_clean

display <- data.frame(
  Method = mdf$method,
  `MISE clean` = fmt(mdf$tmise_clean),
  `MISE contam` = fmt(mdf$tmise_contam),
  `MISE ratio` = sprintf("%.2f\u00d7", mdf$tmise_ratio),
  `Elastic clean` = fmt(mdf$edist_clean),
  `Elastic contam` = fmt(mdf$edist_contam),
  `Elastic ratio` = sprintf("%.2f\u00d7", mdf$edist_ratio),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(display, align = "lcccccc",
  caption = "Template quality for clean vs contaminated data (D12 wiggly/GP/high-rank, 20% phase, noise=0.1).")
```

```{r}
#| label: case-study-findings
#| results: asis
srvf_row <- mdf[mdf$method == "SRVF", ]
if (srvf_row$edist_ratio < 1) {
  cat(sprintf(
    "**SRVF's elastic distance decreases from %s to %s (ratio %.2f\u00d7)**. ",
    fmt(srvf_row$edist_clean), fmt(srvf_row$edist_contam),
    srvf_row$edist_ratio))
  cat("This single realization is consistent with the aggregate phase-outlier ",
      "pattern, but should be read as an illustration rather than proof of mechanism. ",
      "A plausible interpretation is that SRVF absorbs more of the extreme warping ",
      "into phase than into the template in this draw.\n")
} else {
  cat(sprintf(
    "In this particular realization, SRVF elastic distance ratio is %.2f\u00d7. ",
    srvf_row$edist_ratio))
  cat("The paradoxical improvement is a stochastic phenomenon that appears ",
      "in aggregate across many replications.\n")
}
```

```{r}
#| label: template-bias-findings
#| results: asis
# Template MISE degradation ratios
tmise_ratios <- merge(tmise_contam, tmise_base,
  by = c("method", "dgp", "noise_sd"),
  suffixes = c("_contam", "_base"))
tmise_ratios[, ratio := template_mise_contam / template_mise_base]

tmise_ratio_summary <- tmise_ratios[,
  .(median_ratio = round(median(ratio, na.rm = TRUE), 2)),
  by = .(method, outlier_type, contam_frac)
]

# Elastic distance degradation ratios
edist_ratios <- merge(edist_contam, edist_base,
  by = c("method", "dgp", "noise_sd"),
  suffixes = c("_contam", "_base"))
edist_ratios[, ratio := edist_contam / edist_base]

edist_ratio_summary <- edist_ratios[,
  .(median_ratio = round(median(ratio, na.rm = TRUE), 2)),
  by = .(method, outlier_type, contam_frac)
]

cat("### Degradation ratios at 30% contamination\n\n")
cat("Template MISE and elastic distance degradation ",
    "(contaminated / baseline):\n\n")

for (otype in c("shape", "phase")) {
  cat(sprintf("\n**%s outliers:**\n\n", tools::toTitleCase(otype)))
  tmise_sub <- tmise_ratio_summary[
    contam_frac == "0.3" & outlier_type == otype]
  edist_sub <- edist_ratio_summary[
    contam_frac == "0.3" & outlier_type == otype]
  merged <- merge(tmise_sub[, .(method, tmise = median_ratio)],
    edist_sub[, .(method, edist = median_ratio)], by = "method")
  merged <- merged[order(-tmise)]
  cat("| Method | Template MISE | Elastic distance |\n")
  cat("|--------|:---:|:---:|\n")
  for (i in seq_len(nrow(merged))) {
    m <- as.character(merged$method[i])
    cat(sprintf("| %s | %.2f× | %.2f× |\n",
      mlabs[m], merged$tmise[i], merged$edist[i]))
  }
  cat("\n")
}

cat("**Key observation**: Template MISE and elastic distance tell very different ",
    "stories. Landmark has the worst template MISE degradation (up to ~28× for ",
    "shape outliers) but only moderate elastic distance degradation (~2×). This ",
    "means outliers shift the template *position* (L² error) without greatly ",
    "distorting its *shape* (elastic distance). Conversely, SRVF's elastic ",
    "distance sometimes *decreases* under phase contamination (ratio < 1) in the ",
    "aggregate summaries, suggesting a phase/template tradeoff rather than simple ",
    "uniform degradation.\n")
```

# Alignment Error Degradation

Alignment error measures how well curves are aligned to their true (pre-warp)
positions — the end goal of registration. This is a more direct measure of
practical quality than warp MISE.

```{r}
#| label: alignment-error-degradation
#| fig-height: 10
#| fig-cap: "Alignment error vs contamination fraction. Points show medians and vertical bars show IQRs, with separate panels by DGP and noise level."
# Alignment error aggregation
aerr_contam <- aggregate_median_iqr(
  dt = dt_ok,
  metric = "alignment_error",
  by = c("method", "dgp", "noise_sd", "outlier_type", "contam_frac"),
  value_name = "alignment_error"
)
aerr_base <- aggregate_median_iqr(
  dt = dt_base_ok,
  metric = "alignment_error",
  by = c("method", "dgp", "noise_sd"),
  value_name = "alignment_error"
)

plots_aerr <- list()
for (otype in c("shape", "phase")) {
  sub <- aerr_contam[outlier_type == otype]
  base_for_plot <- copy(aerr_base)
  base_for_plot[, outlier_type := otype]
  base_for_plot[, contam_frac := "0"]
  combined <- rbind(sub, base_for_plot, fill = TRUE)
  combined[, frac_num := as.numeric(as.character(contam_frac))]

  p <- ggplot(combined,
    aes(x = frac_num, y = alignment_error, color = method, group = method)) +
    geom_line(linewidth = 0.7) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.015, alpha = 0.3) +
    facet_grid(dgp ~ noise_sd,
      scales = "free_y",
      labeller = labeller(
        noise_sd = function(x) paste0("noise = ", x),
        dgp = dlabs)) +
    scale_color_manual(values = mcols, labels = mlabs) +
    scale_x_continuous(
      breaks = c(0, 0.1, 0.2, 0.3),
      labels = c("0%", "10%", "20%", "30%")) +
    scale_y_log10() +
    labs(
      title = paste0("Alignment error — ", otype, " outliers"),
      x = "Contamination fraction",
      y = "Median alignment error (log scale)",
      color = "Method") +
    theme_benchmark()

  plots_aerr[[otype]] <- p
}

plots_aerr[["shape"]] / plots_aerr[["phase"]] +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: alignment-error-ratios
#| results: asis
# Compute alignment error degradation ratios
aerr_ratios <- merge(aerr_contam, aerr_base,
  by = c("method", "dgp", "noise_sd"),
  suffixes = c("_contam", "_base"))
aerr_ratios[, ratio := alignment_error_contam / alignment_error_base]

aerr_ratio_summary <- aerr_ratios[,
  .(
    median_ratio = round(median(ratio, na.rm = TRUE), 2),
    q25 = round(quantile(ratio, 0.25, na.rm = TRUE), 2),
    q75 = round(quantile(ratio, 0.75, na.rm = TRUE), 2)
  ),
  by = .(method, outlier_type, contam_frac)
]

cat("Alignment error degradation ratio (contaminated / baseline) ",
    "— median [Q25, Q75]:\n\n")
for (otype in c("shape", "phase")) {
  cat(sprintf("\n### %s outliers\n\n", tools::toTitleCase(otype)))
  sub <- aerr_ratio_summary[outlier_type == otype]
  sub[, display := sprintf("%.2f [%.2f, %.2f]", median_ratio, q25, q75)]
  wide <- dcast(sub, method ~ contam_frac, value.var = "display")
  print(knitr::kable(wide))
  cat("\n")
}
```

```{r}
#| label: alignment-error-findings
#| results: asis
# Compare alignment error vs warp MISE rankings
aerr_at_30 <- aerr_ratios[contam_frac == "0.3",
  .(aerr_ratio = median(ratio)), by = method][order(aerr_ratio)]
wmise_at_30 <- contam_ratios[contam_frac == "0.3",
  .(wmise_ratio = median(ratio)), by = method][order(wmise_ratio)]

merged_rank <- merge(aerr_at_30, wmise_at_30, by = "method")
merged_rank[, aerr_rank := rank(aerr_ratio)]
merged_rank[, wmise_rank := rank(wmise_ratio)]

cat("**Alignment error vs warp MISE robustness ranking at 30%:**\n\n")
cat("| Method | Alignment error | Warp MISE |\n")
cat("|--------|:---:|:---:|\n")
for (i in seq_len(nrow(merged_rank))) {
  m <- as.character(merged_rank$method[i])
  cat(sprintf("| %s | %.1f× (rank %d) | %.1f× (rank %d) |\n",
    mlabs[m],
    merged_rank$aerr_ratio[i], merged_rank$aerr_rank[i],
    merged_rank$wmise_ratio[i], merged_rank$wmise_rank[i]))
}

rank_concordance <- cor(merged_rank$aerr_rank, merged_rank$wmise_rank,
  method = "spearman")
cat(sprintf(
  "\nRank concordance (Spearman): %.2f. ", rank_concordance))
if (rank_concordance > 0.8) {
  cat("Alignment error and warp MISE robustness rankings are highly consistent.\n")
} else {
  cat("Rankings differ — method robustness depends on which metric is prioritized.\n")
}
```

# Method Robustness Ranking

Degradation heatmaps at 30% contamination across all primary metrics.

```{r}
#| label: robustness-heatmap-helper
#| fig-height: 6
# Helper function for degradation heatmaps
make_degrad_heatmap <- function(ratios_dt, metric_name) {
  worst_case <- ratios_dt[
    contam_frac == "0.3",
    .(median_ratio = median(ratio, na.rm = TRUE)),
    by = .(method, dgp, outlier_type)
  ]

  ggplot(worst_case,
    aes(x = method, y = interaction(outlier_type, dgp, sep = " | "),
        fill = median_ratio)) +
    geom_tile(color = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.1f×", median_ratio)),
      size = 2.5, color = "white") +
    scale_x_discrete(labels = mlabs) +
    scale_fill_gradient2(
      low = "#2166AC", mid = "#F7F7F7", high = "#B2182B",
      midpoint = 1, limits = c(0.5, NA),
      name = "Degradation\nratio") +
    labs(
      title = metric_name,
      x = NULL, y = NULL) +
    theme_benchmark() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Warp MISE

```{r}
#| label: robustness-heatmap-warp
#| fig-height: 6
make_degrad_heatmap(contam_ratios, "Warp MISE degradation (30% contamination)")
```

## Alignment error

```{r}
#| label: robustness-heatmap-alignment
#| fig-height: 6
make_degrad_heatmap(aerr_ratios, "Alignment error degradation (30% contamination)")
```

## Template MISE

```{r}
#| label: robustness-heatmap-template
#| fig-height: 6
make_degrad_heatmap(tmise_ratios, "Template MISE degradation (30% contamination)")
```

## Template elastic distance

```{r}
#| label: robustness-heatmap-edist
#| fig-height: 6
make_degrad_heatmap(edist_ratios, "Template elastic distance degradation (30% contamination)")
```

# Summary

```{r}
#| label: summary
#| results: asis
# Warp-MISE robustness ranking across all conditions
overall <- contam_ratios[,
  .(
    median_degrad_10 = round(median(ratio[contam_frac == "0.1"], na.rm = TRUE), 2),
    median_degrad_20 = round(median(ratio[contam_frac == "0.2"], na.rm = TRUE), 2),
    median_degrad_30 = round(median(ratio[contam_frac == "0.3"], na.rm = TRUE), 2),
    max_degrad = round(max(ratio, na.rm = TRUE), 1)
  ),
  by = method
]
overall_30_unc <- mean_se_table(contam_ratios[contam_frac == "0.3"], "ratio", "method")
overall <- merge(overall, overall_30_unc[, .(method, mean_degrad_30 = mean, se_degrad_30 = se)],
  by = "method")
overall <- overall[order(median_degrad_30)]

cat("**Warp-MISE robustness ranking** (median warp MISE degradation ratio, ",
    "lower = more robust):\n\n")
overall[, mean_se_30 := fmt_mean_se(mean_degrad_30, se_degrad_30)]
knitr::kable(
  overall[, .(method, median_degrad_10, median_degrad_20, median_degrad_30, mean_se_30, max_degrad)],
  col.names = c("Method", "10%", "20%", "30% median", "30% mean ± SE", "Worst case"),
  caption = "Warp-MISE degradation by contamination fraction. Worst case is max across all conditions.")
cat("\n")

most_robust <- mlabs[as.character(overall$method[1])]
least_robust <- mlabs[as.character(overall$method[nrow(overall)])]
cat(sprintf(
  "**%s** is the most robust method by warp MISE (%.2f× median degradation at 30%% contamination), ",
  most_robust, overall$median_degrad_30[1]))
cat(sprintf(
  "while **%s** is the least robust (%.2f×).\n\n",
  least_robust, overall$median_degrad_30[nrow(overall)]))

# Key findings
cat("### Key findings\n\n")

cat("1. **By warp MISE, Landmark is dramatically more robust** than all other methods. ",
    sprintf("At 30%% contamination, Landmark degrades %.1f× vs %.1f–%.1f× ",
      overall[method == "landmark_auto"]$median_degrad_30,
      overall[method != "landmark_auto", min(median_degrad_30)],
      overall[method != "landmark_auto", max(median_degrad_30)]),
    "for iterative methods. The corresponding 30%% mean ± SE is ",
    overall[method == "landmark_auto", mean_se_30],
    ". This is expected: landmarks are local features ",
    "that are robust to global contamination.\n\n")

cat("2. **Phase outliers are more damaging than shape outliers** for iterative ",
    "template-based methods (SRVF, FDA). Phase outliers warp the template estimate ",
    "via the iterative Procrustes loop, while shape outliers are partially averaged ",
    "out. Affine is equally affected by both types.\n\n")

contam_profile <- dcast(
  contam_ratios,
  method + dgp + noise_sd + outlier_type ~ contam_frac,
  value.var = "ratio"
)
contam_profile[, monotone := `0.1` <= `0.2` & `0.2` <= `0.3`]
monotone_n <- sum(contam_profile$monotone)
monotone_total <- nrow(contam_profile)
cat("3. **No method crashes under contamination.** 0% failure rate across all ",
    "9,000 runs, even at 30% contamination. Degradation is almost always gradual: ",
    monotone_n, "/", monotone_total,
    " method × DGP × noise × outlier profiles show monotone warp-MISE worsening, ",
    "with only a few near-flat landmark exceptions and no sharp breakdown threshold.\n\n")

srvf_noise <- contam_ratios[method == "srvf" & contam_frac == "0.3",
  .(r = median(ratio)), by = noise_sd]
cat(sprintf(
  "4. **SRVF shows an apparent noise-attenuation artifact.** Degradation ratio at 30%% is %.1f× at noise=0.1 but only %.1f× at noise=0.3. ",
  srvf_noise[noise_sd == "0.1"]$r,
  srvf_noise[noise_sd == "0.3"]$r),
  "This is because SRVF's baseline MISE is already poor at high noise ",
  "(derivative amplification), so the denominator is larger. The absolute ",
  "MISE still worsens with noise.\n\n")

cat("5. **Template MISE and elastic distance diverge under contamination.** ",
    "Landmark shows the worst template MISE degradation but only moderate ",
    "elastic distance degradation — outliers shift the template position ",
    "without distorting its shape. SRVF's elastic distance can even improve ",
    "under phase contamination in the aggregated ratios, pointing to a ",
    "phase/template tradeoff rather than uniform template deterioration.\n\n")

cat("6. **Practical implication**: For data that may contain outliers, ",
    "Landmark-based registration (or a preprocessing step to remove outliers) ",
    "is the safest choice if warp recovery is the primary target. But the ",
    "alignment-error ranking differs substantially from the warp-MISE ranking ",
    "(Spearman 0.10), so the preferred method remains metric-dependent. Among ",
    "iterative methods, SRVF is more robust than FDA methods to shape outliers ",
    "but more vulnerable to phase outliers.\n")
```
