---
title: "Registration Benchmark v2 — Study A"
subtitle: "Internal Report — tf Package"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  cache: true
  warning: false
  message: false
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| cache: false
library(ggplot2)
library(data.table)
library(patchwork)

base_dir <- here::here()
source(file.path(base_dir, "sim-analyze.R"))

mcols <- method_colors()
mlabs <- method_labels()
```

```{r}
#| label: load-data
dt <- load_study_a()
cat(sprintf(
  "DGPs: %s\nMethods: %s\nSeverity: %s | Noise: %s\n",
  paste(levels(dt$dgp), collapse = ", "),
  paste(levels(dt$method), collapse = ", "),
  paste(levels(dt$severity), collapse = ", "),
  paste(levels(dt$noise_sd), collapse = ", ")
))
```

Study A compares 5 registration methods across 13 DGPs in a crossed design:
13 DGPs $\times$ 2 severity (0.5, 1.0) $\times$ 2 noise (0, 0.1) $\times$
5 methods = 260 cells $\times$ 100 reps = 26,000 runs.

The DGP design is **not** full factorial (13 of 24 possible template $\times$ warp
$\times$ amplitude combinations), but enables **paired comparisons** that vary
exactly one factor while holding the other two constant.

```{r}
#| label: dgp-table
knitr::kable(
  dgp_factors(),
  caption = "DGP factor structure. 13 DGPs spanning 2 templates, 3 warp types, 4 amplitude types."
)
```

# DGP Gallery {#sec-gallery}

Example realizations for each DGP (n=20, severity=0.5, no noise). Each row shows
the template, the warping functions $h_i(t)$ (dashed = identity), and the resulting
observed curves $x_i(t)$.

::: {.panel-tabset}

## Baseline (D01--D04)

```{r}
#| label: gallery-baseline
#| fig-height: 12
#| fig-cap: "Baseline DGPs: phase variation only (no amplitude variation)."
make_dgp_example("D01") / make_dgp_example("D02") /
  make_dgp_example("D03") / make_dgp_example("D04")
```

## Low-rank amplitude (D05--D08)

```{r}
#| label: gallery-lowrank
#| fig-height: 12
#| fig-cap: "Low-rank amplitude DGPs: rank-1 multiplicative or rank-2 (scale + shift)."
make_dgp_example("D05") / make_dgp_example("D06") /
  make_dgp_example("D07") / make_dgp_example("D08")
```

## High-rank amplitude (D09--D13)

```{r}
#| label: gallery-highrank
#| fig-height: 15
#| fig-cap: "High-rank amplitude DGPs: 3 FPCs from gait knee data. D04/D13 use non-domain-preserving affine warps."
make_dgp_example("D09") / make_dgp_example("D10") /
  make_dgp_example("D11") / make_dgp_example("D12") /
  make_dgp_example("D13")
```

:::

# Overview: DGP $\times$ Method Heatmaps {#sec-overview}

The heatmaps below show the **median** of each metric across all 400 runs per cell
(100 reps $\times$ 2 severity $\times$ 2 noise), giving a quick orientation.
Lighter tiles indicate better performance.

```{r}
#| label: heatmap-warp-mise
#| fig-height: 7
#| fig-cap: "Median warp MISE across all severity/noise conditions. Lighter = lower MISE = better warp recovery."
make_heatmap(dt, "warp_mise", "Warp MISE (median, all conditions)")
```

```{r}
#| label: heatmap-alignment-cc
#| fig-height: 7
#| fig-cap: "Median alignment cross-correlation. Lighter = higher CC = better alignment."
make_heatmap(dt, "alignment_cc", "Alignment CC (median, all conditions)",
  lower_is_better = FALSE)
```

**Key observations:**

- **SRVF** achieves the lowest warp MISE across all smooth-warp DGPs (D01--D03, D05--D12).
- **Affine (S+S)** does *not* dominate on affine DGPs (D04, D13) — see @sec-affine-puzzle for discussion.
- **Landmark (auto)** is competitive on many DGPs but has highest variability.
- **FDA methods** (default and crit 1) show intermediate performance with non-trivial failure rates.
- The heatmaps pool over severity and noise; the conditional analyses below provide sharper comparisons.

# Warp Recovery {#sec-warp}

## By warp type (Q1) {#sec-q1}

*Question: Which method best recovers the true warps, conditional on warp type?*

Warp types represent fundamentally different problems — smooth elastic warps (simple/complex)
vs affine warps — so we stratify all comparisons by warp type. Within each stratum,
we show the full distribution of warp MISE across all DGPs with that warp type.

```{r}
#| label: q1-boxplot
#| fig-height: 9
#| fig-cap: "Distribution of warp MISE by method, stratified by warp type. Each boxplot summarizes all reps across all DGPs with that warp type. Columns: severity x noise conditions."
ggplot(
  dt[failure == FALSE],
  aes(x = method, y = warp_mise, fill = method)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_grid(
    warp_type ~ severity + noise_sd,
    scales = "free_y",
    labeller = labeller(
      severity = function(x) paste0("sev=", x),
      noise_sd = function(x) paste0("noise=", x)
    )
  ) +
  scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_x_discrete(labels = mlabs) +
  scale_y_log10() +
  labs(x = NULL, y = "Warp MISE (log scale)") +
  theme_benchmark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Findings:**

- On **simple** and **complex** warps, SRVF consistently achieves the lowest median warp MISE across all severity/noise conditions.
- On **affine** warps, SRVF and Landmark (auto) outperform Affine (S+S), which is paradoxical for a well-specified method (discussed in @sec-affine-puzzle).
- Higher severity and noise both increase MISE, but the *ranking* of methods is largely stable.
- FDA methods show competitive medians but heavier tails (more variable performance).

```{r}
#| label: q1-summary
q1_summary <- mc_summary(dt, warp_mise,
  by = c("method", "warp_type", "severity"))
q1_summary[, median_fmt := fmt(median)]
q1_summary[, se_fmt := fmt(mc_se)]
q1_wide <- dcast(
  q1_summary[, .(method, warp_type, severity, median_fmt, se_fmt)],
  method + warp_type ~ severity,
  value.var = c("median_fmt", "se_fmt")
)
knitr::kable(
  q1_wide[order(warp_type, method)],
  caption = "Median warp MISE (± MC SE) by method, warp type, and severity, pooling over noise and DGPs within each warp type."
)
```

## Phase-amplitude interaction (Q2) {#sec-q2}

*Question: Does amplitude variation degrade warp recovery?*

We use **amplitude ladders** — sequences of DGPs that hold template and warp type constant
while increasing amplitude complexity (none $\to$ rank-1 $\to$ rank-2 $\to$ high-rank).
An upward slope means that method's warp recovery degrades as amplitude complexity increases.

```{r}
#| label: q2-ladders
#| fig-height: 10
#| fig-cap: "Amplitude ladders: median warp MISE (IQR error bars) vs amplitude complexity. Each panel holds template and warp type fixed. Upward slopes indicate degraded warp recovery under amplitude variation."
ladders <- amplitude_ladders()
plots <- lapply(names(ladders), function(nm) {
  make_ladder_plot(dt, ladders[[nm]], nm)
})
wrap_plots(plots, ncol = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

**Findings:**

- All methods show some degradation as amplitude complexity increases, but the magnitude varies.
- **SRVF** is the most robust: its warp MISE increases modestly from "none" to "highrank" amplitude.
- **FDA methods** degrade more sharply, especially on complex warps with high-rank amplitude.
- **Affine (S+S)** is most affected: MISE increases substantially under high-rank amplitude (D13 at severity=1.0).
- **Landmark (auto)** shows moderate degradation, more stable than FDA methods.

```{r}
#| label: q2-amp-variance-ratio
#| fig-height: 8
#| fig-cap: "Amplitude variance ratio for DGPs with amplitude variation (D05--D13). Values < 1 indicate over-registration (method absorbed amplitude into phase); > 1 indicates under-registration (phase variation remains in amplitude). Median across reps per DGP/method."
dt_amp <- dt[amplitude != "none" & failure == FALSE & !is.na(amp_variance_ratio)]

if (nrow(dt_amp) > 0) {
  ggplot(dt_amp, aes(x = method, y = amp_variance_ratio, fill = method)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
    facet_wrap(~dgp, scales = "free_y") +
    scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_x_discrete(labels = mlabs) +
    scale_y_continuous(trans = "log2",
      breaks = c(0.125, 0.25, 0.5, 1, 2, 4, 8),
      labels = c("1/8", "1/4", "1/2", "1", "2", "4", "8")) +
    labs(x = NULL, y = expression("Amplitude variance ratio (" * log[2] * " scale)")) +
    theme_benchmark() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

The amplitude variance ratio reveals which methods confuse phase and amplitude.
Ratios consistently < 1 (e.g., SRVF on some high-rank DGPs) indicate over-registration
that absorbs amplitude into phase. Ratios > 1 indicate the opposite: residual phase
variation inflates the amplitude component.

## Warp complexity effect (Q3) {#sec-q3}

*Question: How much does warp complexity cost each method?*

**Warp ladders** hold template and amplitude constant while increasing warp complexity
(simple $\to$ complex $\to$ affine).

```{r}
#| label: q3-ladders
#| fig-height: 9
#| fig-cap: "Warp ladders: median warp MISE (IQR error bars) vs warp complexity. Each panel holds template and amplitude fixed."
ladders_w <- warp_ladders()
plots_w <- lapply(names(ladders_w), function(nm) {
  make_ladder_plot(dt, ladders_w[[nm]], nm)
})
wrap_plots(plots_w, ncol = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

**Findings:**

- All methods show higher MISE on complex vs simple warps, as expected.
- The transition to **affine** warps (non-domain-preserving) is revealing:
  on the "harmonic + none" ladder (D01 $\to$ D02 $\to$ D04), SRVF and landmark maintain
  low MISE on affine warps, while Affine (S+S) has higher MISE despite being well-specified.
- On the "wiggly + highrank" ladder (D11 $\to$ D12 $\to$ D13), the affine DGP (D13) shows
  particularly large MISE for Affine (S+S) at high severity — suggesting the combination
  of non-domain-preserving warps and high-rank amplitude is especially challenging for affine regression.

### The affine puzzle {#sec-affine-puzzle}

The Affine (S+S) method is **well-specified** for D04 and D13 (the true warps are
affine), yet it consistently achieves 3--4$\times$ higher warp MISE than SRVF or
Landmark (auto). Several factors likely contribute:

1. **Template estimation**: Affine registration uses the arithmetic mean as template,
   which under non-domain-preserving warps is a biased estimate of the true Fréchet
   mean in shape space. SRVF uses the Karcher mean, which may be more appropriate.
2. **Mild warps**: At severity=0.5, the affine warps are very close to the identity
   ($\text{sd}_a = 0.025$, $\text{sd}_b = 0.015$), giving the affine regression
   little signal to estimate (a, b) from.
3. **Iteration count**: The affine method uses 10 Procrustes iterations, which may
   not suffice for joint template-warp estimation in this setting.

This is a genuine finding: well-specified parametric models can underperform
flexible nonparametric methods when the signal is weak and template estimation
introduces bias.


# Registration Diagnostics {#sec-diag}

## Over/under-registration (Q4) {#sec-q4}

*Question: Do methods systematically over-register or under-register?*

The registration ratio $\|h_{\text{est}} - \text{id}\|^2 / \|h_{\text{true}} - \text{id}\|^2$
measures how much the estimated warp deviates from the identity relative to the true warp.
Values < 1 mean the method didn't warp enough; > 1 means it warped too much.

```{r}
#| label: q4-reg-ratio
#| fig-height: 8
#| fig-cap: "Registration ratio per DGP and method (median across 100 reps, severity=0.5, noise=0). Error bars: IQR. Dashed line at 1 = perfectly calibrated."
make_reg_ratio_plot(dt)
```

```{r}
#| label: q4-reg-ratio-aggregated
#| fig-height: 5
#| fig-cap: "Registration ratio aggregated by method and warp type (median and IQR across all DGPs within each warp type, severity=0.5, noise=0)."
rr_agg <- dt[
  failure == FALSE & !is.na(registration_ratio_median) &
    severity == "0.5" & noise_sd == "0",
  .(
    median_rr = median(registration_ratio_median, na.rm = TRUE),
    q25 = quantile(registration_ratio_median, 0.25, na.rm = TRUE),
    q75 = quantile(registration_ratio_median, 0.75, na.rm = TRUE)
  ),
  by = .(method, warp_type)
]

ggplot(rr_agg, aes(x = method, y = median_rr, color = method)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.2) +
  facet_wrap(~warp_type) +
  scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_x_discrete(labels = mlabs) +
  scale_y_continuous(trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("1/4", "1/2", "1", "2", "4")) +
  labs(x = NULL, y = expression("Registration ratio (" * log[2] * " scale)")) +
  theme_benchmark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Findings:**

- **SRVF** tends to slightly over-register (ratio ~1.0--1.5), especially on complex warps.
- **Landmark (auto)** is well-calibrated on simple warps but shows more variability on complex ones.
- **Affine (S+S)** shows the strongest over-registration signal, particularly on affine DGPs
  where its parametric assumptions should match. This may reflect the template estimation bias noted above.
- **FDA methods** are fairly well-calibrated (ratio near 1) on simple warps, with some under-registration on complex warps.

```{r}
#| label: q4-warp-slope
#| fig-height: 5
#| fig-cap: "Warp slope ratio (range of estimated warp derivative / range of true warp derivative) by method and warp type. Non-affine DGPs only (affine warps have constant derivative). Median and IQR across all conditions. Values > 1 = estimated warps too wiggly; < 1 = too smooth."
dt_slope <- dt[
  warp_type != "affine" & failure == FALSE & !is.na(warp_slope_ratio_median)
]
if (nrow(dt_slope) > 0) {
  slope_agg <- dt_slope[, .(
    median_sr = median(warp_slope_ratio_median, na.rm = TRUE),
    q25 = quantile(warp_slope_ratio_median, 0.25, na.rm = TRUE),
    q75 = quantile(warp_slope_ratio_median, 0.75, na.rm = TRUE)
  ), by = .(method, warp_type)]

  ggplot(slope_agg, aes(x = method, y = median_sr, color = method)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_point(size = 3) +
    geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.2) +
    facet_wrap(~warp_type) +
    scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_x_discrete(labels = mlabs) +
    scale_y_continuous(trans = "log2") +
    labs(x = NULL, y = expression("Warp slope ratio (" * log[2] * " scale)")) +
    theme_benchmark() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

The warp slope ratio reveals that **SRVF** tends to produce warps with slightly
wider derivative ranges than the truth (ratio > 1), consistent with its tendency
to over-register. Landmark and FDA methods produce smoother warps than the truth
on complex warps (ratio < 1), suggesting they under-capture the fine structure of
the true warps.

## Template estimation (Q5) {#sec-q5}

*Question: How well does each method estimate the template?*

Template MISE (L$^2$) has the caveat that it conflates shape error with phase error
in the template estimate. Two templates differing only by a warp are "the same shape"
but have nonzero L$^2$ distance.

```{r}
#| label: q5-template
#| fig-height: 7
#| fig-cap: "Distribution of template MISE by method, conditioned on severity and noise. Boxplots pool across all 13 DGPs (100 reps each)."
ggplot(
  dt[failure == FALSE],
  aes(x = method, y = template_mise, fill = method)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_grid(
    severity ~ noise_sd,
    labeller = labeller(
      severity = function(x) paste0("sev=", x),
      noise_sd = function(x) paste0("noise=", x)
    )
  ) +
  scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_x_discrete(labels = mlabs) +
  scale_y_log10() +
  labs(x = NULL, y = "Template MISE (log scale)") +
  theme_benchmark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Findings:**

- **SRVF** achieves the lowest template MISE across most conditions, consistent with its
  use of the Karcher mean.
- Noise has a moderate effect; severity has a larger effect (higher severity = more phase
  variation = harder template estimation).
- **Affine (S+S)** shows the highest template MISE, which aligns with the earlier finding
  that its template estimation (arithmetic mean) may be biased under non-trivial warping.

# Sensitivity Summary (Q6) {#sec-sensitivity}

*Question: Which methods are most sensitive to noise and severity?*

We compute **degradation ratios**: the ratio of median warp MISE under the harder condition
to the easier condition. A ratio of 2 means the metric doubled. Each dot below represents
one DGP; the crossbar shows the median ratio across DGPs.

## Noise degradation

```{r}
#| label: q6-noise
#| fig-height: 5
#| fig-cap: "Noise degradation: ratio of median warp MISE at noise=0.1 vs noise=0 (severity=0.5 fixed). Each dot = one DGP; crossbar = median across DGPs. Log2 scale: 0 = no effect, 1 = MISE doubled."
ratios <- degradation_ratios(dt)
noise_dt <- ratios$noise[severity == "0.5" & is.finite(noise_ratio)]

if (nrow(noise_dt) > 0) {
  ggplot(noise_dt, aes(x = method, y = noise_ratio, color = method)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.4,
      color = "black", linewidth = 0.4) +
    scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_x_discrete(labels = mlabs) +
    scale_y_continuous(trans = "log2",
      breaks = c(0.5, 1, 2, 4, 8),
      labels = c("0.5", "1", "2", "4", "8")) +
    labs(
      x = NULL,
      y = expression("Noise degradation ratio (" * log[2] * " scale)")
    ) +
    theme_benchmark() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

## Severity degradation

```{r}
#| label: q6-severity
#| fig-height: 5
#| fig-cap: "Severity degradation: ratio of median warp MISE at severity=1.0 vs severity=0.5 (noise=0 fixed). Each dot = one DGP; crossbar = median. Log2 scale."
sev_dt <- ratios$severity[noise_sd == "0" & is.finite(severity_ratio)]

if (nrow(sev_dt) > 0) {
  ggplot(sev_dt, aes(x = method, y = severity_ratio, color = method)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_jitter(width = 0.15, size = 2, alpha = 0.7) +
    stat_summary(fun = median, geom = "crossbar", width = 0.4,
      color = "black", linewidth = 0.4) +
    scale_color_manual(values = mcols, labels = mlabs, guide = "none") +
    scale_x_discrete(labels = mlabs) +
    scale_y_continuous(trans = "log2",
      breaks = c(1, 2, 4, 8, 16, 32),
      labels = c("1", "2", "4", "8", "16", "32")) +
    labs(
      x = NULL,
      y = expression("Severity degradation ratio (" * log[2] * " scale)")
    ) +
    theme_benchmark() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

**Findings:**

- **Noise** has the strongest effect on SRVF (median ratio ~2$\times$), which is
  expected since SRVF operates on derivatives of the curves. Affine and Landmark
  methods are less noise-sensitive.
- **Severity** increases MISE substantially for all methods (ratios 2--8$\times$),
  with the strongest effect on Affine (S+S). This is expected: larger warps are harder
  to estimate, and the affine method's template bias compounds with severity.

```{r}
#| label: q6-summary-table
noise_median <- ratios$noise[
  severity == "0.5" & is.finite(noise_ratio),
  .(noise_ratio = median(noise_ratio, na.rm = TRUE)),
  by = method
]
sev_median <- ratios$severity[
  noise_sd == "0" & is.finite(severity_ratio),
  .(severity_ratio = median(severity_ratio, na.rm = TRUE)),
  by = method
]
deg_summary <- merge(noise_median, sev_median, by = "method", all = TRUE)
deg_summary[, noise_ratio := round(noise_ratio, 2)]
deg_summary[, severity_ratio := round(severity_ratio, 2)]
knitr::kable(
  deg_summary,
  col.names = c("Method", "Noise ratio (median)", "Severity ratio (median)"),
  caption = "Median degradation ratios across DGPs. Noise ratio: MISE(noise=0.1) / MISE(noise=0) at severity=0.5. Severity ratio: MISE(sev=1) / MISE(sev=0.5) at noise=0."
)
```

# Practical (Q7) {#sec-practical}

## Runtime {#sec-runtime}

```{r}
#| label: q7-runtime
#| fig-height: 5
#| fig-cap: "Distribution of time per curve (seconds) by method across all conditions. Log10 scale. Boxplots show all non-failed runs."
ggplot(
  dt[failure == FALSE],
  aes(x = method, y = time_per_curve, fill = method)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  scale_fill_manual(values = mcols, labels = mlabs, guide = "none") +
  scale_x_discrete(labels = mlabs) +
  scale_y_log10() +
  labs(x = NULL, y = "Time per curve (seconds, log scale)") +
  theme_benchmark()
```

```{r}
#| label: q7-runtime-table
time_tbl <- dt[failure == FALSE, .(
  median = round(median(time_per_curve, na.rm = TRUE), 3),
  q25 = round(quantile(time_per_curve, 0.25, na.rm = TRUE), 3),
  q75 = round(quantile(time_per_curve, 0.75, na.rm = TRUE), 3)
), by = method][order(-median)]

knitr::kable(time_tbl,
  col.names = c("Method", "Median (s)", "Q25 (s)", "Q75 (s)"),
  caption = "Time per curve (seconds) across all conditions."
)
```

**FDA methods are by far the slowest** (order of magnitude slower than SRVF),
while Landmark and Affine are the fastest. SRVF provides a good balance of
accuracy and speed.

## Failures {#sec-failures}

```{r}
#| label: q7-failures
fail_tbl <- failure_summary(dt)
fail_tbl[, rate := round(rate, 3)]
knitr::kable(
  fail_tbl[order(method, warp_type)],
  col.names = c("Method", "Warp type", "N total", "N failed", "Rate"),
  caption = "Failure rates by method and warp type."
)
```

```{r}
#| label: q7-error-analysis
#| results: asis
err <- dt[failure == TRUE & nchar(error_msg) > 0]
err[, error_class := fcase(
  grepl("duplicated values", error_msg), "Duplicated arg values in warped curves",
  grepl("computationally singular", error_msg), "Singular linear system",
  grepl("missing value where TRUE/FALSE", error_msg), "NA in logical comparison",
  default = "Other"
)]
err_summary <- err[, .N, by = .(method, error_class)][order(method, -N)]

cat("**Error analysis:**\n\n")
cat(sprintf("Total failures: %d out of %d runs (%.1f%%).\n\n",
  nrow(err), nrow(dt), 100 * nrow(err) / nrow(dt)))

cat("Failures are **concentrated in the FDA methods** (fda_default: ",
  sum(err$method == "fda_default"), "; fda_crit1: ",
  sum(err$method == "fda_crit1"), "). ",
  "The dominant failure modes are:\n\n", sep = "")

cat("1. **Duplicated arg values** (", sum(err$error_class ==
    "Duplicated arg values in warped curves"),
  " cases): The FDA registration algorithm produces warps that map ",
  "multiple input positions to the same output position, creating ",
  "duplicated argument values when constructing the warped curves. ",
  "This typically occurs on DGPs with complex warps or high-rank amplitude ",
  "where the iterative alignment overshoots.\n\n", sep = "")

cat("2. **Singular linear systems** (", sum(err$error_class ==
    "Singular linear system"),
  " cases): The B-spline basis matrix becomes numerically singular, ",
  "usually when the warping concentrates too many evaluation points ",
  "into a narrow interval.\n\n", sep = "")

cat("3. **NA in logical comparison** (", sum(err$error_class ==
    "NA in logical comparison"),
  " cases): Internal convergence check encounters NA values.\n\n", sep = "")

cat("SRVF, Affine, and Landmark methods have near-zero failure rates. ",
  "The 3 Landmark failures are edge cases on specific DGPs where ",
  "landmark detection finds insufficient features.\n\n")
```

# Data Completeness {#sec-completeness}

```{r}
#| label: completeness
comp <- dt[, .N, by = .(dgp, method)]
comp_wide <- dcast(comp, dgp ~ method, value.var = "N")
knitr::kable(comp_wide,
  caption = "Number of runs per DGP x method (target: 400 = 2 severity x 2 noise x 100 reps). Missing cells indicate batches that did not complete on the cluster."
)
```

```{r}
#| label: missing-batches
#| results: asis
expected <- CJ(
  dgp = levels(dt$dgp),
  method = levels(dt$method)
)
actual <- unique(dt[, .(dgp, method)])
missing <- expected[!actual, on = .(dgp, method)]
if (nrow(missing) > 0) {
  cat("**Missing DGP x method combinations** (", nrow(missing),
    " batches still running on LRZ):\n\n", sep = "")
  for (i in seq_len(nrow(missing))) {
    cat(sprintf("- %s / %s\n", missing$dgp[i], missing$method[i]))
  }
} else {
  cat("All DGP x method combinations present.\n")
}
```

# Summary {#sec-summary}

```{r}
#| label: summary-ranking
#| results: asis
cat("## Key findings\n\n")

cat("1. **SRVF is the best general-purpose method** for warp recovery on smooth ",
  "(domain-preserving) warps. It achieves the lowest median warp MISE across ",
  "all simple and complex warp DGPs and is competitive even on affine DGPs.\n\n")

cat("2. **Amplitude variation degrades warp recovery for all methods**, but the ",
  "degradation is modest for SRVF and more severe for FDA and Affine methods. ",
  "High-rank (3-FPC) amplitude is harder than low-rank, confirming that ",
  "phase-amplitude separation is a genuine challenge.\n\n")

cat("3. **The affine method underperforms on affine DGPs** despite being well-specified. ",
  "This appears to stem from biased template estimation (arithmetic mean) and ",
  "weak signal at low severity. See @sec-affine-puzzle.\n\n")

cat("4. **FDA methods have the highest failure rates** (~2-5% depending on DGP), ",
  "driven by numerical issues in the iterative warping (duplicated arg values, ",
  "singular systems). They are also the slowest by an order of magnitude.\n\n")

cat("5. **Noise sensitivity varies by method**: SRVF is most noise-sensitive ",
  "(operates on derivatives), while Landmark and Affine are more robust. ",
  "All methods degrade substantially with increased severity (larger warps).\n\n")

cat("6. **Landmark (auto) is surprisingly competitive** on many DGPs, matching ",
  "or exceeding FDA methods. It has near-zero failure rate and fast runtime. ",
  "Its weakness is variability when landmark detection is unreliable.\n\n")
```

```{r}
#| label: summary-table
rr_med <- dt[
  failure == FALSE & !is.na(registration_ratio_median),
  .(reg_ratio = round(median(registration_ratio_median, na.rm = TRUE), 2)),
  by = method
]
runtime <- dt[failure == FALSE,
  .(time_s = round(median(time_per_curve, na.rm = TRUE), 3)),
  by = method]
fail_rate <- dt[, .(fail_pct = round(100 * mean(failure), 1)), by = method]

summary_dt <- Reduce(function(a, b) merge(a, b, by = "method"), list(
  rr_med, runtime, fail_rate
))

knitr::kable(
  summary_dt[order(method)],
  col.names = c("Method", "Reg. ratio", "Time/curve (s)", "Failure %"),
  caption = "Summary per method. Reg. ratio: median registration ratio (1 = perfectly calibrated). Time/curve: median seconds. Failure %: overall failure rate."
)
```

# Session Info {.unnumbered}

```{r}
#| label: session-info
#| cache: false
sessionInfo()
```
