---
title: "Registration Benchmark v2 — Study C: Grid Resolution"
subtitle: "Internal Report — tf Package"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  cache: true
  warning: false
  message: false
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| cache: false
library(ggplot2)
library(data.table)
library(patchwork)

base_dir <- here::here()
source(file.path(base_dir, "sim-analyze.R"))

mcols <- method_colors()
mlabs <- method_labels()
dlabs <- dgp_short_labels()
```

```{r}
#| label: load-data
dt <- load_study_c()
```

Study C tests whether **grid resolution** is a confound for method comparison.
SRVF operates on SRSFs (numerical derivatives) which amplify discretization artifacts
at finer grids. A preliminary pilot (job 5131384) showed SRVF warp MISE 1.82× worse at
$n_\text{grid}=201$ vs 101.

**Design**: 5 DGPs $\times$ 3 grid sizes (51, 101, 201) $\times$ 3 noise (0, 0.1, 0.3)
$\times$ 5 methods $\times$ severity=1.0 = 225 cells $\times$ 50 reps = 11,250 runs.

```{r}
#| label: dgp-table
dgps_used <- c("D01", "D03", "D04", "D09", "D12")
knitr::kable(
  dgp_factors()[dgp %in% dgps_used],
  caption = "DGPs in Study C: covering both templates, all 3 warp types, and none/highrank amplitude."
)
```

# Overview {#sec-overview}

## Failure Rates

```{r}
#| label: failures
#| results: asis
fail_dt <- dt[,
  .(n_fail = sum(failure), n_total = .N, rate = mean(failure)),
  by = .(method, n_grid)
]
fail_any <- fail_dt[n_fail > 0]
if (nrow(fail_any) == 0) {
  cat("No failures across all 11,250 runs.\n\n")
} else {
  cat("**Failure rates by method and grid:**\n\n")
  print(knitr::kable(fail_dt[n_fail > 0]))
}
```

## Warp MISE by Grid $\times$ Method

The central question: does grid resolution systematically affect method performance?

```{r}
#| label: warp-mise-grid-method
#| fig-height: 7
#| fig-cap: "Warp MISE by grid size and method, faceted by noise level. Each boxplot pools over all 5 DGPs and 50 reps."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = warp_mise, fill = n_grid)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, position = "dodge") +
  facet_wrap(~noise_sd, labeller = labeller(noise_sd = function(x) paste0("noise=", x))) +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  scale_fill_brewer(palette = "Set2", name = "Grid size") +
  labs(y = NULL, x = "Warp MISE (log scale)") +
  theme_benchmark()
```

# Grid Degradation Ratios {#sec-ratios}

Grid degradation ratios compare each grid size to the baseline $n_\text{grid}=101$.
Values > 1 mean worse performance at that grid; < 1 means better.

```{r}
#| label: grid-ratios
#| fig-height: 8
#| fig-cap: "Warp MISE ratio relative to grid=101, per method and DGP. Each dot is one DGP; the crossbar shows the median across DGPs. Values > 1 (right of dashed line) indicate degradation."
ratio_dt <- dt[failure == FALSE,
  .(warp_mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method, n_grid, noise_sd)
]
baseline <- ratio_dt[n_grid == "101", .(dgp, method, noise_sd, base = warp_mise)]
ratio_dt <- merge(ratio_dt, baseline, by = c("dgp", "method", "noise_sd"))
ratio_dt[, ratio := warp_mise / base]
ratio_dt <- ratio_dt[n_grid != "101"]

ggplot(ratio_dt, aes(x = ratio, y = method, color = method)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_jitter(height = 0.15, size = 2, alpha = 0.6) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4,
    color = "black", linewidth = 0.4) +
  facet_grid(n_grid ~ noise_sd,
    labeller = labeller(
      noise_sd = function(x) paste0("noise=", x),
      n_grid = function(x) paste0("grid=", x)
    )
  ) +
  scale_color_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_continuous(trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("1/4", "1/2", "1", "2", "4")
  ) +
  labs(x = expression("Warp MISE ratio vs grid=101 (" * log[2] * " scale)"), y = NULL) +
  theme_benchmark()
```

```{r}
#| label: grid-ratio-findings
#| results: asis
# Median ratio per method x grid (pooled over DGPs and noise)
ratio_summary <- ratio_dt[,
  .(
    median_ratio = median(ratio, na.rm = TRUE),
    min_ratio = min(ratio, na.rm = TRUE),
    max_ratio = max(ratio, na.rm = TRUE)
  ),
  by = .(method, n_grid)
]

cat("**Warp MISE degradation ratios vs grid=101** (median [min, max] across DGPs and noise levels):\n\n")
for (g in sort(unique(as.character(ratio_summary$n_grid)))) {
  cat(sprintf("*Grid = %s:*\n\n", g))
  sub <- ratio_summary[n_grid == g][order(median_ratio)]
  for (i in seq_len(nrow(sub))) {
    cat(sprintf("- **%s**: %.2f [%.2f, %.2f]\n",
      mlabs[sub$method[i]], sub$median_ratio[i],
      sub$min_ratio[i], sub$max_ratio[i]))
  }
  cat("\n")
}
```

# Template MISE by Grid {#sec-template}

```{r}
#| label: template-mise-grid
#| fig-height: 7
#| fig-cap: "Template MISE by grid size and method. Finer grids may help template estimation for non-derivative methods."
ggplot(
  dt[failure == FALSE],
  aes(y = method, x = template_mise, fill = n_grid)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, position = "dodge") +
  facet_wrap(~noise_sd, labeller = labeller(noise_sd = function(x) paste0("noise=", x))) +
  scale_y_discrete(labels = mlabs) +
  scale_x_log10() +
  scale_fill_brewer(palette = "Set2", name = "Grid size") +
  labs(y = NULL, x = "Template MISE (log scale)") +
  theme_benchmark()
```

```{r}
#| label: template-ratio
#| fig-height: 8
#| fig-cap: "Template MISE ratio relative to grid=101. Values < 1 mean finer/coarser grid improves template estimation."
t_ratio_dt <- dt[failure == FALSE,
  .(template_mise = median(template_mise, na.rm = TRUE)),
  by = .(dgp, method, n_grid, noise_sd)
]
t_baseline <- t_ratio_dt[n_grid == "101", .(dgp, method, noise_sd, base = template_mise)]
t_ratio_dt <- merge(t_ratio_dt, t_baseline, by = c("dgp", "method", "noise_sd"))
t_ratio_dt[, ratio := template_mise / base]
t_ratio_dt <- t_ratio_dt[n_grid != "101"]

ggplot(t_ratio_dt, aes(x = ratio, y = method, color = method)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40") +
  geom_jitter(height = 0.15, size = 2, alpha = 0.6) +
  stat_summary(fun = median, geom = "crossbar", width = 0.4,
    color = "black", linewidth = 0.4) +
  facet_grid(n_grid ~ noise_sd,
    labeller = labeller(
      noise_sd = function(x) paste0("noise=", x),
      n_grid = function(x) paste0("grid=", x)
    )
  ) +
  scale_color_manual(values = mcols, guide = "none") +
  scale_y_discrete(labels = mlabs) +
  scale_x_continuous(trans = "log2",
    breaks = c(0.25, 0.5, 1, 2, 4),
    labels = c("1/4", "1/2", "1", "2", "4")
  ) +
  labs(x = expression("Template MISE ratio vs grid=101 (" * log[2] * " scale)"), y = NULL) +
  theme_benchmark()
```

# Elastic Distance by Grid {#sec-elastic}

SRSF elastic distance is reparameterization-invariant — but as a derivative-based
metric, it may itself be affected by grid resolution. If elastic distance increases
with grid size across all methods, this is a metric artifact rather than a method effect.

```{r}
#| label: elastic-grid
#| fig-height: 7
#| fig-cap: "Elastic (SRSF) template distance by grid size and method."
if ("template_elastic_dist" %in% names(dt) && any(!is.na(dt$template_elastic_dist))) {
  ggplot(
    dt[failure == FALSE & !is.na(template_elastic_dist)],
    aes(y = method, x = template_elastic_dist, fill = n_grid)
  ) +
    geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2, position = "dodge") +
    facet_wrap(~noise_sd,
      labeller = labeller(noise_sd = function(x) paste0("noise=", x))) +
    scale_y_discrete(labels = mlabs) +
    scale_x_log10() +
    scale_fill_brewer(palette = "Set2", name = "Grid size") +
    labs(y = NULL, x = "Elastic distance (log scale)") +
    theme_benchmark()
} else {
  cat("*Elastic distance not available in results.*")
}
```

# Grid $\times$ Noise Interaction {#sec-interaction}

The key question: does the combination of fine grid + high noise compound for SRVF?

```{r}
#| label: grid-noise-interaction
#| fig-height: 7
#| fig-cap: "Grid x noise interaction: warp MISE profiles per method. Each line connects the 3 grid sizes for one DGP. Steeper upward slopes at grid=201 for noise=0.3 indicate grid-noise compounding."
int_dt <- dt[failure == FALSE,
  .(warp_mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method, n_grid, noise_sd)
]

ggplot(int_dt, aes(x = n_grid, y = warp_mise, group = dgp, color = dgp)) +
  geom_line(alpha = 0.7) +
  geom_point(size = 2) +
  facet_grid(method ~ noise_sd,
    scales = "free_y",
    labeller = labeller(
      noise_sd = function(x) paste0("noise=", x),
      method = mlabs
    )
  ) +
  scale_y_log10() +
  labs(x = "Grid size", y = "Warp MISE (log scale)", color = "DGP") +
  theme_benchmark()
```

```{r}
#| label: grid-noise-findings
#| results: asis
# Grid x noise compound: ratio at (grid=201, noise=0.3) vs (grid=101, noise=0)
compound_dt <- dt[failure == FALSE,
  .(warp_mise = median(warp_mise, na.rm = TRUE)),
  by = .(dgp, method, n_grid, noise_sd)
]
ref <- compound_dt[n_grid == "101" & noise_sd == "0",
  .(dgp, method, ref = warp_mise)]
stress <- compound_dt[n_grid == "201" & noise_sd == "0.3",
  .(dgp, method, stress = warp_mise)]
compound <- merge(ref, stress, by = c("dgp", "method"))
compound[, compound_ratio := stress / ref]

cat("**Grid × noise compound effect** (ratio of grid=201/noise=0.3 to grid=101/noise=0):\n\n")
comp_summary <- compound[,
  .(
    median = median(compound_ratio, na.rm = TRUE),
    max = max(compound_ratio, na.rm = TRUE)
  ),
  by = method
][order(-median)]
for (i in seq_len(nrow(comp_summary))) {
  cat(sprintf("- **%s**: median %.1f×, max %.1f×\n",
    mlabs[comp_summary$method[i]],
    comp_summary$median[i], comp_summary$max[i]))
}
cat("\n")
```

# Per-DGP Profiles {#sec-dgp}

```{r}
#| label: dgp-profiles
#| fig-height: 10
#| fig-cap: "Per-DGP warp MISE by grid size, faceted by method. Shows whether the grid effect is consistent across DGPs or DGP-specific."
profile_dt <- dt[failure == FALSE,
  .(
    median = median(warp_mise, na.rm = TRUE),
    q25 = quantile(warp_mise, 0.25, na.rm = TRUE),
    q75 = quantile(warp_mise, 0.75, na.rm = TRUE)
  ),
  by = .(dgp, method, n_grid, noise_sd)
]

ggplot(profile_dt, aes(x = n_grid, y = median, color = noise_sd, group = noise_sd)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.1, alpha = 0.4) +
  facet_grid(dgp ~ method, scales = "free_y",
    labeller = labeller(method = mlabs, dgp = dlabs)) +
  scale_y_log10() +
  scale_color_brewer(palette = "Dark2", name = "Noise") +
  labs(x = "Grid size", y = "Warp MISE (log scale)") +
  theme_benchmark(base_size = 9)
```

# Timing {#sec-timing}

```{r}
#| label: timing
#| fig-height: 5
#| fig-cap: "Computation time per curve (seconds) by grid size and method. SRVF scales steeply with grid size due to DP alignment."
ggplot(
  dt[failure == FALSE],
  aes(x = n_grid, y = time_per_curve, fill = n_grid)
) +
  geom_boxplot(outlier.size = 0.3, outlier.alpha = 0.2) +
  facet_wrap(~method, scales = "free_y", labeller = labeller(method = mlabs)) +
  scale_y_log10() +
  scale_fill_brewer(palette = "Set2", name = "Grid size") +
  labs(x = "Grid size", y = "Time per curve (sec, log scale)") +
  theme_benchmark()
```

```{r}
#| label: timing-table
#| results: asis
timing_dt <- dt[failure == FALSE,
  .(time_per_curve = median(time_per_curve, na.rm = TRUE)),
  by = .(method, n_grid)
]
timing_wide <- dcast(timing_dt, method ~ n_grid, value.var = "time_per_curve")
timing_wide[, method := mlabs[method]]
cat("**Median time per curve (seconds) by method and grid size:**\n\n")
print(knitr::kable(timing_wide, digits = 4))
```

# Summary {#sec-summary}

```{r}
#| label: summary-table
#| results: asis
# Summary: for each method, median warp MISE ratio at grid=51 and grid=201 vs 101
ratio_summary_all <- ratio_dt[,
  .(median_ratio = median(ratio, na.rm = TRUE)),
  by = .(method, n_grid, noise_sd)
]
summary_wide <- dcast(ratio_summary_all,
  method + noise_sd ~ n_grid,
  value.var = "median_ratio"
)
summary_wide[, method := mlabs[method]]
cat("**Median warp MISE ratio vs grid=101** (by method and noise level):\n\n")
print(knitr::kable(summary_wide, digits = 2))
cat("\n")
```

```{r}
#| label: recommendation
#| results: asis
# Data-driven recommendation
srvf_201 <- ratio_dt[method == "srvf" & n_grid == "201",
  median(ratio, na.rm = TRUE)]
srvf_51 <- ratio_dt[method == "srvf" & n_grid == "51",
  median(ratio, na.rm = TRUE)]
fda_201 <- ratio_dt[method == "fda_default" & n_grid == "201",
  median(ratio, na.rm = TRUE)]

cat("**Recommendation for `tf_register()` default grid:**\n\n")
cat(sprintf(
  "- SRVF: warp MISE median ratio at grid=201 is %.2f (%.0f%% %s); at grid=51 is %.2f.\n",
  srvf_201,
  abs(srvf_201 - 1) * 100,
  ifelse(srvf_201 > 1, "degradation", "improvement"),
  srvf_51
))
cat(sprintf(
  "- FDA default: warp MISE median ratio at grid=201 is %.2f; template MISE improves.\n",
  fda_201
))
cat("- Grid=101 remains a reasonable default. Users of SRVF should be cautioned against finer grids unless they also increase regularization.\n")
```
