---
title: "Registration Benchmark v2 — Study B: Penalization"
subtitle: "Internal Report — tf Package"
format:
  html:
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true
    code-fold: true
    theme: flatly
    fig-width: 10
    fig-height: 6
    self-contained: true
execute:
  cache: true
  warning: false
  message: false
---

# Setup {.unnumbered}

```{r}
#| label: setup
#| cache: false
library(ggplot2)
library(data.table)
library(patchwork)

base_dir <- here::here()
source(file.path(base_dir, "sim-analyze.R"))

mcols <- method_colors()
mlabs <- method_labels()
dlabs <- dgp_short_labels()
```

```{r}
#| label: load-data
dt <- load_study_b()
```

Study B tests whether regularization (lambda > 0) can rescue registration methods
in regimes where they struggle. Design: 4 DGPs $\times$ 8 lambdas $\times$ 3 methods
(SRVF, FDA default, FDA crit1) $\times$ 3 noise $\times$ 2 severity $\times$ 33 reps
= 19,008 runs.

Lambda grid calibrated by pilot (job 5131464):
`c(0, 1e-4, 1e-3, 0.01, 0.05, 0.1, 1, 10)`.

```{r}
#| label: overview
#| results: asis
cat(sprintf("- Rows: %d | Failures: %d (%.1f%%)\n",
  nrow(dt), sum(dt$failure), 100 * mean(dt$failure)))
cat(sprintf("- DGPs: %s\n", paste(levels(dt$dgp), collapse = ", ")))
cat(sprintf("- Methods: %s\n", paste(levels(dt$method), collapse = ", ")))
cat(sprintf("- Lambdas: %s\n", paste(sort(unique(dt$lambda)), collapse = ", ")))
cat(sprintf("- Noise: %s\n", paste(levels(dt$noise_sd), collapse = ", ")))
```

# MISE vs Lambda Curves {#sec-mise-lambda}

The core question: does the MISE-vs-lambda curve show a U-shape (optimal lambda > 0)?

```{r}
#| label: mise-lambda-curves
#| fig-height: 12
#| fig-cap: "Median warp MISE vs lambda for each method, faceted by DGP (rows) and noise level (columns). Dashed horizontal line = lambda=0 baseline. Points show median across reps; error bars show IQR."
agg <- dt[failure == FALSE,
  .(mise = median(warp_mise, na.rm = TRUE),
    q25 = quantile(warp_mise, 0.25, na.rm = TRUE),
    q75 = quantile(warp_mise, 0.75, na.rm = TRUE)),
  by = .(method, dgp, noise_sd, severity, lambda)
]

# Baseline (lambda = 0) for reference lines
baseline <- agg[lambda == 0, .(method, dgp, noise_sd, severity, base = mise)]

ggplot(agg[severity == "1"],
  aes(x = lambda + 1e-5, y = mise, color = method)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.15, alpha = 0.4) +
  geom_hline(data = baseline[severity == "1"],
    aes(yintercept = base, color = method),
    linetype = "dashed", alpha = 0.5) +
  facet_grid(dgp ~ noise_sd,
    scales = "free_y",
    labeller = labeller(
      noise_sd = function(x) paste0("noise = ", x),
      dgp = dlabs)) +
  scale_x_log10(
    breaks = c(1e-5, 1e-4, 1e-3, 0.01, 0.05, 0.1, 1, 10),
    labels = c("0", "1e-4", "1e-3", "0.01", "0.05", "0.1", "1", "10")) +
  scale_color_manual(values = mcols, labels = mlabs) +
  scale_y_log10() +
  labs(x = expression(lambda ~ "(log scale)"),
    y = "Warp MISE (median, log scale)",
    color = "Method") +
  theme_benchmark() +
  theme(legend.position = "bottom")
```

```{r}
#| label: mise-lambda-sev05
#| fig-height: 12
#| fig-cap: "Same as above but for severity = 0.5."
ggplot(agg[severity == "0.5"],
  aes(x = lambda + 1e-5, y = mise, color = method)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  geom_errorbar(aes(ymin = q25, ymax = q75), width = 0.15, alpha = 0.4) +
  geom_hline(data = baseline[severity == "0.5"],
    aes(yintercept = base, color = method),
    linetype = "dashed", alpha = 0.5) +
  facet_grid(dgp ~ noise_sd,
    scales = "free_y",
    labeller = labeller(
      noise_sd = function(x) paste0("noise = ", x),
      dgp = dlabs)) +
  scale_x_log10(
    breaks = c(1e-5, 1e-4, 1e-3, 0.01, 0.05, 0.1, 1, 10),
    labels = c("0", "1e-4", "1e-3", "0.01", "0.05", "0.1", "1", "10")) +
  scale_color_manual(values = mcols, labels = mlabs) +
  scale_y_log10() +
  labs(x = expression(lambda ~ "(log scale)"),
    y = "Warp MISE (median, log scale)",
    color = "Method") +
  theme_benchmark() +
  theme(legend.position = "bottom")
```

# Optimal Lambda {#sec-optimal}

```{r}
#| label: optimal-lambda
#| results: asis
best <- dt[failure == FALSE,
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(method, dgp, noise_sd, severity, lambda)
]
optimal <- best[, .SD[which.min(mise)], by = .(method, dgp, noise_sd, severity)]

cat("## Optimal lambda per method x DGP x condition\n\n")
opt_wide <- dcast(
  optimal[, .(method, dgp, severity, noise_sd, lambda)],
  method + dgp + severity ~ noise_sd,
  value.var = "lambda"
)
knitr::kable(opt_wide, caption = "Optimal lambda by method, DGP, and severity. Columns show noise levels.")
```

```{r}
#| label: optimal-lambda-findings
#| results: asis
# Check if optimal lambda shifts with noise
cat("\n## Does optimal lambda shift with noise?\n\n")
for (m in levels(dt$method)) {
  sub <- optimal[method == m]
  # Show median optimal lambda per noise level (pooling DGPs and severity)
  by_noise <- sub[, .(
    med_lambda = median(lambda),
    n_zero = sum(lambda == 0),
    n_total = .N
  ), by = noise_sd]
  cat(sprintf("- **%s**: median optimal λ shifts across noise: %s. ",
    mlabs[m],
    paste(sprintf("noise=%s → %g (%d/%d at λ=0)",
      by_noise$noise_sd, by_noise$med_lambda,
      by_noise$n_zero, by_noise$n_total), collapse = "; ")))
  # Flag boundary hits
  n_boundary <- sum(sub$lambda == max(sub$lambda))
  if (n_boundary > 0) {
    cat(sprintf("**%d/%d cells hit boundary (λ=%g).**",
      n_boundary, nrow(sub), max(sub$lambda)))
  }
  cat("\n")
}
```

# Improvement from Penalization {#sec-improvement}

How much does the best lambda improve over lambda=0?

**Note:** This is an *oracle* analysis — the optimal lambda is selected ex post from
the same data used to evaluate it. Ratios $\leq 1$ by construction (the baseline
lambda=0 is always in the candidate set). The practical benefit of penalization
depends on whether a good lambda can be selected a priori or via cross-validation.

```{r}
#| label: improvement-heatmap
#| fig-height: 8
#| fig-cap: "Improvement ratio: MISE(optimal lambda) / MISE(lambda=0). Values < 1 mean penalization helps. Color scale: green = large improvement, white = no change, red = lambda hurts."
base <- best[lambda == 0, .(method, dgp, noise_sd, severity, base_mise = mise)]
merged <- merge(optimal, base, by = c("method", "dgp", "noise_sd", "severity"))
merged[, ratio := mise / base_mise]

ggplot(merged,
  aes(x = interaction(noise_sd, severity, sep = " / sev="),
    y = interaction(dgp, method, sep = " : "),
    fill = ratio)) +
  geom_tile(color = "white") +
  geom_text(aes(label = sprintf("%.2f\n(λ=%g)", ratio, lambda)),
    size = 2.5) +
  scale_fill_gradient2(low = "#2ca02c", mid = "white", high = "#d62728",
    midpoint = 1, limits = c(0, 1.5), oob = scales::squish) +
  labs(x = "noise / severity",
    y = "DGP : Method",
    fill = "MISE ratio\n(vs λ=0)") +
  theme_benchmark() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: improvement-findings
#| results: asis
cat("**Findings:**\n\n")

# Biggest winners
big_win <- merged[ratio < 0.8][order(ratio)]
if (nrow(big_win) > 0) {
  cat("Cases where penalization improves MISE by >20%:\n\n")
  for (i in seq_len(min(nrow(big_win), 10))) {
    r <- big_win[i]
    cat(sprintf("- %s on %s (noise=%s, sev=%s): %.0f%% improvement at λ=%g\n",
      mlabs[r$method], r$dgp, r$noise_sd, r$severity,
      100 * (1 - r$ratio), r$lambda))
  }
}

# Cases where lambda hurts
hurt <- merged[ratio > 1.05]
if (nrow(hurt) > 0) {
  cat(sprintf("\nCases where best lambda *hurts* (ratio > 1.05): %d/%d\n",
    nrow(hurt), nrow(merged)))
}
cat("\n")
```

# FDA Criterion Comparison {#sec-criteria}

Does fda_crit1 respond differently to lambda than fda_default?

```{r}
#| label: criteria-comparison
#| fig-height: 8
#| fig-cap: "Direct comparison of fda_default vs fda_crit1 MISE across lambda values. Each panel is a DGP × noise combination at severity=1.0."
fda_dt <- dt[failure == FALSE & method %in% c("fda_default", "fda_crit1")]
fda_agg <- fda_dt[severity == "1",
  .(mise = median(warp_mise, na.rm = TRUE)),
  by = .(method, dgp, noise_sd, lambda)
]

ggplot(fda_agg,
  aes(x = lambda + 1e-5, y = mise, color = method)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  facet_grid(dgp ~ noise_sd,
    scales = "free_y",
    labeller = labeller(
      noise_sd = function(x) paste0("noise = ", x),
      dgp = dlabs)) +
  scale_x_log10(
    breaks = c(1e-5, 1e-4, 1e-3, 0.01, 0.05, 0.1, 1, 10),
    labels = c("0", "1e-4", "1e-3", "0.01", "0.05", "0.1", "1", "10")) +
  scale_color_manual(values = mcols, labels = mlabs) +
  scale_y_log10() +
  labs(x = expression(lambda ~ "(log scale)"),
    y = "Warp MISE (median, log scale)",
    color = "Method") +
  theme_benchmark() +
  theme(legend.position = "bottom")
```

```{r}
#| label: criteria-findings
#| results: asis
# Compare optimal lambdas for the two FDA methods
fda_opt <- optimal[method %in% c("fda_default", "fda_crit1")]
fda_comp <- dcast(
  fda_opt[, .(method, dgp, noise_sd, severity, lambda, mise)],
  dgp + noise_sd + severity ~ method,
  value.var = list("lambda", "mise")
)

cat("**FDA criterion comparison:**\n\n")
cat("- Optimal lambda tends to be ",
  if (median(fda_comp$lambda_fda_crit1, na.rm = TRUE) >
      median(fda_comp$lambda_fda_default, na.rm = TRUE))
    "**higher**" else "**lower**",
  " for fda_crit1 than fda_default across conditions.\n", sep = "")

# Which criterion wins at its own optimal lambda?
fda_comp[, crit1_wins := mise_fda_crit1 < mise_fda_default]
cat(sprintf("- At each method's optimal lambda, fda_crit1 wins in %d/%d conditions.\n",
  sum(fda_comp$crit1_wins, na.rm = TRUE), nrow(fda_comp)))
cat("\n")
```

# SRVF Lambda Behavior {#sec-srvf}

SRVF's lambda response is expected to differ qualitatively from FDA methods.

```{r}
#| label: srvf-detail
#| fig-height: 6
#| fig-cap: "SRVF warp MISE vs lambda across all DGPs and noise levels (severity=1.0). The lambda response ranges from 'U-shape' (lambda helps) to 'monotone increasing' (lambda only hurts)."
srvf_agg <- dt[failure == FALSE & method == "srvf" & severity == "1",
  .(mise = median(warp_mise, na.rm = TRUE),
    q25 = quantile(warp_mise, 0.25, na.rm = TRUE),
    q75 = quantile(warp_mise, 0.75, na.rm = TRUE)),
  by = .(dgp, noise_sd, lambda)
]

ggplot(srvf_agg,
  aes(x = lambda + 1e-5, y = mise, color = noise_sd)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_ribbon(aes(ymin = q25, ymax = q75, fill = noise_sd),
    alpha = 0.15, color = NA) +
  facet_wrap(~dgp, scales = "free_y", labeller = as_labeller(dlabs)) +
  scale_x_log10(
    breaks = c(1e-5, 1e-4, 1e-3, 0.01, 0.05, 0.1, 1, 10),
    labels = c("0", "1e-4", "1e-3", "0.01", "0.05", "0.1", "1", "10")) +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_log10() +
  labs(x = expression(lambda ~ "(log scale)"),
    y = "Warp MISE (median ± IQR, log scale)",
    color = "Noise", fill = "Noise") +
  theme_benchmark() +
  theme(legend.position = "bottom")
```

```{r}
#| label: srvf-findings
#| results: asis
srvf_opt <- optimal[method == "srvf"]
cat("**SRVF lambda findings:**\n\n")

for (d in levels(dt$dgp)) {
  sub <- srvf_opt[dgp == d]
  lambdas <- sub[, .(noise_sd, lambda, mise)]
  cat(sprintf("- **%s** (%s): optimal λ = %s\n",
    d, dlabs[d],
    paste(sprintf("%g (noise=%s)", lambdas$lambda, lambdas$noise_sd),
      collapse = ", ")))
}
cat("\n")

# Does lambda ever help SRVF substantially?
srvf_merged <- merge(srvf_opt, base[method == "srvf"],
  by = c("method", "dgp", "noise_sd", "severity"))
srvf_merged[, ratio := mise / base_mise]
helps <- srvf_merged[ratio < 0.9]
if (nrow(helps) > 0) {
  cat(sprintf("Lambda helps SRVF (>10%% improvement) in %d/%d conditions:\n",
    nrow(helps), nrow(srvf_merged)))
  for (i in seq_len(nrow(helps))) {
    r <- helps[i]
    cat(sprintf("  - %s noise=%s sev=%s: %.0f%% at λ=%g\n",
      r$dgp, r$noise_sd, r$severity, 100*(1-r$ratio), r$lambda))
  }
} else {
  cat("Lambda does not substantially help SRVF (>10%) in any condition.\n")
}
cat("\n")
```

# Template Quality under Penalization {#sec-template}

Does lambda affect template estimation quality?

```{r}
#| label: template-lambda
#| fig-height: 8
#| fig-cap: "Template MISE vs lambda. Penalization can improve template estimation by smoothing the iterative estimation."
tmpl_agg <- dt[failure == FALSE & severity == "1",
  .(tmise = median(template_mise, na.rm = TRUE)),
  by = .(method, dgp, noise_sd, lambda)
]

ggplot(tmpl_agg,
  aes(x = lambda + 1e-5, y = tmise, color = method)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.5) +
  facet_grid(dgp ~ noise_sd,
    scales = "free_y",
    labeller = labeller(
      noise_sd = function(x) paste0("noise = ", x),
      dgp = dlabs)) +
  scale_x_log10(
    breaks = c(1e-5, 1e-4, 1e-3, 0.01, 0.05, 0.1, 1, 10),
    labels = c("0", "1e-4", "1e-3", "0.01", "0.05", "0.1", "1", "10")) +
  scale_color_manual(values = mcols, labels = mlabs) +
  scale_y_log10() +
  labs(x = expression(lambda ~ "(log scale)"),
    y = "Template MISE (median, log scale)",
    color = "Method") +
  theme_benchmark() +
  theme(legend.position = "bottom")
```

# Summary {#sec-summary}

```{r}
#| label: summary-table
#| results: asis
# Summary table: best lambda and improvement per method x DGP
summary_dt <- merge(optimal, base, by = c("method", "dgp", "noise_sd", "severity"))
summary_dt[, improvement := sprintf("%.0f%%", 100 * (1 - mise / base_mise))]
summary_dt[, cell := sprintf("λ=%g (%s)", lambda, improvement)]

summary_wide <- dcast(
  summary_dt[severity == "1", .(method, dgp, noise_sd, cell)],
  method + dgp ~ noise_sd,
  value.var = "cell"
)

cat("## Optimal lambda and improvement at severity=1.0\n\n")
knitr::kable(
  summary_wide[order(method, dgp)],
  caption = "Optimal lambda and MISE improvement vs lambda=0, by method × DGP × noise (severity=1.0)."
)
```

```{r}
#| label: summary-findings
#| results: asis
cat("## Key takeaways\n\n")

# Count how often lambda > 0 is optimal
n_helps <- nrow(optimal[lambda > 0])
n_total <- nrow(optimal)
cat(sprintf("- Lambda > 0 is optimal in %d/%d (%.0f%%) of method × DGP × condition cells.\n",
  n_helps, n_total, 100 * n_helps / n_total))

# Biggest beneficiary
biggest <- merged[which.min(ratio)]
cat(sprintf("- Largest improvement: %s on %s (noise=%s, sev=%s) — %.0f%% at λ=%g.\n",
  mlabs[biggest$method], biggest$dgp, biggest$noise_sd, biggest$severity,
  100 * (1 - biggest$ratio), biggest$lambda))

# Method that benefits most on average
avg_by_method <- merged[, .(mean_ratio = mean(ratio, na.rm = TRUE)), by = method]
best_method <- avg_by_method[which.min(mean_ratio)]
cat(sprintf("- Method benefiting most from penalization on average: %s (mean ratio %.2f).\n",
  mlabs[best_method$method], best_method$mean_ratio))
cat("\n")
```
